---
title: "Contributing to hvtiPlotR"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{Contributing to hvtiPlotR}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE,
                      warning = FALSE)
```

This guide is for two audiences:

- **Biostatisticians and analysts** who want to port a SAS template to R
  and add it to this package. Start at [Track A](#track-a).
- **R package developers** who want to work on testing, CI, or package
  infrastructure. Start at [Track B](#track-b).

---

# Development environment setup {#setup}

## Prerequisites

- R ≥ 4.1
- RStudio (recommended) or any R-capable editor
- Git

## Installing development dependencies

Clone the repository and install all dependencies declared in `DESCRIPTION`:

```r
# Clone (first time only)
# git clone https://github.com/ehrlinger/hvtiPlotR.git

# Install devtools if needed
install.packages("devtools")

# Install all Imports + Suggests from DESCRIPTION
devtools::install_deps(dependencies = TRUE)
```

## The development workflow loop

```r
# Load the package into the current session without installing
devtools::load_all()

# Regenerate NAMESPACE and .Rd files from roxygen comments
devtools::document()

# Run all tests
devtools::test()

# Full R CMD CHECK (must pass before merging)
devtools::check()

# Build vignettes only
devtools::build_vignettes()
```

`devtools::load_all()` re-sources all `R/*.R` files and makes every exported
function available immediately — much faster than `install.packages()` during
development.

---

# Track A — Porting a SAS template {#track-a}

This track walks through adding a brand-new plot function by porting an
existing SAS template. We use a fictional template
`tp.np.bmi.avrg_curv.binary.sas` as the running example.

## Step 1: Understand the SAS template output

Before writing any R code, identify:

1. **What datasets does the template produce?**
   Most `tp.np.*` templates export a `predict` dataset (the fitted curve)
   and a `means` dataset (binned patient-level summaries). Export these to
   CSV from SAS to use as your development data.

2. **What are the SAS column names?**
   Document the mapping in the roxygen `@description` block. For example:

   | SAS column | R column | Meaning |
   |---|---|---|
   | `iv_bmi` | `time` | x-axis variable (BMI) |
   | `mean_curv` | `estimate` | predicted probability |
   | `cll_p68` | `lower` | 68 % CI lower bound |
   | `clu_p68` | `upper` | 68 % CI upper bound |

3. **Which existing R function is closest?**
   Check `_pkgdown.yml` and the plot-functions vignette. You may only need
   to extend an existing function rather than create a new one.

## Step 2: Create the R source file

Create `R/my-curve-plot.R` (use kebab-case for the filename). Every plot
family lives in its own file — plot functions and their sample-data
companions together.

```r
# File: R/bmi-curve-plot.R

#' Average BMI Curve Plot
#'
#' Plots a nonparametric average curve of a binary outcome against BMI
#' (or any continuous covariate), with an optional confidence ribbon and
#' binned data summary points. Ports
#' \code{tp.np.bmi.avrg_curv.binary.sas}.
#'
#' **SAS column mapping:**
#' - `time`     ← `iv_bmi` (BMI on the x-axis)
#' - `estimate` ← `mean_curv` (predicted probability)
#' - `lower`    ← `cll_p68`  (68 % CI lower)
#' - `upper`    ← `clu_p68`  (68 % CI upper)
#'
#' @param curve_data  Data frame of fitted curve output (one row per x value).
#' @param x_col       Name of the x-axis column. Default `"time"`.
#' @param estimate_col Name of the predicted value column. Default `"estimate"`.
#' @param lower_col   Name of the lower CI column, or `NULL` for no ribbon.
#'   Default `NULL`.
#' @param upper_col   Name of the upper CI column, or `NULL`. Default `NULL`.
#' @param data_points Optional data frame of binned summary points.
#'   Must have columns `x_col` and `"value"`. Default `NULL`.
#' @param line_width  Width of the curve line. Default `1.0`.
#' @param point_size  Size of data summary points. Default `2.5`.
#'
#' @return A bare [ggplot2::ggplot()] object.
#'
#' @seealso [hv_nonparametric()], [sample_bmi_curve_data()]
#'
#' @references SAS template: \code{tp.np.bmi.avrg_curv.binary.sas}.
#'
#' @examples
#' library(ggplot2)
#' dat <- sample_bmi_curve_data(n = 500)
#' bmi_curve_plot(dat, lower_col = "lower", upper_col = "upper") +
#'   scale_colour_manual(values = c("steelblue"), guide = "none") +
#'   scale_fill_manual(values   = c("steelblue"), guide = "none") +
#'   scale_x_continuous(limits = c(18, 45), breaks = seq(20, 45, 5)) +
#'   scale_y_continuous(limits = c(0, 0.5),
#'                      labels = scales::percent) +
#'   labs(x = "BMI (kg/m²)", y = "Prevalence of AF") +
#'   theme_hv_manuscript()
#'
#' @importFrom ggplot2 ggplot aes geom_line geom_ribbon geom_point
#' @importFrom rlang .data
#' @export
bmi_curve_plot <- function(curve_data,
                           x_col        = "time",
                           estimate_col = "estimate",
                           lower_col    = NULL,
                           upper_col    = NULL,
                           data_points  = NULL,
                           line_width   = 1.0,
                           point_size   = 2.5) {

  # ----- Input checks -------------------------------------------------------
  if (!is.data.frame(curve_data))
    stop("`curve_data` must be a data frame.")
  for (col in c(x_col, estimate_col)) {
    if (!(col %in% names(curve_data)))
      stop(sprintf("Column '%s' not found in `curve_data`.", col))
  }

  # ----- Build base plot ----------------------------------------------------
  p <- ggplot2::ggplot(
    curve_data,
    ggplot2::aes(x = .data[[x_col]], y = .data[[estimate_col]])
  )

  # Optional CI ribbon
  if (!is.null(lower_col) && !is.null(upper_col)) {
    p <- p + ggplot2::geom_ribbon(
      ggplot2::aes(ymin = .data[[lower_col]], ymax = .data[[upper_col]],
                   fill = "1"),
      alpha = 0.2
    )
  }

  # Main curve
  p <- p + ggplot2::geom_line(
    ggplot2::aes(colour = "1"),
    linewidth = line_width
  )

  # Optional data summary points
  if (!is.null(data_points)) {
    p <- p + ggplot2::geom_point(
      data = data_points,
      ggplot2::aes(x = .data[[x_col]], y = .data[["value"]],
                   colour = "1", shape = "1"),
      size = point_size
    )
  }

  p
}
```

Key rules enforced here:

- **Column names are strings**, passed via `x_col =`, `estimate_col =`, etc.
  Never use `enquo()` or `{{ }}` — it makes column names opaque to the
  caller.
- **No colours or themes** are applied inside the function. The caller adds
  `scale_colour_*()` and `theme_hv_manuscript()` outside.
- **`.data[[col]]`** is used for tidy evaluation instead of bare variable
  names. This requires `@importFrom rlang .data`.
- The function returns **`p`**, not `print(p)` or `p + theme(...)`.

## Step 3: Add a sample-data generator

Add `sample_bmi_curve_data()` to the same file. The generator should:

- Accept `n` (patient count), `time_max` / range, and `seed`.
- Return a data frame whose column names **match the R defaults** (`time`,
  `estimate`, `lower`, `upper`) — not the SAS column names.
- Produce realistic-looking data at a plausible scale.

```r
#' Sample BMI Curve Data
#'
#' Simulates the fitted curve output from a nonparametric BMI analysis,
#' matching the column layout expected by [bmi_curve_plot()].
#'
#' @param n       Number of simulated patients (controls CI width).
#'   Default `500`.
#' @param bmi_min Lower end of the BMI range. Default `18`.
#' @param bmi_max Upper end of the BMI range. Default `50`.
#' @param n_points Number of points on the prediction grid. Default `200`.
#' @param seed    Random seed. Default `42`.
#'
#' @return A data frame with columns `time` (BMI grid), `estimate`,
#'   `lower`, `upper`.
#'
#' @seealso [bmi_curve_plot()]
#'
#' @examples
#' dat <- sample_bmi_curve_data(n = 300)
#' head(dat)
#' range(dat$time)     # BMI range
#' range(dat$estimate) # probability range, 0-1
#'
#' @importFrom stats plogis runif qnorm
#' @export
sample_bmi_curve_data <- function(n       = 500,
                                  bmi_min = 18,
                                  bmi_max = 50,
                                  n_points = 200,
                                  seed    = 42L) {
  set.seed(seed)
  z    <- stats::qnorm(0.84)   # 68 % CI ≈ 1 SD
  bmi  <- seq(bmi_min, bmi_max, length.out = n_points)
  eta  <- -2 + 0.05 * (bmi - 30)   # logistic link
  est  <- stats::plogis(eta)
  se   <- sqrt(est * (1 - est) / (n * 0.02))

  data.frame(
    time     = bmi,
    estimate = est,
    lower    = pmax(est - z * se, 0),
    upper    = pmin(est + z * se, 1)
  )
}
```

## Step 4: Write roxygen documentation

The documentation block must include all of the following:

| Tag | Required | Notes |
|---|---|---|
| `@description` | Yes | One paragraph; mention the SAS template name |
| `@param` | Yes | One per argument; include default in description |
| `@return` | Yes | Describe the ggplot or data frame structure |
| `@seealso` | Yes | Link to the companion `sample_*()` and related functions |
| `@references` | Yes | The exact SAS template filename(s) |
| `@examples` | Yes | Must be runnable (use `\dontrun{}` only for file I/O) |
| `@importFrom` | Yes | Declare every function used from other packages |
| `@export` | Yes | Must be present for the function to be accessible |

Run `devtools::document()` after editing to check for parse errors.

## Step 5: Register in `_pkgdown.yml`

Add the new functions to the appropriate section in `_pkgdown.yml`. If no
existing section fits, add a new one:

```yaml
- title: "Nonparametric Covariate Curves"
  desc: >
    Average curves plotted against a continuous covariate (BMI, age, etc.)
    rather than against time.
  contents:
  - bmi_curve_plot
  - sample_bmi_curve_data
```

## Step 6: Add a worked example to the plot-functions vignette

Open `vignettes/plot-functions.qmd` and add a new top-level section before
the "Draft Footnotes" section:

````markdown
# BMI Curve Plot

`bmi_curve_plot()` plots a fitted nonparametric average curve of a binary
outcome against BMI ...

```{r bmi_curve_basic}
library(ggplot2)
dat <- sample_bmi_curve_data(n = 500)
bmi_curve_plot(dat, lower_col = "lower", upper_col = "upper") + ...
```
````

## Step 7: Add a row to the SAS migration guide

In `vignettes/sas-migration-guide.qmd`, add a row to the lookup table at
the top:

```markdown
| `tp.np.bmi.avrg_curv.binary.sas` | np | `bmi_curve_plot()` | [BMI curve](#np-bmi) |
```

Then add the corresponding section with a runnable example further down in
the `# Nonparametric temporal trends` family.

## Step 8: Write tests

Create `tests/testthat/test_bmi_curve_plot.R`:

```r
library(testthat)
library(ggplot2)

test_that("sample_bmi_curve_data returns a data frame with required columns", {
  dat <- sample_bmi_curve_data(n = 100, seed = 1)
  expect_true(is.data.frame(dat))
  expect_true(all(c("time", "estimate", "lower", "upper") %in% names(dat)))
})

test_that("sample_bmi_curve_data respects n_points", {
  dat <- sample_bmi_curve_data(n_points = 50, seed = 1)
  expect_equal(nrow(dat), 50)
})

test_that("bmi_curve_plot returns a ggplot", {
  dat <- sample_bmi_curve_data(n = 100, seed = 1)
  p   <- bmi_curve_plot(dat)
  expect_s3_class(p, "ggplot")
})

test_that("bmi_curve_plot adds ribbon when CI columns are supplied", {
  dat    <- sample_bmi_curve_data(n = 100, seed = 1)
  p_ci   <- bmi_curve_plot(dat, lower_col = "lower", upper_col = "upper")
  layers <- sapply(p_ci$layers, function(l) class(l$geom)[1])
  expect_true("GeomRibbon" %in% layers)
})

test_that("bmi_curve_plot errors on missing column", {
  dat <- sample_bmi_curve_data(n = 100)
  expect_error(bmi_curve_plot(dat, x_col = "no_such_col"))
})
```

Run with `devtools::test(filter = "bmi")`.

## Step 9: Update NEWS.md

Add a bullet to the current dev version at the top of `NEWS.md`:

```markdown
* Added `bmi_curve_plot()` and `sample_bmi_curve_data()` — nonparametric
  average curve of a binary outcome against a continuous covariate (BMI).
  Ports `tp.np.bmi.avrg_curv.binary.sas`.
```

## Step 10: Final checklist before opening a PR

```r
devtools::document()   # regenerate NAMESPACE + .Rd — must complete without errors
devtools::test()       # all tests pass
devtools::check()      # 0 errors, 0 warnings, 0 notes (ideally)
```

Then open a pull request against `main` on GitHub.

---

# Track B — Package infrastructure {#track-b}

## Package structure overview

```
hvtiPlotR/
├── R/                    # Source: one file per plot family
├── man/                  # Auto-generated .Rd files — do not edit manually
├── tests/
│   └── testthat/         # One test_*.R per source file
├── vignettes/            # Quarto (.qmd) vignettes
├── inst/                 # Bundled files (ClevelandClinic.pptx, extdata/)
├── DESCRIPTION           # Package metadata and dependency declarations
├── NAMESPACE             # Auto-generated — do not edit manually
├── _pkgdown.yml          # Website reference and articles layout
└── NEWS.md               # User-visible changelog
```

`NAMESPACE` and `man/*.Rd` are both auto-generated by `devtools::document()`
from the roxygen comments in `R/*.R`. **Never edit these files by hand.**

## Adding a dependency

- **Runtime dependency** (used inside a plot function): add to `Imports` in
  `DESCRIPTION` and declare with `@importFrom pkg fn` in the roxygen block.
- **Vignette/test only**: add to `Suggests`. Wrap any code that uses it in
  `if (requireNamespace("pkg", quietly = TRUE)) { ... }` or use
  `#| eval: false` in the vignette chunk.

```r
# Check before adding — is it already available in base R or an existing Import?
# Keep Imports lean; every new dependency adds installation weight.
```

## Code style and conventions {#conventions}

### File and function naming

| Item | Convention | Example |
|---|---|---|
| Source files | `kebab-case.R` | `bmi-curve-plot.R` |
| Plot functions | `snake_case()` | `bmi_curve_plot()` |
| Sample generators | `sample_` prefix | `sample_bmi_curve_data()` |
| Internal helpers | `.` prefix | `.bmi_compute_ci()` |

Internal helpers (prefixed with `.`) should have `@keywords internal` and
**no** `@export` — they will not appear in `NAMESPACE` or generate `.Rd`
files.

### The bare-ggplot pattern

Every plot function must:

1. Accept model-output data frames (not raw patient data).
2. Map all column references through string arguments (`x_col =`, etc.).
3. Use `.data[[col]]` for tidy evaluation inside `aes()`.
4. Return an unstyled ggplot — no `scale_*()`, no `labs()`, no theme.
5. Not call `print()` or `invisible()`.

This pattern lets callers layer on their own choices and enables the
`+` composition grammar documented in `vignettes/plot-decorators.qmd`.

### Tidy evaluation

```r
# Correct — column name is a string; .data masks the data frame
ggplot2::aes(x = .data[[x_col]], y = .data[[estimate_col]])

# Wrong — bare symbol; fails when x_col is a variable
ggplot2::aes(x = x_col, y = estimate_col)

# Wrong — enquo; column name is opaque to the caller
ggplot2::aes(x = !!rlang::enquo(x_col))
```

Always declare `.data` in the roxygen block:
```r
#' @importFrom rlang .data
```

## Testing {#testing}

### Test file layout

Each source file `R/my-plot.R` should have a corresponding
`tests/testthat/test_my_plot.R`. The minimum set of tests for a new plot
function:

| Test | What to check |
|---|---|
| Sample data shape | Correct class, column names, and number of rows |
| Sample data types | Numeric / logical / factor columns are the right type |
| Plot return class | `expect_s3_class(p, "ggplot")` |
| Layer presence | CI ribbon added when `lower_col` is provided |
| Error on bad input | `expect_error(fn(dat, x_col = "no_such_col"))` |

### Snapshot tests

The `tests/testthat/_snaps/` directory stores snapshot outputs for
`expect_snapshot()` tests. To update snapshots after an intentional change:

```r
testthat::snapshot_review()   # review diffs interactively
testthat::snapshot_accept()   # accept all pending diffs
```

### Running checks

```r
devtools::test()               # run all tests
devtools::test(filter = "bmi") # run tests matching "bmi"
devtools::check()              # full R CMD CHECK
rcmdcheck::rcmdcheck(          # more detailed output
  args = "--no-manual",
  error_on = "warning"
)
```

### What R CMD CHECK validates

- All examples in `@examples` blocks run without error.
- All exported functions have documentation.
- `NAMESPACE` matches the actual exports.
- No undefined global variables (lintr / `R CMD CHECK` NOTE).
- Vignette chunks marked `eval: true` run successfully.

## Vignette conventions {#vignette-conventions}

Vignettes live in `vignettes/` as `.qmd` files and are built with Quarto
(`VignetteBuilder: quarto`). Rules:

- All code chunks that read files, write files, or access a network must
  have `#| eval: false`.
- Use `here::here()` for file paths in eval-false chunks (not hard-coded
  absolute paths).
- Use `system.file("ClevelandClinic.pptx", package = "hvtiPlotR")` for the
  bundled PPT template — never a hard-coded path.
- After adding a new vignette, register it in `_pkgdown.yml` under
  `articles`.

## Releasing a new version {#releasing}

1. Update the version in `DESCRIPTION` (follow
   [semantic versioning](https://semver.org/): `MAJOR.MINOR.PATCH`).
2. Move the dev entries in `NEWS.md` to a new `# hvtiPlotR X.Y.Z` heading.
3. Run `devtools::check()` — zero errors and zero warnings required.
4. Tag the release commit: `git tag -a vX.Y.Z -m "Release X.Y.Z"`.
5. Push the tag: `git push origin vX.Y.Z`.
6. The pkgdown GitHub Action rebuilds the documentation site automatically.

---

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
