---
title: "Decorating and Saving hvtiPlotR Plots"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{Decorating and Saving hvtiPlotR Plots}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Installed vignettes embed every figure, so 1x resolution halves the tarball
# (17 MB to 9 MB). The pkgdown site links its figures and keeps retina 2.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.retina = if (identical(Sys.getenv("IN_PKGDOWN"), "true")) 2 else 1)
```

```{r libraries}
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(ggplot2)
library(hvtiPlotR)
```

# The Composition Pattern

Every hvtiPlotR plot is built in two steps: a constructor (`hv_*()`) that
shapes the data, followed by `plot()` that renders a bare `ggplot` object. No
color scales, axis labels, or theme are applied by either step. You add those
by chaining layers with `+`:

```
plot(hv_*(...)) +
  scale_color_*() +   # data colors
  scale_fill_*()   +   # fill colors
  labs()           +   # axis labels, title, caption
  annotate()       +   # text/arrows placed on the panel
  coord_cartesian() +  # viewport cropping
  theme_hv_manuscript()         # non-data formatting
```

This vignette walks through each decorator in turn, using `hv_trends()` and
`hv_survival()` as representative base plots.

```{r base_data}
# Trends data — multi-group continuous outcome over time
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

# KM data — survival curve
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)
```

# Themes

The **hvtiPlotR** package provides four themes, one per output target.

| Theme | Target |
|-------|--------|
| `theme_hv_manuscript()` | Journal PDF, black-on-white |
| `theme_hv_poster()` | Conference poster, slightly larger text |
| `theme_hv_ppt_light()` | PowerPoint on white/light background |
| `theme_hv_ppt_dark()` | PowerPoint on dark/blue background |

## Manuscript

`theme_hv_manuscript()` is the default for journal submissions: white
background, black text, minimal chrome. Font sizes are tuned for 8.5 x 11 inch
letter-size PDFs where the figure may be reduced to a column width. Use this
whenever the figure ends up in a Word or PDF document sent to a journal.

```{r theme_hv_manuscript}
p_ms <- p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript()

p_ms
```

## Poster

`theme_hv_poster()` bumps the axis text and tick weights up from the manuscript
baseline and removes the panel grid, so the figure reads clearly at arm's
length on a 36 x 48 inch foam board. It is also the theme for in-room
slide presentations when you are not using the PowerPoint themes; the
larger tick labels and the boxed panel hold up on a projector better than the
manuscript variant. Pass `base_family = "sans"` (or another family) to switch
the font face.

```{r theme_hv_poster}
p_poster <- p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()

p_poster
```

## Light PowerPoint

`theme_hv_ppt_light()` matches slides with a white or light-gray background:
the Cleveland Clinic standard template, most default Office themes, and any
deck where the content area is light. Text and lines are dark, so the figure
reads without modification when placed on the light slide. Pair with
`save_ppt()` to insert it as editable DrawingML.

```{r theme_hv_ppt_light}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_ppt_light()
```

## Dark PowerPoint

`theme_hv_ppt_dark()` flips the palette for dark-background slides, the navy or
dark-blue gradient decks where white-on-dark text is the convention. The theme
sets a transparent plot background, so the slide's background shows through
behind the panel. The `plot.background` override in the chunk below simulates
that navy background in the rendered vignette; you would omit it when saving
to an actual `.pptx` file.

```{r theme_hv_ppt_dark}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_ppt_dark() +
  theme(plot.background = element_rect(fill = "navy", color = "navy"))
```

## One decorator for a whole deck

The two dark-slide chunks above show the shape of the problem. Every plot in a
deck wants the same theme, the same colors and the same shapes, and copying
those lines figure by figure is how one slide quietly drifts out of step with
the rest. A theme cannot carry any of it. `theme()` governs the
non-data ink, the text, panel, grid and ticks, and nothing that draws from the
data, so no theme element sets a series color. Passing the layer to the theme
instead does not rescue it: `theme_hv_ppt_dark(geom_point(color = "yellow"))`
forwards its `...` to `theme()`, which errors on anything that is not a theme
element. Colors and shapes belong to the scales, added with `+`.

`hv_ppt_series()` returns the theme and both scales as a single object.
ggplot2's `+` unrolls a plain list, so you define the look once and add it
wherever you need it:

```{r hv_ppt_series}
ppt <- hv_ppt_series(mode = "dark")

p_base +
  labs(x = "Surgery Year", y = "Outcome") +
  ppt +
  theme(plot.background = element_rect(fill = "navy", color = "navy"))
```

The colors are the Okabe-Ito colorblind-safe palette, reordered so the
highest-luminance hues come first on a dark slide and the darker ones first on
a light slide. Color and shape both map the grouping column, which is worth
keeping even though it looks redundant on your monitor: a projector in a dark
room flattens color differences that read cleanly on screen, and the shapes
survive where the hues do not. Pass `colors` or `shapes` when a deck calls for
a specific set.

## Naming the series without a legend

House style carries no legend. Every `theme_hv_*()` sets
`legend.position = "none"`, and `hv_ppt_series()` leaves that alone, because a
CORR figure names the series where the series sits rather than sending the
reader out to a key and back.

The label takes the same ink as the axis, white on a dark slide and black on a
light slide or in a manuscript, not the color of the curve it names:

```{r hv_ppt_series_annotate}
p_base +
  labs(x = "Surgery Year", y = "Outcome") +
  ppt +
  annotate("text", x = 2012, y = 56, label = "Group I",
           color = "white", size = 5, hjust = 0) +
  theme(plot.background = element_rect(fill = "navy", color = "navy"))
```

`hv_ppt_palette()` hands back the colors themselves when a figure needs them
outside the decorator, so a script reaches for the one definition instead of
pasted hex codes:

```{r hv_ppt_palette}
hv_ppt_palette("dark")
hv_ppt_palette("light", n = 4)
```

If a working draft does want a key, pass `legend.position` straight through:
`hv_ppt_series(legend.position = "top", name = "Group")`. Color and shape
share `name`, so ggplot2 merges them into one legend rather than stacking two.

## Designing the slide around the figure

`hv_ppt_series()` settles how the plot looks. What the slide around it says is
still your call, and a few habits decide whether the audience reads the figure
or squints at it.

- **Orient the room first.** Before any result, say which color is which
  group: "Group I is blue, Group II is vermilion." The audience cannot read
  the finding until they know the key.
- **Make the headline a sentence.** Put a one-sentence assertion over the
  figure, the finding the plot is evidence for, in place of a topic label.
  "Survival after repair is higher in Group I" tells the room what to look
  for; "Survival by group" leaves them to work it out. This is the
  assertion-evidence structure of Garner and Alley (2013).
- **Label the series on the plot.** A name placed beside its curve keeps the
  word and the line together, Mayer's spatial contiguity principle, and it is
  why the house themes draw no legend (see above).
- **Set labels inside the figure at 32 pt.** The group's house rule is a
  32 pt floor for slide text, and it covers the words on the plot too: series
  names, region labels such as RCA and LCA, annotations. The PowerPoint themes
  default to `base_size = 32` for axis text; text you add with `annotate()` or
  `geom_text()` takes its size in millimeters, so pass `size = 32 / .pt`.
  This is a house rule, not a tested threshold.
- **Highlight the series that carries the message.** When one group is the
  point of the slide, give it the strong color and let the others recede to
  gray. That is Mayer's signaling principle, and his coherence principle
  argues for dropping anything on the slide that does not serve the
  assertion.
- **Keep the axes and the panel fixed across a build-up.** When a slide
  sequence adds one series at a time, set the limits once with
  `coord_cartesian()` and reuse them. `save_ppt()` already anchors the panel
  at the same slide coordinates on every slide through its `panel_box`
  argument; keep that one box for the whole sequence. The frame stays still
  and only the data changes.
- **If a pie chart is unavoidable, order the slices as you will discuss
  them**, not from largest to smallest. A sorted bar or dot chart is still the
  better choice.
- **Hold the contrast floor.** Every color in `hv_ppt_palette()` clears 3:1
  against its background, black for the dark ordering and a white slide for
  the light one, which is the contrast WCAG 2.2 success criterion 1.4.11 asks
  of a chart's graphical objects. `theme_hv_ppt_light()` leaves its panel
  transparent so the slide shows through, so the light guarantee assumes a
  white template, as the house light template is. The colors are Okabe-Ito's, except that on a light slide
  a darkened yellow replaces their orange, which is too faint on white. If
  you pass your own `colors`, check them against the same floor.

The [slide design chapter](https://ehrlinger.github.io/hvtiGraphics/slide_design.html)
of the hvtiGraphics recipes book carries the rest of the house guidance,
worked through on full decks.

# Color Scales

`scale_color_*` controls line and point colors; `scale_fill_*` controls filled areas (ribbons, bars). Both take the same `name` (legend title) and `guide` (legend display) arguments; set them once and both scales update.

## Manual colors

Use `scale_color_manual()` when assigning specific brand or convention colors
to known levels.

```{r scale_color_manual}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()
```

## ColorBrewer palettes

`scale_color_brewer()` applies a ColorBrewer palette. These are perceptually uniform
and print-safe, so they survive a black-and-white PDF. Use
`"RdBu"` or `"PuOr"` for ordered or diverging data, `"Blues"` for sequential.
For categorical groups use `scale_color_manual(values = hv_ppt_palette("light"))`
instead, as below: six colorblind-safe colors. Avoid `"Set1"` and `"RdYlGn"`,
whose red against green fails for readers with deuteranopia.

```{r scale_color_brewer}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()
```

## Suppressing legends

Pass `guide = "none"` to any scale and that aesthetic drops out of the legend.
You'd do this when the axis labels or an annotation already name the group;
a redundant legend only takes up panel space.

```{r scale_no_legend}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), guide = "none") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    guide = "none"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()
```

# Labels and Annotations

## labs()

`labs()` sets axis labels, the plot title, legend title, subtitle, and caption text. We set
these outside the constructor so you can override them per project without
touching the function itself.

```{r labs_example}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, \u03b1 = 0.6827 (1 SD)"
  ) +
  theme_hv_poster()
```

## annotate()

`annotate()` places text, segments, rectangles, or arrows at fixed data coordinates,
such as a sample-size callout in the corner or a label pointing to an
event of interest.

```{r annotate_example}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), color = "gray40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, color = "gray40") +
  theme_hv_poster()
```

## coord_cartesian()

`coord_cartesian()` crops the viewport without dropping data, preserving LOESS
fits computed on the full range.

```{r coord_cartesian}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  theme_hv_poster()
```

# Saving Figures

## Manuscript PDF

Use `ggsave()` with `width = 11, height = 8.5` (US Letter landscape) for
manuscript figures. Assign the fully composed plot to a variable first so the
same object is both displayed in the session and written to disk.

```{r save_manuscript}
#| eval: false
ggsave(
  filename = "../graphs/trends_manuscript.pdf",
  plot     = p_ms,
  width    = 11,
  height   = 8.5
)
```

## Poster PDF

Poster figures are typically larger and use `theme_hv_poster()`. Adjust
dimensions to match the poster panel size.

```{r save_poster}
#| eval: false
ggsave(
  filename = "../graphs/trends_poster.pdf",
  plot     = p_poster,
  width    = 14,
  height   = 10
)
```

## PowerPoint slides

`save_ppt()` inserts ggplot objects into a PowerPoint file as **editable
DrawingML vector graphics** via the `officer` and `rvg` packages. Shapes,
lines, and text remain selectable in PowerPoint after export.

Key arguments:

| Argument | Default | Notes |
|---|---|---|
| `object` | (required) | A single ggplot **or** a named/unnamed list of ggplots |
| `template` | shipped `hv_ppt_template.pptx` | Existing `.pptx` used as the slide template |
| `powerpoint` | (required) | Output file path |
| `slide_titles` | `"Plot"` | Character vector recycled to the number of plots |
| `layout` | `"Title and Content"` | Slide layout from the template |
| `width` / `height` | `10.1` / `5.8` | Plot area in inches |
| `left` / `top` | `0.0` / `1.2` | Position from slide edges, in inches |

Apply `theme_hv_ppt_dark()` or `theme_hv_ppt_light()` before saving to
match the slide background.

### Single slide

Build a fully themed plot with `theme_hv_ppt_dark()` or `theme_hv_ppt_light()`,
then pass it as `object`. The `template` argument points to an existing
`.pptx` file whose slide master and layouts carry the brand fonts and
background; `save_ppt()` appends a new slide rather than overwriting the
template.

```{r save_ppt_single}
#| eval: false
template <- system.file("extdata", "hv_ppt_template.pptx", package = "hvtiPlotR")

p_ppt <- p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  theme_hv_ppt_dark()

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = here::here("graphs", "trends_slides.pptx"),
  slide_titles = "Temporal Trends by Group"
)
```

### Multiple slides from a list

Pass a named list of plots and a matching vector of titles to produce one slide
per plot in a single call. This is the pattern we use for batch-report decks
where each outcome gets its own slide. The list keeps plots in order and the
names serve as a paper trail. Every plot in the list should carry the same
theme so the deck looks consistent.

```{r save_ppt_list}
#| eval: false
dta_km2 <- sample_survival_data(n = 400, seed = 99)
km2     <- hv_survival(dta_km2)

p_km_ppt <- plot(km2) +
  scale_color_manual(values = c(All = "white"), guide = "none") +
  scale_fill_manual(values  = c(All = "white"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_ppt_dark()

save_ppt(
  object       = list(trends = p_ppt, survival = p_km_ppt),
  template     = template,
  powerpoint   = here::here("graphs", "multi_slide_deck.pptx"),
  slide_titles = c("Temporal Trends by Group", "Overall Survival")
)
```

### Fixed panel placement across slides

When plots have different y-axis ranges ("0-1" vs "0-10000"), the axis-label
widths differ, which shifts the plot panel inside a fixed `ph_location()`. On
dark PPT themes with a visible panel background (black panel on a blue-gradient
slide), the box appears to jump between slides.

Pass `panel_box = list(width, height, left, top)` to anchor the **panel content
area** at the same slide coordinates on every slide. `save_ppt()` calls
`hv_ph_location()` for each plot, measures the axis/legend/title chrome
around the panel, and adjusts per-slide placement so the panel lands at the
target rectangle regardless of label width. Axis labels then extend outside the
panel box as needed; device dimensions vary per plot.

```{r save_ppt_panel_box}
#| eval: false
save_ppt(
  object       = list(trends = p_ppt, survival = p_km_ppt),
  template     = template,
  powerpoint   = here::here("graphs", "multi_slide_deck_aligned.pptx"),
  slide_titles = c("Temporal Trends by Group", "Overall Survival"),
  panel_box    = list(width = 10, height = 5, left = 1.5, top = 1.5)
)
```

Make the `left` and `top` entries of `panel_box` large enough to leave room for the widest
axis labels in the deck, or `hv_ph_location()` warns that the plot
chrome spills off the slide edge.

## Multi-panel PDF (EDA batch output)

When generating multiple plots in a loop, `patchwork::wrap_plots()` arranges
them into a grid and `ggsave()` writes each page. This is typical for EDA
batches where you have a dozen or more outcomes to inspect. You build the
list with `lapply()`, chunk it into pages of 9, and let the loop handle
pagination. The `per_page` constant is easy to adjust if you want a 2x2 or
4x4 grid instead.

```{r save_multipanel}
#| eval: false
# Build a list of plots (e.g. from an hv_eda() lapply loop)
plot_list <- lapply(
  c("ef", "lv_mass", "peak_grad"),
  function(yv) {
    dta_eda <- sample_eda_data()
    plot(hv_eda(dta_eda, x_col = "op_years", y_col = yv,
                 y_label = yv)) +
      scale_color_manual(values = c("steelblue"), guide = "none") +
      labs(x = "Years") +
      theme_hv_poster()
  }
)

per_page <- 9L
for (pg in seq(1, length(plot_list), by = per_page)) {
  idx     <- seq(pg, min(pg + per_page - 1L, length(plot_list)))
  pg_plot <- patchwork::wrap_plots(plot_list[idx], nrow = 3, ncol = 3)
  ggsave(
    filename = sprintf(here::here("graphs", "eda_page%02d.pdf"),
                       ceiling(pg / per_page)),
    plot     = pg_plot,
    width    = 14,
    height   = 14
  )
}
```

# Legend Positioning

ggplot2 places the legend outside the panel by default. For publication figures
we usually move it inside or drop it, because the axis labels often do the
identification work already.

## Inside the panel

Pass fractional coordinates `c(x, y)` to `legend.position` inside `theme()`.
`c(0, 0)` is the bottom-left corner; `c(1, 1)` is the top-right.

```{r legend_inside}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = NULL) +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = NULL
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.position  = c(0.15, 0.2),        # bottom-left of panel
    legend.background = element_rect(fill = "white", color = "gray80",
                                     linewidth = 0.3)
  )
```

## Outside the panel (explicit sides)

Set `legend.position` to `"right"`, `"left"`, `"top"`, or `"bottom"` to place
the legend outside the panel area. `"bottom"` with
`legend.direction = "horizontal"` is the most common choice for multi-group
figures where a long label would crowd a corner inside the panel.

```{r legend_outside}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.position = "bottom",             # "right" | "left" | "top" | "bottom"
    legend.direction = "horizontal"
  )
```

## Suppress all legends

Two ways to suppress legends, with slightly different scope.
`theme(legend.position = "none")` hides every legend for the plot in one
shot. `guide = "none"` on a `scale_*()` call drops only that aesthetic's
legend, for when you want a color legend but no shape legend (or vice
versa). Both reclaim the panel width.

```{r legend_none}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()
```

`guide = "none"` on every `scale_*()` call removes all legend entries. This is
preferred over `theme(legend.position = "none")` when only some aesthetics
have legends and others do not.

## Legend text and key size

The legend inherits font size from the active theme, which is often slightly
larger than needed when the legend sits inside a crowded panel. `legend.text`
controls the label font; `legend.key.size` shrinks the color/shape swatch.
Both accept any `element_text()` or `unit()` value.

```{r legend_text}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.text  = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  )
```

# Theme Overrides

`theme_hv_manuscript()` sets a complete non-data formatting baseline. Layer additional
`theme()` calls after it to adjust individual elements without touching the
rest.

## Axis text size

Override `axis.text` to resize tick labels and `axis.title` for the axis
title, independent of the rest of the theme. This is useful when a figure is
being resized for a different output target, for example scaling down for a
two-column journal layout where the default poster-sized text would be too
large.

```{r theme_axis_text}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    axis.text  = element_text(size = 10),   # tick labels
    axis.title = element_text(size = 12)    # axis titles
  )
```

## Removing minor grid lines

Minor grid lines (`panel.grid.minor`) add density without adding precision;
they can make a busy figure look cluttered, especially when there are many data
series. Setting `element_blank()` removes them while keeping the major grid
intact.

```{r theme_grid}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    panel.grid.minor = element_blank()
  )
```

## Rotating x-axis labels

Useful for time-point labels or long category names. The combination of
`angle = 45`, `hjust = 1`, and `vjust = 1` keeps the rotated label
right-aligned to its tick mark; without the `hjust`/`vjust` corrections the
labels will float off-center.

```{r theme_axis_rotate}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

## Adding a plot title and subtitle

Titles are stripped from the base themes (they are rarely used in journal
figures), but can be added back:

```{r theme_title}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title    = "Overall Survival",
    subtitle = paste0("n = ", nrow(dta_km), " patients"),
    x        = "Years after Operation",
    y        = "Freedom from Death (%)"
  ) +
  theme_hv_poster() +
  theme(
    plot.title    = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0)
  )
```

## Expanding plot margins

Add breathing room around the panel when a figure is placed directly
on a poster without a surrounding text frame, or when axis labels clip against
a tight device boundary. The `margin()` arguments follow top / right / bottom
/ left order, matching CSS convention.

```{r theme_margin}
p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20, unit = "pt")
  )
```

# Multi-panel Figures with patchwork

The `patchwork` package composes multiple ggplot objects into a single figure.
The two patterns we reach for most are side-by-side comparisons and a main
plot stacked above a companion table or risk panel.

## Side-by-side plots

The `|` operator from patchwork places two plots next to each other in the
same row, sharing the figure height. This is the typical layout for comparing
two outcomes (trends and survival, for example) on a single manuscript
figure. Both panels are built independently and composed at the last step, so
you can adjust each one without touching the other.

```{r patchwork_side_by_side}
library(patchwork)

p_ms <- p_base +
  scale_color_manual(values = hv_ppt_palette("light"), name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()

p_km_ms <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()

p_ms | p_km_ms
```

`|` places plots side by side; `/` stacks them vertically.

## Controlling relative widths and heights

`plot_layout(widths = ...)` sets relative widths as a numeric vector:
`c(2, 1)` makes the left panel twice as wide as the right. Use `heights` the
same way for stacked layouts. Use it when one panel has a wide
y-axis label or a tall legend that throws off a 50/50 split.

```{r patchwork_layout}
(p_ms | p_km_ms) +
  plot_layout(widths = c(2, 1))   # left panel twice as wide as right
```

## Stacking a plot above a risk table

A common pattern with survival curves is to pair the plot with a
numbers-at-risk panel. `hv_survival()` stores the risk table as a data
frame at `km$tables$risk`, with columns `strata`, `report_time`, `n.risk`.
Build a ggplot text panel from it, then stack with `/`.

```{r patchwork_risk_table}
#| fig-height: 5
risk_df <- km$tables$risk

rt_panel <- ggplot(risk_df,
                   aes(x = report_time, y = factor(strata),
                       label = n.risk)) +
  geom_text(size = 3) +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = NULL) +
  theme_hv_poster() +
  theme(
    axis.line  = element_blank(),
    axis.ticks = element_blank()
  )

p_km_ms / rt_panel +
  plot_layout(heights = c(4, 1))
```

## Shared axis labels and panel tags

`plot_annotation()` adds a shared title or panel tags (A, B, C...) across all
panels. Tags are required by most journals for multi-panel figures and are
referenced in the caption as "Panel A shows..."; setting
`tag_levels = "A"` generates them automatically. The `&` operator applies a
shared `theme()` call to every panel at once, so you only need to set
`plot.tag` formatting in one place.

```{r patchwork_annotation}
(p_ms | p_km_ms) +
  plot_annotation(
    title = "Figure 1. Outcomes after cardiac surgery",
    tag_levels = "A"
  ) &
  theme(plot.tag = element_text(size = 12, face = "bold"))
```

## Saving a patchwork composite

Assign the composed object to a variable and pass it to `ggsave()`. For
PowerPoint, save each panel individually with `save_ppt()`, because patchwork flattens
everything into a single raster, so the shapes and text are no longer editable
in PowerPoint.

```{r patchwork_save}
#| eval: false
combined <- (p_ms | p_km_ms) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 12, face = "bold"))

ggsave(
  filename = here::here("graphs", "fig1_combined.pdf"),
  plot     = combined,
  width    = 14,
  height   = 7
)
```
