---
title: "hvtiPlotR Plot Functions"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{hvtiPlotR Plot Functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(ggplot2)
library(hvtiPlotR)
```

Each hvtiPlotR plot function returns a bare ggplot object with no colour
scales, axis labels, or theme applied. The caller composes those with the usual
+ operator. See the companion vignette "Decorating and Saving hvtiPlotR Plots"
for full coverage of scale_, labs(), annotate(), themes, and ggsave() patterns.

# Template Reference Map

The table below maps each hvtiPlotR function to the original SAS and R
templates it ports. Functions marked with — have no direct predecessor and were
designed specifically for this package.

| hvtiPlotR Function | SAS Template(s) | R Template(s) |
|---|---|---|
| `mirror_histogram()` | — | — |
| `stacked_histogram()` | — | — |
| `covariate_balance()` | — | — |
| `goodness_followup()` | `tp.dp.goodness_followup.*` | — |
| `goodness_event_plot()` | `tp.dp.goodness_event.*` | — |
| `survival_curve()` | `tp.hp.dead.sas` (basic) | `tp.hp.dead.number_risk.R` |
| `hazard_plot()` | `tp.hp.dead.*`, `tp.hp.event.weighted.sas`, `tp.hp.repeated*.sas`, `tp.hp.numtreat.survdiff.matched.sas` | `tp.hp.dead.number_risk.R` |
| `survival_difference_plot()` | `tp.hp.dead.life-gained.sas`, `tp.hp.numtreat.survdiff.matched.sas` | — |
| `nnt_plot()` | `tp.hp.numtreat.survdiff.matched.sas` | — |
| `nonparametric_curve_plot()` | `tp.np.*.avrg_curv.*`, `tp.np.*.u.trend.*`, `tp.np.*.double.*`, `tp.np.*.mult.*`, `tp.np.*.phases.*`, `tp.np.z0axdpo.*` | — |
| `nonparametric_ordinal_plot()` | `tp.np.*.ordinal.*` | — |
| `eda_plot()` | — | — |
| `spaghetti_plot()` | — | — |
| `trends_plot()` | — | — |
| `longitudinal_counts_plot()` | `tp.dp.longitudinal_patients_measures.*` | — |
| `longitudinal_counts_table()` | `tp.dp.longitudinal_patients_measures.*` | — |
| `alluvial_plot()` | — | — |
| `upset_plot()` | — | — |

: {tbl-colwidths="[28,44,28]"}

# Mirrored Propensity Score Histogram

A common figure in propensity-matched analyses is the mirrored histogram, which displays the propensity score distributions for two treatment groups before and after matching. The **hvtiPlotR** package provides the `mirror_histogram()` function to generate this figure.

The function accepts a data frame with columns for the propensity score, group indicator, and match indicator. The `sample_mirror_histogram_data()` function generates example data suitable for testing.

```{r}
#| label: mirror_histogram
#| caption: Mirrored Propensity Score Histogram
# Generate sample data for the mirrored histogram
mirror_dta <- sample_mirror_histogram_data(n = 2000)

# Generate the mirrored histogram
mhist <- mirror_histogram(
  data = mirror_dta,
  score_col = "prob_t",
  group_col = "tavr",
  match_col = "match",
  group_levels = c(0, 1),
  group_labels = c("SAVR", "TF-TAVR"),
  matched_value = 1,
  score_multiplier = 100,
  binwidth = 5,
  alpha = 0.8
)

# Display the plot
mhist$plot
```

The lighter bars show the full (pre-match) propensity score distribution for each group, while the darker overlaid bars show the matched subset. The upper panel corresponds to the first group label and the lower panel to the second.

The function also returns diagnostics summarising group counts and standardized mean differences (SMD) before and after matching:

```{r}
#| label: mirror_histogram_diagnostics
#| caption: Mirrored Histogram Diagnostics
# Standardized mean difference before matching
mhist$diagnostics$smd_before

# Standardized mean difference after matching
mhist$diagnostics$smd_matched

# Group counts before matching
mhist$diagnostics$group_counts_before

# Group counts after matching
mhist$diagnostics$group_counts_matched
```

# Stacked Histogram

A common exploratory figure is the stacked histogram, which shows how the composition of a numeric variable changes over time or across another grouping dimension. The **hvtiPlotR** package provides the `stacked_histogram()` function to generate this figure.

The function returns a bare `ggplot` object — no colour scales, axis labels, or theme are applied — so the caller can add those freely with the usual `+` operator.

The `sample_stacked_histogram_data()` function generates a reproducible example dataset with `year` and `category` columns.

```{r}
#| label: stacked_histogram_data
# Generate sample data
hist_dta <- sample_stacked_histogram_data(n_years = 20, start_year = 2000,
                                           n_categories = 3)
head(hist_dta)
```

## Count histogram

The default `position = "stack"` shows raw counts within each bin, equivalent to the `plot.sas` frequency histogram.

```{r}
#| label: stacked_histogram_count
# Build the bare plot
p_count <- stacked_histogram(hist_dta, x_col = "year", group_col = "category")

# Layer on colour scales, labels, and a theme
p_count +
  scale_fill_brewer(palette = "Set1", name = "Category") +
  scale_color_brewer(palette = "Set1", name = "Category") +
  labs(x = "Year", y = "Count") +
  hvti_theme("manuscript")
```

## Proportion (fill) histogram

Setting `position = "fill"` rescales each bin so the bars sum to 1, making it easy to compare the relative composition across years.

```{r}
#| label: stacked_histogram_fill
# Build the proportional variant
p_fill <- stacked_histogram(hist_dta, x_col = "year", group_col = "category",
                             position = "fill")

# Use manual colours and custom legend labels
p_fill +
  scale_fill_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    labels = c("1" = "Group A", "2" = "Group B", "3" = "Group C"),
    name   = "Category"
  ) +
  scale_color_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    guide  = "none"
  ) +
  labs(x = "Year", y = "Proportion") +
  hvti_theme("manuscript")
```

## Saving

```{r}
#| label: stacked_histogram_save
#| eval: false
p_final <- p_fill +
  scale_fill_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    name   = "Category"
  ) +
  scale_color_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    guide  = "none"
  ) +
  labs(x = "Year", y = "Proportion") +
  hvti_theme("manuscript")

ggsave(
  filename = "../graphs/stacked_histogram.pdf",
  plot     = p_final,
  width    = 11,
  height   = 8
)
```

# Goodness-of-Follow-Up Plot

The goodness-of-follow-up plot is a standard quality-control figure in
longitudinal outcome analyses. It displays each patient as a point at their
operation date (x-axis) and follow-up duration (y-axis), with a short vertical
tick below each point. A dashed diagonal line marks the maximum potential
follow-up given the study start, study end, and follow-up closing date. The
**hvtiPlotR** package provides `goodness_followup()` to build this figure.

The function returns a bare `ggplot` object with no colour, shape, axis, or
label scales applied — those are added by the caller with standard `ggplot2`
modifiers.

## Sample data

```{r}
#| label: gfup_data
gfup_dta <- sample_goodness_followup_data(n = 300, seed = 42)
head(gfup_dta)
```

## Death follow-up plot

```{r}
#| label: gfup_basic
gfup <- goodness_followup(
  data        = gfup_dta,
  origin_year = 1990,
  study_start = as.Date("1990-01-01"),
  study_end   = as.Date("2019-12-31"),
  close_date  = as.Date("2021-08-06"),
  alpha       = 0.8
)

# Bare plot — no scales or labels yet
gfup$death_plot
```

## Adding scales, labels, and annotations

Scale, label, and annotation layers are composed with the usual `+` operator.
`scale_color_manual()` and `scale_shape_manual()` map the binary alive/dead
state to colours and point shapes. `annotate()` places group-identifying text
directly on the panel.

```{r}
#| label: gfup_styled
library(RColorBrewer)

gfup$death_plot +
  # Colour alive = blue, dead = red (Set1 palette positions 2 and 1)
  scale_color_manual(
    values   = brewer.pal(3, "Set1")[c(2, 1)],
    labels   = c("Alive", "Dead"),
    na.value = "black",
    drop     = FALSE
  ) +
  scale_shape_manual(
    values = c(1, 4),
    labels = c("Alive", "Dead")
  ) +
  # Axis tick placement
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33,   3)) +
  # Clip the panel to the study window
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  # Axis and legend labels
  labs(
    x     = "Operation Date",
    y     = "Follow-up (years)",
    color = "Status",
    shape = "Status"
  ) +
  # Annotate directly on the panel
  annotate("text", x = 1993, y = 31, label = "Alive at close",
           hjust = 0, size = 3.5) +
  annotate("text", x = 1993, y = 28, label = "Deceased",
           hjust = 0, size = 3.5, color = brewer.pal(3, "Set1")[1]) +
  theme(legend.position = "none")
```

The diagonal dashed line represents the maximum potential follow-up. Points
sitting above the line indicate patients with longer follow-up than expected
from the study window, typically due to passive surveillance supplementing
active cross-sectional follow-up.

## Saving

```{r}
#| label: gfup_save
#| eval: false
gfup_final <- gfup$death_plot +
  scale_color_manual(
    values   = brewer.pal(3, "Set1")[c(2, 1)],
    labels   = c("Alive", "Dead"),
    na.value = "black",
    drop     = FALSE
  ) +
  scale_shape_manual(values = c(1, 4), labels = c("Alive", "Dead")) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(x = "Operation Date", y = "Follow-up (years)",
       color = "Status", shape = "Status") +
  annotate("text", x = 1993, y = 31, label = "Alive at close",
           hjust = 0, size = 3.5) +
  annotate("text", x = 1993, y = 28, label = "Deceased",
           hjust = 0, size = 3.5, color = brewer.pal(3, "Set1")[1]) +
  theme(legend.position = "none") +
  hvti_theme("manuscript")

ggsave(
  filename = "../graphs/dp_goodness-of-followup.pdf",
  plot     = gfup_final,
  height   = 6,
  width    = 6
)
```

## Non-fatal event panel

When the dataset includes a non-fatal competing event (e.g. relapse,
reoperation), pass `event_col`, `event_time_col`, and optionally
`death_for_event_col` to generate a second panel alongside the death panel.

```{r}
#| label: gfup_event_data
gfup_event_dta <- sample_goodness_followup_data(n = 300, seed = 42)
```

```{r}
#| label: gfup_event_panel
goodness_event_plot(
  gfup_event_dta,
  event_col           = "ev_event",
  event_time_col      = "iv_event",
  death_for_event_col = "deads",
  event_levels        = c("No event", "Relapse", "Death"),
  origin_year         = 1990,
  study_start         = as.Date("1990-01-01"),
  study_end           = as.Date("2019-12-31"),
  close_date          = as.Date("2021-08-06"),
  alpha               = 0.8
) +
  scale_color_manual(
    values = c("No event" = "blue", "Relapse" = "green3", "Death" = "red"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("No event" = 1L, "Relapse" = 2L, "Death" = 4L),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(
    x = "Operation Date",
    y = "Follow-up (years)",
    color = "Event", shape = "Event"
  ) +
  annotate("text", x = 1993, y = 31,
           label = "Systematic follow-up", hjust = 0, size = 3.5) +
  theme(legend.position = c(0.85, 0.15)) +
  hvti_theme("manuscript")
```

The death panel (from `goodness_followup()`) and event panel (from
`goodness_event_plot()`) share the same diagonal reference line and can be
saved individually with `ggsave()`.

# Covariate Balance Plot

The covariate balance plot is the standard quality-control figure for
propensity score matching and IPTW analyses. Each covariate occupies a
labelled row; points show the standardized mean difference (SMD) for each
comparison group (e.g. before and after matching). A solid vertical line marks
zero balance; dotted vertical lines mark an imbalance threshold (default ±10%).

The **hvtiPlotR** package provides `covariate_balance()` to build this figure.
It returns a bare `ggplot` object — no colour, shape, axis labels, or theme
applied — so all styling is added with the usual `+` operator.

Input data must be in **long format**: one row per covariate × group
combination with columns for the covariate name, the group label, and the
numeric SMD value.

## Sample data

```{r}
#| label: cov_balance_data
dta_cb <- sample_covariate_balance_data(n_vars = 12)
head(dta_cb)
```

## Bare plot

```{r}
#| label: cov_balance_bare
covariate_balance(dta_cb, alpha = 0.8)
```

## Adding colour, shape, and axis scales

```{r}
#| label: cov_balance_scales
covariate_balance(dta_cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-45, 35),
    breaks = seq(-40, 30, 10)
  ) +
  labs(
    x = "Standardized difference (%)",
    y = ""
  ) +
  theme(legend.position = c(0.20, 0.95))
```

## Adding directional annotations and theme

`annotate()` places explanatory text directly on the panel to indicate which
direction of imbalance favours each group.

```{r}
#| label: cov_balance_annotated
n_vars <- length(unique(dta_cb$variable))

covariate_balance(dta_cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-45, 35),
    breaks = seq(-40, 30, 10)
  ) +
  labs(
    x = "Standardized difference: Group A \u2013 Group B (%)",
    y = ""
  ) +
  annotate("text", x = -32, y = 0.4,        label = "More likely Group B", size = 4) +
  annotate("text", x =  22, y = n_vars + 1, label = "More likely Group A", size = 4) +
  theme(legend.position = c(0.20, 0.95)) +
  hvti_theme("manuscript")
```

## Controlling covariate order

Pass `var_levels` to control the bottom-to-top display order of covariates.

```{r}
#| label: cov_balance_order
covariate_balance(
  dta_cb,
  var_levels = rev(unique(dta_cb$variable)),
  alpha      = 0.8
) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  hvti_theme("manuscript")
```

## Saving

```{r}
#| label: cov_balance_save
#| eval: false
cb_final <- covariate_balance(dta_cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(-45, 35), breaks = seq(-40, 30, 10)) +
  labs(x = "Standardized difference (%)", y = "") +
  annotate("text", x = -32, y = 0.4,        label = "More likely Group B", size = 4) +
  annotate("text", x =  22, y = n_vars + 1, label = "More likely Group A", size = 4) +
  theme(legend.position = c(0.20, 0.95)) +
  hvti_theme("manuscript")

ggsave(
  filename = "../graphs/lp_cov-balance.pdf",
  plot     = cb_final,
  height   = 7,
  width    = 8
)
```

# Kaplan-Meier Survival Curve

`survival_curve()` estimates the Kaplan-Meier product-limit survival function
and returns five companion bare plots matching the SAS `%kaplan` macro output
(`PLOTS`, `PLOTC`, `PLOTH`, `PLOTL`), plus tidy data frames for tables and
further computation. Confidence intervals use the logit transform with a
default confidence level of 0.6827 (one standard deviation), reproducing the
SAS macro default.

## Sample data

```{r}
#| label: km_data
dta_km <- sample_survival_data(n = 500, seed = 42)
head(dta_km)
```

## Survival curve (PLOTS=1)

```{r}
#| label: km_result
km <- survival_curve(dta_km, alpha = 0.8)

# Bare plot — no scales or labels yet
km$survival_plot
```

## Adding scales, labels, and annotations

```{r}
#| label: km_styled
km$survival_plot +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(
    breaks = seq(0, 100, 20),
    labels = function(x) paste0(x, "%")
  ) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    x     = "Years after Operation",
    y     = "Freedom from Death (%)",
    title = "Overall Survival"
  ) +
  annotate("text", x = 1, y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  hvti_theme("manuscript")
```

## Numbers at risk and report table

```{r}
#| label: km_tables
km$risk_table
km$report_table
```

## Saving

```{r}
#| label: km_save
#| eval: false
km_final <- km$survival_plot +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  hvti_theme("manuscript")

ggsave("../graphs/km_survival.pdf", km_final, width = 8, height = 6)
```

## Stratified analysis

```{r}
#| label: km_strata_data
dta_km_s <- sample_survival_data(
  n             = 500,
  strata_levels = c("Type A", "Type B"),
  hazard_ratios = c(1, 1.4),
  seed          = 42
)

km_s <- survival_curve(dta_km_s, strata_col = "valve_type", alpha = 0.8)

km_s$survival_plot +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_fill_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Survival by Valve Type") +
  theme(legend.position = c(0.15, 0.20)) +
  hvti_theme("manuscript")
```

## Cumulative hazard (PLOTC=1)

```{r}
#| label: km_cumhaz
km$cumhaz_plot +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Cumulative Hazard H(t)",
       title = "Nelson-Aalen Cumulative Hazard") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  hvti_theme("manuscript")
```

## Log-log survival plot (Weibull/PH check)

Parallel lines across strata indicate proportional hazards.

```{r}
#| label: km_loglog
km_s$loglog_plot +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  labs(x = "log(Years after Operation)", y = "log(-log S(t))",
       title = "Log-Log Survival — Proportional-Hazards Check") +
  hvti_theme("manuscript")
```

## Hazard rate (PLOTH=1)

The raw point estimates are noisy; add `geom_smooth()` for a publication-ready
smoothed hazard curve.

```{r}
#| label: km_hazard
km$hazard_plot +
  geom_smooth(
    aes(x = mid_time, y = hazard, color = strata),
    method = "loess", se = FALSE, span = 0.6
  ) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Instantaneous Hazard",
       title = "Hazard Rate") +
  hvti_theme("manuscript")
```

## Integrated survivorship / restricted mean survival (PLOTL=1)

```{r}
#| label: km_life
km$life_plot +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation",
       y = "Restricted Mean Survival (years)",
       title = "Integral of Survivorship") +
  hvti_theme("manuscript")
```

# EDA Barplots and Scatterplots

The **hvtiPlotR** package provides `eda_plot()` for exploratory data analysis
of all variables in a dataset against a reference time axis. It replicates the
`Function_DataPlotting()` workflow from `tp.dp.EDA_barplots_scatterplots.R`
and `tp.dp.EDA_barplots_scatterplots_varnames.R`, replacing base-R graphics
with composable `ggplot2` objects.

Three helper functions support the workflow:

- `eda_classify_var()` — detects whether a column is continuous (`"Cont"`),
  numeric-categorical (`"Cat_Num"`), or character-categorical (`"Cat_Char"`),
  matching the `UniqueLimit` logic from the template.
- `eda_select_vars()` — subsets and reorders columns by name or
  space-separated string, replacing the `Order_Variables()` /
  `Mod_Data <- dta[, Order_Var]` pattern from the varnames template.
- `sample_eda_data()` — generates a reproducible mixed-type dataset for
  demonstration.

`eda_plot()` always returns a bare `ggplot` object. Colour scales, axis
labels, annotations, and the `hvti_theme()` are added by the caller.

## Sample data

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)

# Inspect auto-detected types for each column
sapply(dta_eda, eda_classify_var)
```

## Binary categorical: count barplot

Numeric 0/1 columns are classified as `"Cat_Num"`. `NA` values appear as an
explicit `"(Missing)"` fill level so they can be coloured with
`scale_fill_manual()`. The `y_label` argument sets the plot title and
fill-legend name in place of the raw column name.

```{r}
#| label: eda_binary_count
eda_plot(dta_eda, x_col = "year", y_col = "male",
         y_label = "Sex") +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick", "(Missing)" = "grey80"),
    labels = c("0" = "Female", "1" = "Male", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  hvti_theme("manuscript")
```

## Binary categorical: percentage barplot

Setting `show_percent = TRUE` switches `geom_bar()` to `position = "fill"`.

```{r}
#| label: eda_binary_percent
eda_plot(dta_eda, x_col = "year", y_col = "cabg",
         y_label = "Concomitant CABG", show_percent = TRUE) +
  scale_fill_manual(
    values = c("0" = "grey70", "1" = "steelblue", "(Missing)" = "grey90"),
    labels = c("0" = "No CABG", "1" = "CABG", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Surgery Year", y = "Proportion") +
  hvti_theme("manuscript")
```

## Ordinal and multi-level categorical

```{r}
#| label: eda_ordinal
eda_plot(dta_eda, x_col = "year", y_col = "nyha",
         y_label = "Preoperative NYHA Class") +
  scale_fill_brewer(
    palette = "RdYlGn", direction = -1,
    labels  = c("1" = "I", "2" = "II", "3" = "III", "4" = "IV",
                "(Missing)" = "Missing"),
    name    = "NYHA"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  hvti_theme("manuscript")
```

## Character categorical

```{r}
#| label: eda_char_cat
eda_plot(dta_eda, x_col = "year", y_col = "valve_morph",
         y_label = "Valve Morphology") +
  scale_fill_manual(
    values = c(Bicuspid   = "steelblue",
               Tricuspid  = "firebrick",
               Unicuspid  = "goldenrod3",
               "(Missing)" = "grey80"),
    name = "Morphology"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  hvti_theme("manuscript")
```

## Continuous: scatter + LOESS

Continuous columns produce a scatter plot with a LOESS smoother overlay.
Where `y_col` is `NA`, a rug mark is drawn on the x-axis.

```{r}
#| label: eda_continuous
eda_plot(dta_eda, x_col = "op_years", y_col = "ef",
         y_label = "Ejection Fraction (%)") +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       caption = "Tick marks on x-axis: observations with missing EF") +
  hvti_theme("manuscript")
```

## Variable selection and labels (varnames template pattern)

`eda_select_vars()` subsets columns by name, matching the `Var_CatList` /
`Var_ContList` + `Order_Variables()` workflow. A named vector whose names are
column names and whose values are human-readable labels drives a single
`lapply()` loop.

```{r}
#| label: eda_varnames_binary
bin_vars <- c(male = "Sex (Male)", cabg = "Concomitant CABG")
sub_bin  <- eda_select_vars(dta_eda, c("year", names(bin_vars)))

p_bin <- lapply(names(bin_vars), function(cn) {
  eda_plot(sub_bin, x_col = "year", y_col = cn,
           y_label = bin_vars[[cn]]) +
    scale_fill_brewer(palette = "Set1", direction = -1, name = NULL) +
    scale_x_discrete(breaks = seq(2005, 2020, 5)) +
    labs(x = "Surgery Year", y = "Count") +
    hvti_theme("manuscript")
})
p_bin[[1]]
p_bin[[2]]
```

```{r}
#| label: eda_varnames_ordinal
cat_vars <- c(nyha        = "NYHA Class",
              valve_morph = "Valve Morphology")
sub_cat <- eda_select_vars(dta_eda, c("year", names(cat_vars)))

p_cat <- lapply(names(cat_vars), function(cn) {
  eda_plot(sub_cat, x_col = "year", y_col = cn,
           y_label = cat_vars[[cn]]) +
    scale_fill_brewer(palette = "Set2", name = NULL) +
    scale_x_discrete(breaks = seq(2005, 2020, 5)) +
    labs(x = "Surgery Year", y = "Count") +
    hvti_theme("manuscript")
})
p_cat[[1]]
p_cat[[2]]
```

```{r}
#| label: eda_varnames_continuous
cont_vars <- c(ef        = "Ejection Fraction (%)",
               lv_mass   = "LV Mass Index (g/m\u00b2)",
               peak_grad = "Peak Gradient (mmHg)")
sub_cont <- eda_select_vars(dta_eda, c("op_years", names(cont_vars)))

p_cont <- lapply(names(cont_vars), function(cn) {
  eda_plot(sub_cont, x_col = "op_years", y_col = cn,
           y_label = cont_vars[[cn]]) +
    scale_colour_manual(values = c("steelblue"), guide = "none") +
    scale_x_continuous(breaks = seq(0, 15, 5)) +
    labs(x = "Years from First Surgery Year") +
    hvti_theme("manuscript")
})
p_cont[[1]]
p_cont[[2]]
p_cont[[3]]
```

## Saving EDA plots

`gridExtra::marrangeGrob()` arranges multiple plots into a grid and
`ggsave()` writes each page to a separate PDF file.

```{r}
#| label: eda_save
#| eval: false
all_plots <- c(p_bin, p_cat, p_cont)
per_page  <- 9L  # 3 x 3 grid

for (pg in seq(1, length(all_plots), by = per_page)) {
  idx  <- seq(pg, min(pg + per_page - 1L, length(all_plots)))
  grob <- gridExtra::marrangeGrob(all_plots[idx], nrow = 3, ncol = 3)
  ggsave(
    filename = sprintf("../graphs/eda_page%02d.pdf", ceiling(pg / per_page)),
    plot     = grob,
    width    = 14,
    height   = 14
  )
}
```

# Alluvial (Sankey) Plot

`alluvial_plot()` produces an alluvial (Sankey-style) diagram using
`ggalluvial`. Each row of the input is a unique combination of axis values
with an associated patient count; flows are drawn proportional to that count.
Ports `tp.dp.female_bicus_preAR_sankey.R`.

## Sample data

```{r}
#| label: alluvial_data
dta_al  <- sample_alluvial_data(n = 300, seed = 42)
axes    <- c("pre_ar", "procedure", "post_ar")
head(dta_al)
```

## Bare plot

```{r}
#| label: alluvial_bare
alluvial_plot(dta_al, axes = axes, y_col = "freq")
```

## Fill flows by pre-operative grade

```{r}
#| label: alluvial_filled
alluvial_plot(dta_al, axes = axes, y_col = "freq",
              fill_col = "pre_ar") +
  scale_fill_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    name = "Pre-op AR"
  ) +
  scale_colour_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c("Pre-op AR", "Procedure", "Post-op AR"),
    expand = c(0.05, 0.05)
  ) +
  labs(y = "Patients (n)",
       title = "AV Regurgitation: Pre- to Post-operative") +
  hvti_theme("manuscript")
```

## Two-axis before / after comparison

```{r}
#| label: alluvial_two_axis
alluvial_plot(
  dta_al,
  axes        = c("pre_ar", "post_ar"),
  y_col       = "freq",
  fill_col    = "pre_ar",
  axis_labels = c("Pre-operative", "Post-operative")
) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1,
                    name = "AR Grade") +
  scale_colour_brewer(palette = "RdYlGn", direction = -1,
                      guide = "none") +
  annotate("text", x = 1.5, y = 250,
           label = "Improvement after surgery",
           size = 3.5, fontface = "italic") +
  labs(y = "Patients (n)",
       title = "AV Regurgitation Before and After Surgery") +
  hvti_theme("manuscript")
```

## Saving

```{r}
#| label: alluvial_save
#| eval: false
p_al <- alluvial_plot(dta_al, axes = axes, y_col = "freq",
                      fill_col = "pre_ar") +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  scale_colour_brewer(palette = "RdYlGn", direction = -1, guide = "none") +
  labs(y = "Patients (n)") +
  hvti_theme("manuscript")

ggsave("../graphs/alluvial.pdf", p_al, width = 8, height = 6)
```

# Hazard Plot

`hazard_plot()` plots pre-computed parametric survival, hazard, or
cumulative-hazard curves from a fitted Weibull (or other parametric) model,
with optional Kaplan-Meier empirical overlay and population life-table
reference. It ports the entire `tp.hp.dead.*` SAS template family.

The input data comes from two sources that map directly to the SAS output:

| Column set | SAS dataset | R function |
|---|---|---|
| Parametric prediction grid | `predict` (`SSURVIV`, `SCLLSURV`, `SCLUSURV`) | `sample_hazard_data()` |
| KM empirical overlay | `plout` (`CUM_SURV`, `CL_LOWER`, `CL_UPPER`) | `sample_hazard_empirical()` |
| Population life table | `smatched` | `sample_life_table()` |

## Sample data

```{r}
#| label: hazard_data
dat_hp  <- sample_hazard_data(n = 500, time_max = 10)
emp_hp  <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)
head(dat_hp)
```

## Survival curve with KM overlay

The most common `tp.hp.dead.sas` output: smooth parametric survival with
confidence band plus discrete KM empirical circles and error bars.

```{r}
#| label: hazard_survival
hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)") +
  hvti_theme("manuscript")
```

## Hazard rate curve

Switching `estimate_col` to `"hazard"` (and the matching CI columns) shows the
instantaneous hazard rate (%/year) instead of survival.

```{r}
#| label: hazard_rate
hazard_plot(
  dat_hp,
  estimate_col = "hazard",
  lower_col    = "haz_lower",
  upper_col    = "haz_upper"
) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 30),
                     labels = function(x) paste0(x, "%/yr")) +
  labs(x = "Years", y = "Hazard (%/year)") +
  hvti_theme("manuscript")
```

## Cumulative hazard

Used for readmission and repeated-event analyses (`tp.hp.event.weighted.sas`,
`tp.hp.repeated_events.sas`). The `"cumhaz"` column equals `-log(S) * 100`.

```{r}
#| label: hazard_cumhaz
hazard_plot(
  dat_hp,
  estimate_col  = "cumhaz",
  lower_col     = "cumhaz_lower",
  upper_col     = "cumhaz_upper"
) +
  scale_colour_manual(values = c("darkorange"), guide = "none") +
  scale_fill_manual(values = c("darkorange"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Cumulative Hazard (%)") +
  hvti_theme("manuscript")
```

## Stratified by group

Pass `group_col` to compare two or more groups
(`tp.hp.dead.tkdn.stratified.sas`).

```{r}
#| label: hazard_stratified
dat_strat <- sample_hazard_data(
  n = 400, time_max = 10,
  groups = c("No Takedown" = 1.0, "Takedown" = 0.65)
)
emp_strat <- sample_hazard_empirical(
  n = 400, time_max = 10, n_bins = 6,
  groups = c("No Takedown" = 1.0, "Takedown" = 0.65)
)

hazard_plot(
  dat_strat,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  group_col     = "group",
  empirical     = emp_strat,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(
    values = c("No Takedown" = "steelblue", "Takedown" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("No Takedown" = "steelblue", "Takedown" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Esophagostomy", y = "Survival (%)") +
  hvti_theme("manuscript")
```

## Life-table overlay

`tp.hp.dead.age_with_population_life_table.sas` and
`tp.hp.dead.uslife.stratifed.sas` overlay US population life-table survival
(dashed lines) on age-stratified study curves. Pass the life-table data frame
to `reference` and set `ref_group_col` to vary the linetype by age group.

```{r}
#| label: hazard_lifetable
dat_age <- sample_hazard_data(
  n = 600, time_max = 12,
  groups = c("<65" = 0.5, "65\u201380" = 1.0, "\u226580" = 1.8)
)
emp_age <- sample_hazard_empirical(
  n = 600, time_max = 12, n_bins = 6,
  groups = c("<65" = 0.5, "65\u201380" = 1.0, "\u226580" = 1.8)
)
lt <- sample_life_table(
  age_groups = c("<65", "65\u201380", "\u226580"),
  age_mids   = c(55, 72, 85),
  time_max   = 12
)

hazard_plot(
  dat_age,
  estimate_col     = "survival",
  lower_col        = "surv_lower",
  upper_col        = "surv_upper",
  group_col        = "group",
  empirical        = emp_age,
  emp_lower_col    = "lower",
  emp_upper_col    = "upper",
  reference        = lt,
  ref_estimate_col = "survival",
  ref_group_col    = "group"
) +
  scale_colour_manual(
    values = c("<65" = "steelblue", "65\u201380" = "forestgreen",
               "\u226580" = "firebrick"),
    name   = "Age group"
  ) +
  scale_fill_manual(
    values = c("<65" = "steelblue", "65\u201380" = "forestgreen",
               "\u226580" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)",
       caption = "Dashed lines: US population life table") +
  hvti_theme("manuscript")
```

# Survival Difference (Life-Gained) Plot

`survival_difference_plot()` plots the difference `S_2(t) - S_1(t)` between
two groups over time, with an optional confidence band. This ports
`tp.hp.dead.life-gained.sas` (HAZDIFL macro output).

## Sample data

```{r}
#| label: surv_diff_data
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("Control" = 1.0, "Treatment" = 0.70)
)
head(diff_dat)
```

## Survival difference curve

```{r}
#| label: surv_diff_plot
survival_difference_plot(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                      colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-5, 40),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival Difference (%)") +
  hvti_theme("manuscript")
```

## Multiple comparisons

Combine several two-group differences into one long data frame and use
`group_col` to overlay them.

```{r}
#| label: surv_diff_multi
d1 <- sample_survival_difference_data(
  groups = c("Medical Mgmt" = 1.0, "TF-TAVR" = 0.70), seed = 1L
)
d1$comparison <- "TF-TAVR vs Medical Mgmt"

d2 <- sample_survival_difference_data(
  groups = c("TA-TAVR" = 0.90, "TF-TAVR" = 0.70), seed = 2L
)
d2$comparison <- "TF-TAVR vs TA-TAVR"

d3 <- sample_survival_difference_data(
  groups = c("AVR" = 0.80, "TF-TAVR" = 0.70), seed = 3L
)
d3$comparison <- "TF-TAVR vs AVR"

survival_difference_plot(rbind(d1, d2, d3),
                         group_col = "comparison") +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                      colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Survival Difference (%)") +
  hvti_theme("manuscript")
```

# Number Needed to Treat (NNT) Plot

`nnt_plot()` plots the number needed to treat and absolute risk reduction over
time, porting the NNT component of `tp.hp.numtreat.survdiff.matched.sas`.

## Sample data

```{r}
#| label: nnt_data
nnt_dat <- sample_nnt_data(
  n      = 500,
  time_max = 20,
  groups = c("SVG" = 1.0, "ITA" = 0.75)
)
head(nnt_dat)
```

## NNT curve

NNT decreases over time as the treatment benefit accumulates.

```{r}
#| label: nnt_curve
nnt_plot(
  nnt_dat,
  lower_col = "nnt_lower",
  upper_col = "nnt_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) +
  labs(x = "Years", y = "Number Needed to Treat") +
  hvti_theme("manuscript")
```

## Absolute risk reduction

The same data, plotted as ARR (%) instead of NNT.

```{r}
#| label: nnt_arr
nnt_plot(
  nnt_dat,
  estimate_col = "arr",
  lower_col    = "arr_lower",
  upper_col    = "arr_upper"
) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Absolute Risk Reduction (%)") +
  hvti_theme("manuscript")
```

## Saving

```{r}
#| label: hazard_save
#| eval: false
p_hp <- hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  labs(x = "Years", y = "Survival (%)") +
  hvti_theme("manuscript")

ggsave("../graphs/hazard_survival.pdf", p_hp, width = 11.5, height = 8)
```
