---
title: "Decorating and Saving hvtiPlotR Plots"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{Decorating and Saving hvtiPlotR Plots}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(ggplot2)
library(hvtiPlotR)
```

# The Composition Pattern

Every hvtiPlotR plot is built in two steps: a constructor (`hv_*()`) that
shapes the data, followed by `plot()` that renders a bare `ggplot` object. No
colour scales, axis labels, or theme are applied by either step — you add those
by chaining layers with `+`:

```
plot(hv_*(...)) +
  scale_colour_*() +   # data colours
  scale_fill_*()   +   # fill colours
  labs()           +   # axis labels, title, caption
  annotate()       +   # text/arrows placed on the panel
  coord_cartesian() +  # viewport cropping
  theme_hv_manuscript()         # non-data formatting
```

This vignette demonstrates each decorator in turn, using `hv_trends()` and
`hv_survival()` as representative base plots.

```{r base_data}
# Trends data — multi-group continuous outcome over time
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

# KM data — survival curve
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)
```

# Themes

The **hvtiPlotR** package provides four themes via `theme_hv_*()`. The
`style` argument selects the output target.

| Style | Target |
|-------|--------|
| `"manuscript"` | Journal PDF, black-on-white |
| `"poster"` | Conference poster, slightly larger text |
| `"light_ppt"` | PowerPoint on white/light background |
| `"dark_ppt"` | PowerPoint on dark/blue background |

## Manuscript

`theme_hv_manuscript()` is the default for journal submissions -- white
background, black text, minimal chrome. Font sizes are tuned for 8.5 x 11 inch
letter-size PDFs where the figure may be reduced to a column width. Use this
whenever the figure ends up in a Word or PDF document sent to a journal.

```{r theme_hv_manuscript}
p_ms <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript()

p_ms
```

## Poster

`theme_hv_poster()` bumps the axis text and tick weights up from the manuscript
baseline and removes the panel grid, so the figure reads clearly at arm's
length on a 36 x 48 inch foam board. It is also the go-to theme for in-room
slide presentations when you are not using the PowerPoint themes -- the
larger tick labels and the boxed panel hold up on a projector better than the
manuscript variant. Pass `base_family = "sans"` (or another family) to switch
the font face.

```{r theme_hv_poster}
p_poster <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()

p_poster
```

## Light PowerPoint

`theme_hv_ppt_light()` matches slides with a white or light-grey background --
the Cleveland Clinic standard template, most default Office themes, and any
deck where the content area is light. Text and lines are dark, so the figure
reads without modification when placed on the light slide. Pair with
`save_ppt()` to insert it as editable DrawingML.

```{r theme_hv_ppt_light}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_ppt_light()
```

## Dark PowerPoint

`theme_hv_ppt_dark()` flips the palette for dark-background slides -- navy or
dark-blue gradient decks where white-on-dark text is the convention. The theme
sets a transparent plot background, so the slide's background shows through
behind the panel. The `plot.background` override in the chunk below simulates
that navy background in the rendered vignette; you would omit it when saving
to an actual `.pptx` file.

```{r theme_hv_ppt_dark}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_ppt_dark() +
  theme(plot.background = element_rect(fill = "navy", colour = "navy"))
```

# Colour Scales

`scale_colour_*` controls line and point colours; `scale_fill_*` controls filled areas (ribbons, bars). Both take the same `name` (legend title) and `guide` (legend display) arguments — set them once and both scales update.

## Manual colours

Use `scale_colour_manual()` when assigning specific brand or convention colours
to known levels.

```{r scale_colour_manual}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()
```

## ColorBrewer palettes

`scale_colour_brewer()` applies a ColorBrewer palette — perceptually uniform
and print-safe, which matters when figures go to black-and-white PDF. Use
`palette = "Set1"` for categorical data, `"RdYlGn"` for diverging, `"Blues"`
for sequential.

```{r scale_colour_brewer}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()
```

## Suppressing legends

Pass `guide = "none"` to any scale and that aesthetic drops out of the legend.
You'd do this when the axis labels or an annotation already name the group —
a redundant legend only takes up panel space.

```{r scale_no_legend}
p_base +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    guide = "none"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()
```

# Labels and Annotations

## labs()

`labs()` sets axis labels, the plot title, legend title, subtitle, and caption text. We set
these outside the constructor so you can override them per project without
touching the function itself.

```{r labs_example}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, \u03b1 = 0.6827 (1 SD)"
  ) +
  theme_hv_poster()
```

## annotate()

`annotate()` places text, segments, rectangles, or arrows at fixed data coordinates —
useful for a sample-size callout in the corner or a label pointing to an
event of interest.

```{r annotate_example}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), colour = "grey40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, colour = "grey40") +
  theme_hv_poster()
```

## coord_cartesian()

`coord_cartesian()` crops the viewport without dropping data, preserving LOESS
fits computed on the full range.

```{r coord_cartesian}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  theme_hv_poster()
```

# Saving Figures

## Manuscript PDF

Use `ggsave()` with `width = 11, height = 8.5` (US Letter landscape) for
manuscript figures. Assign the fully composed plot to a variable first so the
same object is both displayed in the session and written to disk.

```{r save_manuscript}
#| eval: false
ggsave(
  filename = "../graphs/trends_manuscript.pdf",
  plot     = p_ms,
  width    = 11,
  height   = 8.5
)
```

## Poster PDF

Poster figures are typically larger and use `theme_hv_poster()`. Adjust
dimensions to match the poster panel size.

```{r save_poster}
#| eval: false
ggsave(
  filename = "../graphs/trends_poster.pdf",
  plot     = p_poster,
  width    = 14,
  height   = 10
)
```

## PowerPoint slides

`save_ppt()` inserts ggplot objects into a PowerPoint file as **editable
DrawingML vector graphics** via the `officer` and `rvg` packages — shapes,
lines, and text remain selectable in PowerPoint after export.

Key arguments:

| Argument | Default | Notes |
|---|---|---|
| `object` | — | A single ggplot **or** a named/unnamed list of ggplots |
| `template` | `"../graphs/RD.pptx"` | Existing `.pptx` used as the slide template |
| `powerpoint` | `"../graphs/pptExample.pptx"` | Output file path |
| `slide_titles` | `"Plot"` | Character vector recycled to the number of plots |
| `layout` | `"Title and Content"` | Slide layout from the template |
| `width` / `height` | `10.1` / `5.8` | Plot area in inches |
| `left` / `top` | `0.0` / `1.2` | Position from slide edges, in inches |

Apply `theme_hv_ppt_dark()` or `theme_hv_ppt_light()` before saving to
match the slide background.

### Single slide

Build a fully themed plot -- `theme_hv_ppt_dark()` or `theme_hv_ppt_light()`
-- then pass it as `object`. The `template` argument points to an existing
`.pptx` file whose slide master and layouts carry the brand fonts and
background; `save_ppt()` appends a new slide rather than overwriting the
template.

```{r save_ppt_single}
#| eval: false
template <- system.file("ClevelandClinic.pptx", package = "hvtiPlotR")

p_ppt <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  theme_hv_ppt_dark()

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = here::here("graphs", "trends_slides.pptx"),
  slide_titles = "Temporal Trends by Group"
)
```

### Multiple slides from a list

Pass a named list of plots and a matching vector of titles to produce one slide
per plot in a single call. This is the pattern we use for batch-report decks
where each outcome gets its own slide -- the list keeps plots in order and the
names serve as a paper trail. Every plot in the list should carry the same
theme so the deck looks consistent.

```{r save_ppt_list}
#| eval: false
dta_km2 <- sample_survival_data(n = 400, seed = 99)
km2     <- hv_survival(dta_km2)

p_km_ppt <- plot(km2) +
  scale_color_manual(values = c(All = "white"), guide = "none") +
  scale_fill_manual(values  = c(All = "white"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_ppt_dark()

save_ppt(
  object       = list(trends = p_ppt, survival = p_km_ppt),
  template     = template,
  powerpoint   = here::here("graphs", "multi_slide_deck.pptx"),
  slide_titles = c("Temporal Trends by Group", "Overall Survival")
)
```

### Fixed panel placement across slides

When plots have different y-axis ranges ("0-1" vs "0-10000"), the axis-label
widths differ, which shifts the plot panel inside a fixed `ph_location()`. On
dark PPT themes with a visible panel background (black panel on a blue-gradient
slide), that shift is visually jarring — the box appears to move between slides.

Pass `panel_box = list(width, height, left, top)` to anchor the **panel content
area** at the same slide coordinates on every slide. `save_ppt()` calls
`hv_ph_location()` for each plot, measures the axis/legend/title chrome
around the panel, and adjusts per-slide placement so the panel lands at the
target rectangle regardless of label width. Axis labels then extend outside the
panel box as needed; device dimensions vary per plot.

```{r save_ppt_panel_box}
#| eval: false
save_ppt(
  object       = list(trends = p_ppt, survival = p_km_ppt),
  template     = template,
  powerpoint   = here::here("graphs", "multi_slide_deck_aligned.pptx"),
  slide_titles = c("Temporal Trends by Group", "Overall Survival"),
  panel_box    = list(width = 10, height = 5, left = 1.5, top = 1.5)
)
```

Make `panel_left` / `panel_top` large enough to leave room for the widest
axis labels in the deck — otherwise `hv_ph_location()` warns that the plot
chrome spills off the slide edge.

## Multi-panel PDF (EDA batch output)

When generating multiple plots in a loop, `patchwork::wrap_plots()` arranges
them into a grid and `ggsave()` writes each page. This is typical for EDA
batches where you have a dozen or more outcomes to inspect -- you build the
list with `lapply()`, chunk it into pages of 9, and let the loop handle
pagination. The `per_page` constant is easy to adjust if you want a 2x2 or
4x4 grid instead.

```{r save_multipanel}
#| eval: false
# Build a list of plots (e.g. from an hv_eda() lapply loop)
plot_list <- lapply(
  c("ef", "lv_mass", "peak_grad"),
  function(yv) {
    dta_eda <- sample_eda_data()
    plot(hv_eda(dta_eda, x_col = "op_years", y_col = yv,
                 y_label = yv)) +
      scale_colour_manual(values = c("steelblue"), guide = "none") +
      labs(x = "Years") +
      theme_hv_poster()
  }
)

per_page <- 9L
for (pg in seq(1, length(plot_list), by = per_page)) {
  idx     <- seq(pg, min(pg + per_page - 1L, length(plot_list)))
  pg_plot <- patchwork::wrap_plots(plot_list[idx], nrow = 3, ncol = 3)
  ggsave(
    filename = sprintf(here::here("graphs", "eda_page%02d.pdf"),
                       ceiling(pg / per_page)),
    plot     = pg_plot,
    width    = 14,
    height   = 14
  )
}
```

# Legend Positioning

ggplot2 places the legend outside the panel by default. For publication figures
we usually move it inside or drop it — the axis labels often do the
identification work already.

## Inside the panel

Pass fractional coordinates `c(x, y)` to `legend.position` inside `theme()`.
`c(0, 0)` is the bottom-left corner; `c(1, 1)` is the top-right.

```{r legend_inside}
p_base +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = NULL
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.position  = c(0.15, 0.2),        # bottom-left of panel
    legend.background = element_rect(fill = "white", colour = "grey80",
                                     linewidth = 0.3)
  )
```

## Outside the panel (explicit sides)

Set `legend.position` to `"right"`, `"left"`, `"top"`, or `"bottom"` to place
the legend outside the panel area. `"bottom"` with
`legend.direction = "horizontal"` is the most common choice for multi-group
figures where a long label would crowd a corner inside the panel.

```{r legend_outside}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.position = "bottom",             # "right" | "left" | "top" | "bottom"
    legend.direction = "horizontal"
  )
```

## Suppress all legends

Two ways to suppress legends, with slightly different scope.
`theme(legend.position = "none")` hides every legend for the plot in one
shot. `guide = "none"` on a `scale_*()` call drops only that aesthetic's
legend — handy when you want a colour legend but no shape legend (or vice
versa). Both reclaim the panel width either way.

```{r legend_none}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()
```

`guide = "none"` on every `scale_*()` call removes all legend entries. This is
preferred over `theme(legend.position = "none")` when only some aesthetics
have legends and others do not.

## Legend text and key size

The legend inherits font size from the active theme, which is often slightly
larger than needed when the legend sits inside a crowded panel. `legend.text`
controls the label font; `legend.key.size` shrinks the colour/shape swatch.
Both accept any `element_text()` or `unit()` value.

```{r legend_text}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    legend.text  = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  )
```

# Theme Overrides

`theme_hv_manuscript()` sets a complete non-data formatting baseline. Layer additional
`theme()` calls after it to adjust individual elements without touching the
rest.

## Axis text size

Override `axis.text` to resize tick labels and `axis.title` for the axis
title, independent of the rest of the theme. This is useful when a figure is
being resized for a different output target -- for example, scaling down for a
two-column journal layout where the default poster-sized text would be too
large.

```{r theme_axis_text}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    axis.text  = element_text(size = 10),   # tick labels
    axis.title = element_text(size = 12)    # axis titles
  )
```

## Removing minor grid lines

Minor grid lines (`panel.grid.minor`) add density without adding precision --
they can make a busy figure look cluttered, especially when there are many data
series. Setting `element_blank()` removes them while keeping the major grid
intact.

```{r theme_grid}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    panel.grid.minor = element_blank()
  )
```

## Rotating x-axis labels

Useful for time-point labels or long category names. The combination of
`angle = 45`, `hjust = 1`, and `vjust = 1` keeps the rotated label
right-aligned to its tick mark; without the `hjust`/`vjust` corrections the
labels will float off-center.

```{r theme_axis_rotate}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

## Adding a plot title and subtitle

Titles are stripped from the base themes (they are rarely used in journal
figures), but can be added back:

```{r theme_title}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title    = "Overall Survival",
    subtitle = paste0("n = ", nrow(dta_km), " patients"),
    x        = "Years after Operation",
    y        = "Freedom from Death (%)"
  ) +
  theme_hv_poster() +
  theme(
    plot.title    = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, colour = "grey40", hjust = 0)
  )
```

## Expanding plot margins

Add breathing room around the panel -- useful when a figure is placed directly
on a poster without a surrounding text frame, or when axis labels clip against
a tight device boundary. The `margin()` arguments follow top / right / bottom
/ left order, matching CSS convention.

```{r theme_margin}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster() +
  theme(
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20, unit = "pt")
  )
```

# Multi-panel Figures with patchwork

The `patchwork` package composes multiple ggplot objects into a single figure.
The two patterns we reach for most are side-by-side comparisons and a main
plot stacked above a companion table or risk panel.

## Side-by-side plots

The `|` operator from patchwork places two plots next to each other in the
same row, sharing the figure height. This is the typical layout for comparing
two outcomes -- trends and survival, for example -- on a single manuscript
figure. Both panels are built independently and composed at the last step, so
you can adjust each one without touching the other.

```{r patchwork_side_by_side}
library(patchwork)

p_ms <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_poster()

p_km_ms <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_poster()

p_ms | p_km_ms
```

`|` places plots side by side; `/` stacks them vertically.

## Controlling relative widths and heights

`plot_layout(widths = ...)` sets relative widths as a numeric vector --
`c(2, 1)` makes the left panel twice as wide as the right. Use `heights` the
same way for stacked layouts. This is the right tool when one panel has a wide
y-axis label or a tall legend that throws off a 50/50 split.

```{r patchwork_layout}
(p_ms | p_km_ms) +
  plot_layout(widths = c(2, 1))   # left panel twice as wide as right
```

## Stacking a plot above a risk table

A common pattern with survival curves is to pair the plot with a
numbers-at-risk panel. `hv_survival()` stores the risk table as a data
frame at `km$tables$risk` — columns `strata`, `report_time`, `n.risk`.
Build a ggplot text panel from it, then stack with `/`.

```{r patchwork_risk_table}
#| fig-height: 5
risk_df <- km$tables$risk

rt_panel <- ggplot(risk_df,
                   aes(x = report_time, y = factor(strata),
                       label = n.risk)) +
  geom_text(size = 3) +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = NULL) +
  theme_hv_poster() +
  theme(
    axis.line  = element_blank(),
    axis.ticks = element_blank()
  )

p_km_ms / rt_panel +
  plot_layout(heights = c(4, 1))
```

## Shared axis labels and panel tags

`plot_annotation()` adds a shared title or panel tags (A, B, C...) across all
panels. Tags are required by most journals for multi-panel figures and are
referenced in the caption as "Panel A shows..." -- setting
`tag_levels = "A"` generates them automatically. The `&` operator applies a
shared `theme()` call to every panel at once, so you only need to set
`plot.tag` formatting in one place.

```{r patchwork_annotation}
(p_ms | p_km_ms) +
  plot_annotation(
    title = "Figure 1. Outcomes after cardiac surgery",
    tag_levels = "A"
  ) &
  theme(plot.tag = element_text(size = 12, face = "bold"))
```

## Saving a patchwork composite

Assign the composed object to a variable and pass it to `ggsave()`. For
PowerPoint, save each panel individually with `save_ppt()` — patchwork flattens
everything into a single raster, so the shapes and text are no longer editable
in PowerPoint.

```{r patchwork_save}
#| eval: false
combined <- (p_ms | p_km_ms) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 12, face = "bold"))

ggsave(
  filename = here::here("graphs", "fig1_combined.pdf"),
  plot     = combined,
  width    = 14,
  height   = 7
)
```
