---
title: "Decorating and Saving hvtiPlotR Plots"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{Decorating and Saving hvtiPlotR Plots}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(ggplot2)
library(hvtiPlotR)
```

# The Composition Pattern

Every hvtiPlotR plot function returns a bare `ggplot` object — no colour scales, axis labels, or theme are applied. Decoration is added by chaining layers with `+`:

```
bare_plot() +
  scale_colour_*() +   # data colours
  scale_fill_*()   +   # fill colours
  labs()           +   # axis labels, title, caption
  annotate()       +   # text/arrows placed on the panel
  coord_cartesian() +  # viewport cropping
  hvti_theme()         # non-data formatting
```

This vignette demonstrates each decorator in turn, using `trends_plot()` and
`survival_curve()` as representative base plots.

```{r base_data}
# Trends data — multi-group continuous outcome over time
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base <- trends_plot(dta_trends)

# KM data — survival curve
dta_km  <- sample_survival_data(n = 500, seed = 42)
km      <- survival_curve(dta_km, alpha = 0.8)
```

# Themes

The **hvtiPlotR** package provides four themes via `hvti_theme(style)`. The
`style` argument selects the output target.

| Style | Target |
|-------|--------|
| `"manuscript"` | Journal PDF, black-on-white |
| `"poster"` | Conference poster, slightly larger text |
| `"light_ppt"` | PowerPoint on white/light background |
| `"dark_ppt"` | PowerPoint on dark/blue background |

## Manuscript

```{r theme_manuscript}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")
```

## Poster

```{r theme_poster}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("poster")
```

## Light PowerPoint

```{r theme_light_ppt}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("light_ppt")
```

## Dark PowerPoint

```{r theme_dark_ppt}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("dark_ppt") +
  theme(plot.background = element_rect(fill = "navy", colour = "navy"))
```

# Colour Scales

`scale_colour_*` controls line and point colours; `scale_fill_*` controls filled areas (ribbons, bars). Both share the same `name` (legend title) and `guide` (legend display) arguments.

## Manual colours

Use `scale_colour_manual()` when assigning specific brand or convention colours
to known levels.

```{r scale_colour_manual}
km$survival_plot +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  hvti_theme("manuscript")
```

## ColorBrewer palettes

`scale_colour_brewer()` applies a ColorBrewer palette — safe, perceptually
uniform, and print-friendly. Use `palette = "Set1"` for categorical data,
`"RdYlGn"` for diverging, `"Blues"` for sequential.

```{r scale_colour_brewer}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")
```

## Suppressing legends

Pass `guide = "none"` to any scale to remove its legend entry. Use this when
colour is self-evident from axis labels or annotations.

```{r scale_no_legend}
p_base +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    guide = "none"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")
```

# Labels and Annotations

## labs()

`labs()` sets the axis, legend, title, subtitle, and caption text. Set axis
labels here rather than inside the plot function so they can be overridden per
project.

```{r labs_example}
km$survival_plot +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, \u03b1 = 0.6827 (1 SD)"
  ) +
  hvti_theme("manuscript")
```

## annotate()

`annotate()` places text, segments, or rectangles at fixed data coordinates.
Use it for sample size callouts, phase labels, or directional arrows.

```{r annotate_example}
km$survival_plot +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), colour = "grey40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, colour = "grey40") +
  hvti_theme("manuscript")
```

## coord_cartesian()

`coord_cartesian()` crops the viewport without dropping data, preserving LOESS
fits computed on the full range.

```{r coord_cartesian}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  hvti_theme("manuscript")
```

# Saving Figures

## Manuscript PDF

Use `ggsave()` with `width = 11, height = 8.5` (US Letter landscape) for
manuscript figures. Assign the fully composed plot to a variable first so the
same object is both displayed in the session and written to disk.

```{r save_manuscript}
#| eval: false
p_ms <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  hvti_theme("manuscript")

ggsave(
  filename = "../graphs/trends_manuscript.pdf",
  plot     = p_ms,
  width    = 11,
  height   = 8.5
)
```

## Poster PDF

Poster figures are typically larger and use `hvti_theme("poster")`. Adjust
dimensions to match the poster panel size.

```{r save_poster}
#| eval: false
p_poster <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  hvti_theme("poster")

ggsave(
  filename = "../graphs/trends_poster.pdf",
  plot     = p_poster,
  width    = 14,
  height   = 10
)
```

## PowerPoint slides

`save_ppt()` from the **hvtiPlotR** package inserts a ggplot object into a
PowerPoint file as an editable vector graphic (via the `officer` and `rvg`
packages). Apply `hvti_theme("dark_ppt")` or `hvti_theme("light_ppt")` before
saving.

```{r save_ppt_example}
#| eval: false
p_ppt <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  hvti_theme("dark_ppt")

save_ppt(
  plot     = p_ppt,
  filename = "../graphs/trends_slides.pptx",
  width    = 10,
  height   = 7.5
)
```

## Multi-panel PDF (EDA batch output)

When generating multiple plots in a loop, `gridExtra::marrangeGrob()` arranges
them into a grid and `ggsave()` writes each page.

```{r save_multipanel}
#| eval: false
# Build a list of plots (e.g. from an eda_plot() lapply loop)
plot_list <- lapply(
  c("ef", "lv_mass", "peak_grad"),
  function(yv) {
    dta_eda <- sample_eda_data()
    eda_plot(dta_eda, x_col = "op_years", y_col = yv,
             y_label = yv) +
      scale_colour_manual(values = c("steelblue"), guide = "none") +
      labs(x = "Years") +
      hvti_theme("manuscript")
  }
)

per_page <- 9L
for (pg in seq(1, length(plot_list), by = per_page)) {
  idx  <- seq(pg, min(pg + per_page - 1L, length(plot_list)))
  grob <- gridExtra::marrangeGrob(plot_list[idx], nrow = 3, ncol = 3)
  ggsave(
    filename = sprintf("../graphs/eda_page%02d.pdf", ceiling(pg / per_page)),
    plot     = grob,
    width    = 14,
    height   = 14
  )
}
```
