---
title: "Decorating and Saving hvtiPlotR Plots"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{Decorating and Saving hvtiPlotR Plots}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(ggplot2)
library(hvtiPlotR)
```

# The Composition Pattern

Every hvtiPlotR plot is built in two steps: a constructor (`hvti_*()`) that
shapes the data, followed by `plot()` that renders a bare `ggplot` object. No
colour scales, axis labels, or theme are applied by either step. Decoration is
added by chaining layers with `+`:

```
plot(hvti_*(...)) +
  scale_colour_*() +   # data colours
  scale_fill_*()   +   # fill colours
  labs()           +   # axis labels, title, caption
  annotate()       +   # text/arrows placed on the panel
  coord_cartesian() +  # viewport cropping
  hvti_theme()         # non-data formatting
```

This vignette demonstrates each decorator in turn, using `hvti_trends()` and
`hvti_survival()` as representative base plots.

```{r base_data}
# Trends data — multi-group continuous outcome over time
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hvti_trends(dta_trends))

# KM data — survival curve
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hvti_survival(dta_km)
```

# Themes

The **hvtiPlotR** package provides four themes via `hvti_theme(style)`. The
`style` argument selects the output target.

| Style | Target |
|-------|--------|
| `"manuscript"` | Journal PDF, black-on-white |
| `"poster"` | Conference poster, slightly larger text |
| `"light_ppt"` | PowerPoint on white/light background |
| `"dark_ppt"` | PowerPoint on dark/blue background |

## Manuscript

```{r theme_manuscript}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")
```

## Poster

```{r theme_poster}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("poster")
```

## Light PowerPoint

```{r theme_light_ppt}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("light_ppt")
```

## Dark PowerPoint

```{r theme_dark_ppt}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("dark_ppt") +
  theme(plot.background = element_rect(fill = "navy", colour = "navy"))
```

# Colour Scales

`scale_colour_*` controls line and point colours; `scale_fill_*` controls filled areas (ribbons, bars). Both share the same `name` (legend title) and `guide` (legend display) arguments.

## Manual colours

Use `scale_colour_manual()` when assigning specific brand or convention colours
to known levels.

```{r scale_colour_manual}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  hvti_theme("manuscript")
```

## ColorBrewer palettes

`scale_colour_brewer()` applies a ColorBrewer palette — safe, perceptually
uniform, and print-friendly. Use `palette = "Set1"` for categorical data,
`"RdYlGn"` for diverging, `"Blues"` for sequential.

```{r scale_colour_brewer}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")
```

## Suppressing legends

Pass `guide = "none"` to any scale to remove its legend entry. Use this when
colour is self-evident from axis labels or annotations.

```{r scale_no_legend}
p_base +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    guide = "none"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")
```

# Labels and Annotations

## labs()

`labs()` sets the axis, legend, title, subtitle, and caption text. Set axis
labels here rather than inside the plot function so they can be overridden per
project.

```{r labs_example}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, \u03b1 = 0.6827 (1 SD)"
  ) +
  hvti_theme("manuscript")
```

## annotate()

`annotate()` places text, segments, or rectangles at fixed data coordinates.
Use it for sample size callouts, phase labels, or directional arrows.

```{r annotate_example}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), colour = "grey40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, colour = "grey40") +
  hvti_theme("manuscript")
```

## coord_cartesian()

`coord_cartesian()` crops the viewport without dropping data, preserving LOESS
fits computed on the full range.

```{r coord_cartesian}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  hvti_theme("manuscript")
```

# Saving Figures

## Manuscript PDF

Use `ggsave()` with `width = 11, height = 8.5` (US Letter landscape) for
manuscript figures. Assign the fully composed plot to a variable first so the
same object is both displayed in the session and written to disk.

```{r save_manuscript}
#| eval: false
p_ms <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  hvti_theme("manuscript")

ggsave(
  filename = "../graphs/trends_manuscript.pdf",
  plot     = p_ms,
  width    = 11,
  height   = 8.5
)
```

## Poster PDF

Poster figures are typically larger and use `hvti_theme("poster")`. Adjust
dimensions to match the poster panel size.

```{r save_poster}
#| eval: false
p_poster <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  hvti_theme("poster")

ggsave(
  filename = "../graphs/trends_poster.pdf",
  plot     = p_poster,
  width    = 14,
  height   = 10
)
```

## PowerPoint slides

`save_ppt()` inserts ggplot objects into a PowerPoint file as **editable
DrawingML vector graphics** via the `officer` and `rvg` packages — shapes,
lines, and text remain selectable in PowerPoint after export.

Key arguments:

| Argument | Default | Notes |
|---|---|---|
| `object` | — | A single ggplot **or** a named/unnamed list of ggplots |
| `template` | `"../graphs/RD.pptx"` | Existing `.pptx` used as the slide template |
| `powerpoint` | `"../graphs/pptExample.pptx"` | Output file path |
| `slide_titles` | `"Plot"` | Character vector recycled to the number of plots |
| `layout` | `"Title and Content"` | Slide layout from the template |
| `width` / `height` | `10.1` / `5.8` | Plot area in inches |
| `left` / `top` | `0.0` / `1.2` | Position from slide edges, in inches |

Apply `hvti_theme("dark_ppt")` or `hvti_theme("light_ppt")` before saving to
match the slide background.

### Single slide

```{r save_ppt_single}
#| eval: false
template <- system.file("ClevelandClinic.pptx", package = "hvtiPlotR")

p_ppt <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome (%)") +
  hvti_theme("dark_ppt")

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = here::here("graphs", "trends_slides.pptx"),
  slide_titles = "Temporal Trends by Group"
)
```

### Multiple slides from a list

Pass a named list of plots and a matching vector of titles to produce one slide
per plot in a single call.

```{r save_ppt_list}
#| eval: false
dta_km2 <- sample_survival_data(n = 400, seed = 99)
km2     <- hvti_survival(dta_km2)

p_km_ppt <- plot(km2) +
  scale_color_manual(values = c(All = "white"), guide = "none") +
  scale_fill_manual(values  = c(All = "white"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  hvti_theme("dark_ppt")

save_ppt(
  object       = list(trends = p_ppt, survival = p_km_ppt),
  template     = template,
  powerpoint   = here::here("graphs", "multi_slide_deck.pptx"),
  slide_titles = c("Temporal Trends by Group", "Overall Survival")
)
```

## Multi-panel PDF (EDA batch output)

When generating multiple plots in a loop, `gridExtra::marrangeGrob()` arranges
them into a grid and `ggsave()` writes each page.

```{r save_multipanel}
#| eval: false
# Build a list of plots (e.g. from an hvti_eda() lapply loop)
plot_list <- lapply(
  c("ef", "lv_mass", "peak_grad"),
  function(yv) {
    dta_eda <- sample_eda_data()
    plot(hvti_eda(dta_eda, x_col = "op_years", y_col = yv,
                 y_label = yv)) +
      scale_colour_manual(values = c("steelblue"), guide = "none") +
      labs(x = "Years") +
      hvti_theme("manuscript")
  }
)

per_page <- 9L
for (pg in seq(1, length(plot_list), by = per_page)) {
  idx  <- seq(pg, min(pg + per_page - 1L, length(plot_list)))
  grob <- gridExtra::marrangeGrob(plot_list[idx], nrow = 3, ncol = 3)
  ggsave(
    filename = sprintf(here::here("graphs", "eda_page%02d.pdf"),
                       ceiling(pg / per_page)),
    plot     = grob,
    width    = 14,
    height   = 14
  )
}
```

# Legend Positioning

ggplot2 places the legend outside the panel by default. For publication figures
it is often cleaner to place it inside the panel or suppress it entirely.

## Inside the panel

Pass fractional coordinates `c(x, y)` to `legend.position` inside `theme()`.
`c(0, 0)` is the bottom-left corner; `c(1, 1)` is the top-right.

```{r legend_inside}
p_base +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = NULL
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript") +
  theme(
    legend.position  = c(0.15, 0.2),        # bottom-left of panel
    legend.background = element_rect(fill = "white", colour = "grey80",
                                     linewidth = 0.3)
  )
```

## Outside the panel (explicit sides)

```{r legend_outside}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript") +
  theme(
    legend.position = "bottom",             # "right" | "left" | "top" | "bottom"
    legend.direction = "horizontal"
  )
```

## Suppress all legends

```{r legend_none}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  hvti_theme("manuscript")
```

`guide = "none"` on every `scale_*()` call removes all legend entries. This is
preferred over `theme(legend.position = "none")` when only some aesthetics
have legends and others do not.

## Legend text and key size

```{r legend_text}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript") +
  theme(
    legend.text  = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  )
```

# Theme Overrides

`hvti_theme()` sets a complete non-data formatting baseline. Layer additional
`theme()` calls after it to adjust individual elements without touching the
rest.

## Axis text size

```{r theme_axis_text}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript") +
  theme(
    axis.text  = element_text(size = 10),   # tick labels
    axis.title = element_text(size = 12)    # axis titles
  )
```

## Removing minor grid lines

```{r theme_grid}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript") +
  theme(
    panel.grid.minor = element_blank()
  )
```

## Rotating x-axis labels

Useful for time-point labels or long category names.

```{r theme_axis_rotate}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

## Adding a plot title and subtitle

Titles are stripped from the base themes (they are rarely used in journal
figures), but can be added back:

```{r theme_title}
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title    = "Overall Survival",
    subtitle = paste0("n = ", nrow(dta_km), " patients"),
    x        = "Years after Operation",
    y        = "Freedom from Death (%)"
  ) +
  hvti_theme("manuscript") +
  theme(
    plot.title    = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, colour = "grey40", hjust = 0)
  )
```

## Expanding plot margins

Add breathing room around the panel, for example when a figure is placed
directly on a poster without a surrounding text frame.

```{r theme_margin}
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("poster") +
  theme(
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20, unit = "pt")
  )
```

# Multi-panel Figures with patchwork

The `patchwork` package composes multiple ggplot objects into a single figure.
Use it to place two related plots side by side, or to stack a main plot above
a companion table or risk panel.

## Side-by-side plots

```{r patchwork_side_by_side}
library(patchwork)

p_ms <- p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  hvti_theme("manuscript")

p_km_ms <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  hvti_theme("manuscript")

p_ms | p_km_ms
```

`|` places plots side by side; `/` stacks them vertically.

## Controlling relative widths and heights

```{r patchwork_layout}
(p_ms | p_km_ms) +
  plot_layout(widths = c(2, 1))   # left panel twice as wide as right
```

## Stacking a plot above a risk table

A common pattern with survival curves is to pair the plot with a
numbers-at-risk panel. `hvti_survival()` stores the risk table as a data
frame at `km$tables$risk` — columns `strata`, `report_time`, `n.risk`.
Build a ggplot text panel from it, then stack with `/`.

```{r patchwork_risk_table}
#| fig-height: 5
risk_df <- km$tables$risk

rt_panel <- ggplot(risk_df,
                   aes(x = report_time, y = factor(strata),
                       label = n.risk)) +
  geom_text(size = 3) +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = NULL) +
  hvti_theme("manuscript") +
  theme(
    axis.line  = element_blank(),
    axis.ticks = element_blank()
  )

p_km_ms / rt_panel +
  plot_layout(heights = c(4, 1))
```

## Shared axis labels and panel tags

`plot_annotation()` adds a shared title or tags (A, B, C…) across all panels.

```{r patchwork_annotation}
(p_ms | p_km_ms) +
  plot_annotation(
    title = "Figure 1. Outcomes after cardiac surgery",
    tag_levels = "A"
  ) &
  theme(plot.tag = element_text(size = 12, face = "bold"))
```

## Saving a patchwork composite

Assign the composed object to a variable and pass it to `ggsave()`. For
PowerPoint, save each panel individually with `save_ppt()` (patchwork composites
are not editable DrawingML objects).

```{r patchwork_save}
#| eval: false
combined <- (p_ms | p_km_ms) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 12, face = "bold"))

ggsave(
  filename = here::here("graphs", "fig1_combined.pdf"),
  plot     = combined,
  width    = 14,
  height   = 7
)
```
