---
title: "hvtiPlotR Plot Functions"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{hvtiPlotR Plot Functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(ggplot2)
library(hvtiPlotR)
```

All hvtiPlotR plot functions follow a two-step workflow. First, call the
constructor (`hv_*()`) to validate and prepare data; this returns an S3
object of class `c("hv_<concept>", "hv_data")`. Then call `plot()` on that
object to obtain a bare `ggplot` with no colour scales, axis labels, or theme
applied. Compose those with the usual `+` operator. See the companion vignette
"Decorating and Saving hvtiPlotR Plots" for full coverage of `scale_()`,
`labs()`, `annotate()`, themes, and `ggsave()` patterns.

# Template Reference Map

The table below maps each hvtiPlotR constructor to the original SAS and R
templates it ports. Functions marked with — have no direct predecessor and were
designed specifically for this package. All functions have worked examples in
the sections below.

| hvtiPlotR Constructor | SAS Template(s) | R Template(s) |
|---|---|---|
| `hv_mirror_hist()` | — | `tp.lp.mirror-histogram_SAVR-TF-TAVR.R`, `tp.lp.mirror_histo_before_after_wt.R` |
| `hv_stacked()` | — | — |
| `hv_balance()` | — | `tp.lp.propen.cov_balance.R` |
| `hv_followup()` | `tp.dp.goodness_followup.*`, `tp.dp.goodness_event.*` | — |
| `hv_survival()` | `tp.hp.dead.sas` (basic) | `tp.hp.dead.number_risk.R` |
| `hazard_plot()` | `tp.hp.dead.*`, `tp.hp.event.weighted.sas`, `tp.hp.repeated*.sas`, `tp.hp.numtreat.survdiff.matched.sas`, `tp.hs.dead.*`, `tp.hs.uslife_*` | `tp.hp.dead.number_risk.R` |
| `survival_difference_plot()` | `tp.hp.dead.life-gained.sas`, `tp.hp.numtreat.survdiff.matched.sas`, `tp.hs.dead.compare_benefit.setup.sas` | — |
| `nnt_plot()` | `tp.hp.numtreat.survdiff.matched.sas` | — |
| `hv_nonparametric()` | `tp.np.*.avrg_curv.*`, `tp.np.*.u.trend.*`, `tp.np.*.double.*`, `tp.np.*.mult.*`, `tp.np.*.phases.*`, `tp.np.z0axdpo.*` | — |
| `hv_ordinal()` | `tp.np.*.ordinal.*` | — |
| `hv_eda()` | — | — |
| `hv_spaghetti()` | — | `tp.dp.spaghetti.echo.R` |
| `hv_trends()` | `tp.lp.trends.sas`, `tp.lp.trends.age.sas`, `tp.lp.trends.polytomous.sas`, `tp.rp.trends.sas` | `tp.dp.trends.R` |
| `hv_longitudinal()` | `tp.dp.longitudinal_patients_measures.*` | — |
| `hv_alluvial()` | — | — |
| `hv_sankey()` | — | PAM cluster stability analysis |
| `hv_upset()` | — | — |

Note: `hazard_plot()`, `survival_difference_plot()`, and `nnt_plot()` retain
the legacy single-call API pending migration to the two-step constructor pattern.

: {tbl-colwidths="[28,44,28]"}

# Mirrored Propensity Score Histogram

A common figure in propensity-matched analyses is the mirrored histogram, which displays the propensity score distributions for two treatment groups before and after matching. The **hvtiPlotR** package provides the `hv_mirror_hist()` constructor to prepare the data, and `plot()` to render the figure.

The constructor accepts a data frame with columns for the propensity score, group indicator, and match indicator. The `sample_mirror_histogram_data()` function generates example data suitable for testing.

Two display modes are selected by the arguments supplied to the constructor:

- **Binary-match mode** (`match_col`): upper bars show all observations before matching; overlaid bars show the matched subset. This reproduces `tp.lp.mirror-histogram_SAVR-TF-TAVR.R`.
- **Weighted IPTW mode** (`weight_col`): upper bars show raw counts; overlaid bars show per-bin IPTW weight sums. This reproduces `tp.lp.mirror_histo_before_after_wt.R`.

## Binary-match mode (SAVR vs. TF-TAVR)

```{r}
#| label: mirror_histogram_build
#| caption: Mirrored Propensity Score Histogram — binary-match mode
mirror_dta <- sample_mirror_histogram_data(n = 2000, separation = 1.5)

mh <- hv_mirror_hist(
  data             = mirror_dta,
  score_col        = "prob_t",
  group_col        = "tavr",
  match_col        = "match",
  group_levels     = c(0, 1),
  group_labels     = c("SAVR", "TF-TAVR"),
  matched_value    = 1,
  score_multiplier = 100,
  binwidth         = 5
)
```

## Bare plot

```{r}
#| label: mirror_histogram_bare
p <- plot(mh, alpha = 0.8)
p
```

## Adding scales, labels, and theme

```{r}
#| label: mirror_histogram_decorated
p +
  ggplot2::scale_fill_manual(
    values = c(
      before_g0  = "white",  matched_g0 = "green1",
      before_g1  = "white",  matched_g1 = "green4"
    ),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::annotate("text", x = 20, y =  Inf, vjust =  2,
                    label = mh$meta$group_labels[1], size = 7) +
  ggplot2::annotate("text", x = 20, y = -Inf, vjust = -1,
                    label = mh$meta$group_labels[2], size = 7) +
  ggplot2::labs(x = "Propensity Score (%)", y = "Number of Patients") +
  hv_theme("poster")
```

The lighter bars show the full (pre-match) distribution for each group; the darker overlaid bars show the matched subset. Upper panel = first group label; lower panel = second group label. `scale_y_continuous(labels = abs)` converts the internal negative counts to positive labels. `y = Inf`/`-Inf` with `vjust` anchors each annotation near the panel edge regardless of data scale, so the positions adapt automatically to different dataset sizes.

The constructor stores diagnostics with group counts and standardized mean differences (SMD) before and after matching in the `$tables$diagnostics` list:

```{r}
#| label: mirror_histogram_diagnostics
#| caption: Mirrored Histogram Diagnostics
mh$tables$diagnostics$smd_before
mh$tables$diagnostics$smd_matched
mh$tables$diagnostics$group_counts_before
mh$tables$diagnostics$group_counts_matched
```

## Weighted IPTW mode (Limited vs. Extended)

```{r}
#| label: mirror_histogram_weighted_build
#| caption: Mirrored Propensity Score Histogram — weighted IPTW mode
wt_dta <- sample_mirror_histogram_data(
  n = 2000, separation = 1.5, add_weights = TRUE
)

mh_wt <- hv_mirror_hist(
  data             = wt_dta,
  score_col        = "prob_t",
  group_col        = "tavr",
  group_levels     = c(0, 1),
  group_labels     = c("Limited", "Extended"),
  weight_col       = "mt_wt",
  score_multiplier = 100,
  binwidth         = 5
)
```

## Bare plot {#mirror-iptw-bare}

```{r}
#| label: mirror_histogram_weighted_bare
p_wt <- plot(mh_wt, alpha = 0.8)
p_wt
```

## Adding scales, labels, and theme {#mirror-iptw-decorated}

```{r}
#| label: mirror_histogram_weighted_decorated
p_wt +
  ggplot2::scale_fill_manual(
    values = c(
      before_g0   = "white", weighted_g0 = "blue",
      before_g1   = "white", weighted_g1 = "red"
    ),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::annotate("text", x = 30, y =  Inf, vjust =  2,
                    label = mh_wt$meta$group_labels[1], color = "blue", size = 5) +
  ggplot2::annotate("text", x = 30, y = -Inf, vjust = -1,
                    label = mh_wt$meta$group_labels[2], color = "red",  size = 5) +
  ggplot2::labs(x = "Propensity Score (%)", y = "#") +
  hv_theme("poster")
```

Weighted diagnostics include the effective N and weighted SMD:

```{r}
#| label: mirror_histogram_weighted_diagnostics
mh_wt$tables$diagnostics$smd_weighted
mh_wt$tables$diagnostics$effective_n_by_group
```

# Stacked Histogram

A common exploratory figure is the stacked histogram, which shows how the composition of a numeric variable changes over time or across another grouping dimension. The **hvtiPlotR** package provides the `hv_stacked()` constructor and a `plot()` method to generate this figure.

`plot()` returns a bare `ggplot` object — no colour scales, axis labels, or theme are applied — so the caller can add those freely with the usual `+` operator.

The `sample_stacked_histogram_data()` function generates a reproducible example dataset with `year` and `category` columns.

```{r}
#| label: stacked_histogram_data
# Generate sample data
hist_dta <- sample_stacked_histogram_data(n_years = 20, start_year = 2000,
                                           n_categories = 3)
head(hist_dta)
```

## Count histogram

The default `position = "stack"` shows raw counts within each bin, equivalent to the `plot.sas` frequency histogram.

```{r}
#| label: stacked_histogram_count
# Build the S3 object, then render the bare plot
sh <- hv_stacked(hist_dta, x_col = "year", group_col = "category")
p_count <- plot(sh)

# Layer on colour scales, labels, and a theme
p_count +
  scale_fill_brewer(palette = "Set1", name = "Category") +
  scale_color_brewer(palette = "Set1", name = "Category") +
  labs(x = "Year", y = "Count") +
  hv_theme("poster")
```

## Proportion (fill) histogram

Setting `position = "fill"` rescales each bin so the bars sum to 1, making it easy to compare the relative composition across years.

```{r}
#| label: stacked_histogram_fill
# Build the proportional variant
sh2   <- hv_stacked(hist_dta, x_col = "year", group_col = "category",
                       position = "fill")
p_fill <- plot(sh2)

# Use manual colours and custom legend labels
p_final <- p_fill +
  scale_fill_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    labels = c("1" = "Group A", "2" = "Group B", "3" = "Group C"),
    name   = "Category"
  ) +
  scale_color_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    guide  = "none"
  ) +
  labs(x = "Year", y = "Proportion") +
  hv_theme("poster")

p_final
```

## Saving

```{r}
#| label: stacked_histogram_save
#| eval: false
ggsave(
  filename = "../graphs/stacked_histogram.pdf",
  plot     = p_final,
  width    = 11,
  height   = 8
)
```

# Goodness-of-Follow-Up Plot

The goodness-of-follow-up plot is a standard quality-control figure in
longitudinal outcome analyses. It displays each patient as a point at their
operation date (x-axis) and follow-up duration (y-axis), with a short vertical
tick below each point. A dashed diagonal line marks the maximum potential
follow-up given the study start, study end, and follow-up closing date. The
**hvtiPlotR** package provides the `hv_followup()` constructor and a `plot()`
method to build this figure.

`plot()` returns a bare `ggplot` object with no colour, shape, axis, or label
scales applied — those are added by the caller with standard `ggplot2`
modifiers. The `type` argument selects the panel: `"followup"` (default, the
death/censoring scatter) or `"event"` (competing non-fatal event panel).

## Sample data

```{r}
#| label: gfup_data
gfup_dta <- sample_goodness_followup_data(n = 300, seed = 42)
head(gfup_dta)
```

## Death follow-up plot

```{r}
#| label: gfup_basic
gf <- hv_followup(
  data        = gfup_dta,
  origin_year = 1990,
  study_start = as.Date("1990-01-01"),
  study_end   = as.Date("2019-12-31"),
  close_date  = as.Date("2021-08-06")
)

# Bare plot — no scales or labels yet
plot(gf)
```

## Adding scales, labels, and annotations

Scale, label, and annotation layers are composed with the usual `+` operator.
`scale_color_manual()` and `scale_shape_manual()` map the binary alive/dead
state to colours and point shapes. `annotate()` places group-identifying text
directly on the panel.

```{r}
#| label: gfup_styled
library(RColorBrewer)

gfup_final <- plot(gf, alpha = 0.8) +
  # Colour alive = blue, dead = red (Set1 palette positions 2 and 1)
  scale_color_manual(
    values   = brewer.pal(3, "Set1")[c(2, 1)],
    labels   = c("Alive", "Dead"),
    na.value = "black",
    drop     = FALSE
  ) +
  scale_shape_manual(
    values = c(1, 4),
    labels = c("Alive", "Dead")
  ) +
  # Axis tick placement
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33,   3)) +
  # Clip the panel to the study window
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  # Axis and legend labels
  labs(
    x     = "Operation Date",
    y     = "Follow-up (years)",
    color = "Status",
    shape = "Status"
  ) +
  # Annotate directly on the panel
  annotate("text", x = 1993, y = 31, label = "Alive at close",
           hjust = 0, size = 3.5) +
  annotate("text", x = 1993, y = 28, label = "Deceased",
           hjust = 0, size = 3.5, color = brewer.pal(3, "Set1")[1]) +
  theme(legend.position = "none") +
  hv_theme("poster")

gfup_final
```

The diagonal dashed line represents the maximum potential follow-up. Points
sitting above the line indicate patients with longer follow-up than expected
from the study window, typically due to passive surveillance supplementing
active cross-sectional follow-up.

## Saving

```{r}
#| label: gfup_save
#| eval: false
ggsave(
  filename = "../graphs/dp_goodness-of-followup.pdf",
  plot     = gfup_final,
  height   = 6,
  width    = 6
)
```

## Non-fatal event panel

When the dataset includes a non-fatal competing event (e.g. relapse,
reoperation), pass `event_col`, `event_time_col`, and optionally
`death_for_event_col` to the constructor and then call `plot()` with
`type = "event"` to render the event panel.

```{r}
#| label: gfup_event_data
gfup_event_dta <- sample_goodness_followup_data(n = 300, seed = 42)
```

```{r}
#| label: gfup_event_panel
gf2 <- hv_followup(
  gfup_event_dta,
  event_col           = "ev_event",
  event_time_col      = "iv_event",
  death_for_event_col = "deads",
  event_levels        = c("No event", "Relapse", "Death"),
  origin_year         = 1990,
  study_start         = as.Date("1990-01-01"),
  study_end           = as.Date("2019-12-31"),
  close_date          = as.Date("2021-08-06")
)

plot(gf2, type = "event", alpha = 0.8) +
  scale_color_manual(
    values = c("No event" = "blue", "Relapse" = "green3", "Death" = "red"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("No event" = 1L, "Relapse" = 2L, "Death" = 4L),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(1990, 2020, 3)) +
  scale_y_continuous(breaks = seq(0, 33, 3)) +
  coord_cartesian(ylim = c(0, 33), xlim = c(1990, 2020)) +
  labs(
    x = "Operation Date",
    y = "Follow-up (years)",
    color = "Event", shape = "Event"
  ) +
  annotate("text", x = 1993, y = 31,
           label = "Systematic follow-up", hjust = 0, size = 3.5) +
  theme(legend.position = c(0.85, 0.15)) +
  hv_theme("poster")
```

The default follow-up panel (`plot(gf)`) and the event panel (`plot(gf2, type = "event")`)
share the same diagonal reference line and can be saved individually with `ggsave()`.

# Covariate Balance Plot

The covariate balance plot is the standard quality-control figure for
propensity score matching and IPTW analyses. Each covariate occupies a
labelled row; points show the standardized mean difference (SMD) for each
comparison group (e.g. before and after matching). A solid vertical line marks
zero balance; dotted vertical lines mark an imbalance threshold (default ±10%).

The **hvtiPlotR** package provides `hv_balance()` to prepare data and
`plot()` to render the figure, superseding `tp.lp.propen.cov_balance.R`.
`plot()` returns a bare `ggplot` object — no colour, shape, axis labels, or
theme applied — so all styling is added with the usual `+` operator.

Input data must be in **long format**: one row per covariate × group
combination with columns for the covariate name, the group label, and the
numeric SMD value. When the source dataset arrives in wide format (one column
per time-point, as the SAS export does), reshape it first:

```{r}
#| label: cov_balance_wide_reshape
# Simulate a wide-format export (e.g. from SAS or a summary table)
dta_wide <- data.frame(
  variable      = c("Age", "Female sex", "Hypertension", "Diabetes", "COPD"),
  `Before match` = c(22.1, -15.3,  18.7, -9.4,  11.2),
  `After match`  = c( 3.5,   2.1,  -1.8,  4.0,  -2.3),
  check.names = FALSE
)

dta_long <- reshape(
  dta_wide,
  direction = "long",
  varying   = c("Before match", "After match"),
  v.names   = "std_diff",
  timevar   = "group",
  times     = c("Before match", "After match"),
  idvar     = "variable"
)
head(dta_long)
```

## Sample data

```{r}
#| label: cov_balance_data
dta_cb <- sample_covariate_balance_data(n_vars = 12)
head(dta_cb)

# Build the S3 object once; reuse for all plot variants below
cb <- hv_balance(dta_cb)
```

## Bare plot

```{r}
#| label: cov_balance_bare
plot(cb, alpha = 0.8)
```

## Adding colour, shape, and axis scales

```{r}
#| label: cov_balance_scales
plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-45, 35),
    breaks = seq(-40, 30, 10)
  ) +
  labs(
    x = "Standardized difference (%)",
    y = ""
  ) +
  theme(legend.position = c(0.20, 0.95))
```

## Adding directional annotations and theme

`annotate()` places explanatory text directly on the panel to indicate which
direction of imbalance favours each group. The example below uses the exact
labels, x-scale, annotation positions, and legend placement from
`tp.lp.propen.cov_balance.R` (SAVR vs. TF-TAVR study).

```{r}
#| label: cov_balance_annotated
n_vars <- length(unique(dta_cb$variable))

cb_final <- plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(
    limits = c(-40, 30),
    breaks = seq(-40, 30, 10)
  ) +
  labs(x = "Standardized difference: SAVR - TF:TAVR (%)", y = "") +
  annotate("text", x = -30, y = 0,        label = "More likely TF-TAVR", size = 4.5) +
  annotate("text", x =  22, y = n_vars,   label = "More likely SAVR",    size = 4.5) +
  theme(legend.position = c(0.20, 0.935)) +
  hv_theme("poster")

cb_final
```

## Controlling covariate order

Pass `var_levels` to the constructor to control the bottom-to-top display order
of covariates.

```{r}
#| label: cov_balance_order
cb_ord <- hv_balance(dta_cb, var_levels = rev(unique(dta_cb$variable)))

plot(cb_ord, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  hv_theme("poster")
```

## Saving

```{r}
#| label: cov_balance_save
#| eval: false
ggsave(
  filename = here::here("graphs", "lp_cov-balance-SAVR_TF-TAVR.pdf"),
  plot     = cb_final,
  height   = 7,
  width    = 8
)
```

# Kaplan-Meier Survival Curve

`hv_survival()` estimates the Kaplan-Meier product-limit survival function
and stores all five companion plots' data (matching the SAS `%kaplan` macro
output `PLOTS`, `PLOTC`, `PLOTH`, `PLOTL`) plus tidy table data frames in the
returned S3 object. Call `plot()` to render a bare ggplot for the selected
`type` (default `"survival"`). Tidy tables are accessible via `$tables`.
Confidence intervals use the logit transform with a default confidence level
of 0.95.

## Sample data

```{r}
#| label: km_data
dta_km <- sample_survival_data(n = 500, seed = 42)
head(dta_km)

# Build the S3 object once; reuse across all plot variants below
km <- hv_survival(dta_km)
```

## Survival curve (PLOTS=1)

```{r}
#| label: km_result
# Bare plot — no scales or labels yet
plot(km)
```

## Adding scales, labels, and annotations

```{r}
#| label: km_styled
km_final <- plot(km, alpha = 0.8) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(
    breaks = seq(0, 100, 20),
    labels = function(x) paste0(x, "%")
  ) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    x     = "Years after Operation",
    y     = "Freedom from Death (%)",
    title = "Overall Survival"
  ) +
  annotate("text", x = 1, y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  hv_theme("poster")

km_final
```

## Numbers at risk and report table

```{r}
#| label: km_tables
km$tables$risk
km$tables$report
```

## Saving

```{r}
#| label: km_save
#| eval: false
ggsave("../graphs/km_survival.pdf", km_final, width = 8, height = 6)
```

## Stratified analysis

```{r}
#| label: km_strata_data
dta_km_s <- sample_survival_data(
  n             = 500,
  strata_levels = c("Type A", "Type B"),
  hazard_ratios = c(1, 1.4),
  seed          = 42
)

km_s <- hv_survival(dta_km_s, group_col = "valve_type")

plot(km_s, alpha = 0.8) +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_fill_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Survival by Valve Type") +
  theme(legend.position = c(0.15, 0.20)) +
  hv_theme("poster")
```

## Cumulative hazard (PLOTC=1)

```{r}
#| label: km_cumhaz
plot(km, type = "cumhaz") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Cumulative Hazard H(t)",
       title = "Nelson-Aalen Cumulative Hazard") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  hv_theme("poster")
```

## Log-log survival plot (Weibull/PH check)

Parallel lines across strata indicate proportional hazards.

```{r}
#| label: km_loglog
plot(km_s, type = "loglog") +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  labs(x = "log(Years after Operation)", y = "log(-log S(t))",
       title = "Log-Log Survival — Proportional-Hazards Check") +
  hv_theme("poster")
```

## Hazard rate (PLOTH=1)

The raw point estimates are noisy; add `geom_smooth()` for a publication-ready
smoothed hazard curve.

```{r}
#| label: km_hazard
plot(km, type = "hazard") +
  geom_smooth(
    aes(x = mid_time, y = hazard, color = strata),
    method = "loess", se = FALSE, span = 0.6
  ) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Instantaneous Hazard",
       title = "Hazard Rate") +
  hv_theme("poster")
```

## Integrated survivorship / restricted mean survival (PLOTL=1)

```{r}
#| label: km_life
plot(km, type = "life") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation",
       y = "Restricted Mean Survival (years)",
       title = "Integral of Survivorship") +
  hv_theme("poster")
```

# EDA Barplots and Scatterplots

The **hvtiPlotR** package provides `hv_eda()` for exploratory data analysis
of all variables in a dataset against a reference time axis. It replicates the
`Function_DataPlotting()` workflow from `tp.dp.EDA_barplots_scatterplots.R`
and `tp.dp.EDA_barplots_scatterplots_varnames.R`, replacing base-R graphics
with composable `ggplot2` objects.

Three helper functions support the workflow:

- `eda_classify_var()` — detects whether a column is continuous (`"Cont"`),
  numeric-categorical (`"Cat_Num"`), or character-categorical (`"Cat_Char"`),
  matching the `UniqueLimit` logic from the template.
- `eda_select_vars()` — subsets and reorders columns by name or
  space-separated string, replacing the `Order_Variables()` /
  `Mod_Data <- dta[, Order_Var]` pattern from the varnames template.
- `sample_eda_data()` — generates a reproducible mixed-type dataset for
  demonstration.

`plot()` on an `hv_eda` object always returns a bare `ggplot`. Colour scales,
axis labels, annotations, and `hv_theme()` are added by the caller.

## Sample data

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)

# Inspect auto-detected types for each column
sapply(dta_eda, eda_classify_var)
```

## Binary categorical: count barplot

Numeric 0/1 columns are classified as `"Cat_Num"`. `NA` values appear as an
explicit `"(Missing)"` fill level so they can be coloured with
`scale_fill_manual()`. The `y_label` argument sets the plot title and
fill-legend name in place of the raw column name.

```{r}
#| label: eda_binary_count
plot(hv_eda(dta_eda, x_col = "year", y_col = "male",
              y_label = "Sex")) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick", "(Missing)" = "grey80"),
    labels = c("0" = "Female", "1" = "Male", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  hv_theme("poster")
```

## Binary categorical: percentage barplot

Setting `show_percent = TRUE` switches `geom_bar()` to `position = "fill"`.

```{r}
#| label: eda_binary_percent
plot(hv_eda(dta_eda, x_col = "year", y_col = "cabg",
              y_label = "Concomitant CABG", show_percent = TRUE)) +
  scale_fill_manual(
    values = c("0" = "grey70", "1" = "steelblue", "(Missing)" = "grey90"),
    labels = c("0" = "No CABG", "1" = "CABG", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Surgery Year", y = "Proportion") +
  hv_theme("poster")
```

## Ordinal and multi-level categorical

```{r}
#| label: eda_ordinal
plot(hv_eda(dta_eda, x_col = "year", y_col = "nyha",
              y_label = "Preoperative NYHA Class")) +
  scale_fill_brewer(
    palette = "RdYlGn", direction = -1,
    labels  = c("1" = "I", "2" = "II", "3" = "III", "4" = "IV",
                "(Missing)" = "Missing"),
    name    = "NYHA"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  hv_theme("poster")
```

## Character categorical

```{r}
#| label: eda_char_cat
plot(hv_eda(dta_eda, x_col = "year", y_col = "valve_morph",
              y_label = "Valve Morphology")) +
  scale_fill_manual(
    values = c(Bicuspid   = "steelblue",
               Tricuspid  = "firebrick",
               Unicuspid  = "goldenrod3",
               "(Missing)" = "grey80"),
    name = "Morphology"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  hv_theme("poster")
```

## Continuous: scatter + LOESS

Continuous columns produce a scatter plot with a LOESS smoother overlay.
Where `y_col` is `NA`, a rug mark is drawn on the x-axis.

```{r}
#| label: eda_continuous
plot(hv_eda(dta_eda, x_col = "op_years", y_col = "ef",
              y_label = "Ejection Fraction (%)")) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       caption = "Tick marks on x-axis: observations with missing EF") +
  hv_theme("poster")
```

## Variable selection and labels (varnames template pattern)

`eda_select_vars()` subsets columns by name, matching the `Var_CatList` /
`Var_ContList` + `Order_Variables()` workflow. A named vector whose names are
column names and whose values are human-readable labels drives a single
`lapply()` loop.

```{r}
#| label: eda_varnames_binary
bin_vars <- c(male = "Sex (Male)", cabg = "Concomitant CABG")
sub_bin  <- eda_select_vars(dta_eda, c("year", names(bin_vars)))

p_bin <- lapply(names(bin_vars), function(cn) {
  plot(hv_eda(sub_bin, x_col = "year", y_col = cn,
                y_label = bin_vars[[cn]])) +
    scale_fill_brewer(palette = "Set1", direction = -1, name = NULL) +
    scale_x_discrete(breaks = seq(2005, 2020, 5)) +
    labs(x = "Surgery Year", y = "Count") +
    hv_theme("poster")
})
p_bin[[1]]
p_bin[[2]]
```

```{r}
#| label: eda_varnames_ordinal
cat_vars <- c(nyha        = "NYHA Class",
              valve_morph = "Valve Morphology")
sub_cat <- eda_select_vars(dta_eda, c("year", names(cat_vars)))

p_cat <- lapply(names(cat_vars), function(cn) {
  plot(hv_eda(sub_cat, x_col = "year", y_col = cn,
                y_label = cat_vars[[cn]])) +
    scale_fill_brewer(palette = "Set2", name = NULL) +
    scale_x_discrete(breaks = seq(2005, 2020, 5)) +
    labs(x = "Surgery Year", y = "Count") +
    hv_theme("poster")
})
p_cat[[1]]
p_cat[[2]]
```

```{r}
#| label: eda_varnames_continuous
cont_vars <- c(ef        = "Ejection Fraction (%)",
               lv_mass   = "LV Mass Index (g/m\u00b2)",
               peak_grad = "Peak Gradient (mmHg)")
sub_cont <- eda_select_vars(dta_eda, c("op_years", names(cont_vars)))

p_cont <- lapply(names(cont_vars), function(cn) {
  plot(hv_eda(sub_cont, x_col = "op_years", y_col = cn,
                y_label = cont_vars[[cn]])) +
    scale_colour_manual(values = c("steelblue"), guide = "none") +
    scale_x_continuous(breaks = seq(0, 15, 5)) +
    labs(x = "Years from First Surgery Year") +
    hv_theme("poster")
})
p_cont[[1]]
p_cont[[2]]
p_cont[[3]]
```

## Saving EDA plots

`gridExtra::marrangeGrob()` arranges multiple plots into a grid and
`ggsave()` writes each page to a separate PDF file.

```{r}
#| label: eda_save
#| eval: false
all_plots <- c(p_bin, p_cat, p_cont)
per_page  <- 9L  # 3 x 3 grid

for (pg in seq(1, length(all_plots), by = per_page)) {
  idx  <- seq(pg, min(pg + per_page - 1L, length(all_plots)))
  grob <- gridExtra::marrangeGrob(all_plots[idx], nrow = 3, ncol = 3)
  ggsave(
    filename = sprintf("../graphs/eda_page%02d.pdf", ceiling(pg / per_page)),
    plot     = grob,
    width    = 14,
    height   = 14
  )
}
```

# Alluvial (Sankey) Plot

`hv_alluvial()` prepares an alluvial (Sankey-style) diagram using
`ggalluvial`. Each row of the input is a unique combination of axis values
with an associated patient count; `plot()` draws flows proportional to that
count. Ports `tp.dp.female_bicus_preAR_sankey.R`.

## Sample data

```{r}
#| label: alluvial_data
dta_al  <- sample_alluvial_data(n = 300, seed = 42)
axes    <- c("pre_ar", "procedure", "post_ar")
head(dta_al)
```

## Bare plot

```{r}
#| label: alluvial_bare
al <- hv_alluvial(dta_al, axes = axes, y_col = "freq")
plot(al)
```

## Fill flows by pre-operative grade

```{r}
#| label: alluvial_filled
al_filled <- hv_alluvial(dta_al, axes = axes, y_col = "freq",
                            fill_col = "pre_ar")

plot(al_filled) +
  scale_fill_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    name = "Pre-op AR"
  ) +
  scale_colour_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c("Pre-op AR", "Procedure", "Post-op AR"),
    expand = c(0.05, 0.05)
  ) +
  labs(y = "Patients (n)",
       title = "AV Regurgitation: Pre- to Post-operative") +
  hv_theme("poster")
```

## Two-axis before / after comparison

```{r}
#| label: alluvial_two_axis
al2 <- hv_alluvial(
  dta_al,
  axes        = c("pre_ar", "post_ar"),
  y_col       = "freq",
  fill_col    = "pre_ar",
  axis_labels = c("Pre-operative", "Post-operative")
)

plot(al2) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1,
                    name = "AR Grade") +
  scale_colour_brewer(palette = "RdYlGn", direction = -1,
                      guide = "none") +
  annotate("text", x = 1.5, y = 250,
           label = "Improvement after surgery",
           size = 3.5, fontface = "italic") +
  labs(y = "Patients (n)",
       title = "AV Regurgitation Before and After Surgery") +
  hv_theme("poster")
```

## Saving

```{r}
#| label: alluvial_save
#| eval: false
p_al <- plot(al_filled) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  scale_colour_brewer(palette = "RdYlGn", direction = -1, guide = "none") +
  labs(y = "Patients (n)") +
  hv_theme("poster")

ggsave("../graphs/alluvial.pdf", p_al, width = 8, height = 6)
```

# Cluster Stability Sankey Plot

`hv_sankey()` prepares a Sankey diagram showing how patients flow between
labelled clusters as the number of clusters K increases. It ports the PAM
cluster stability figure from the HVTI clustering analysis pipeline. Each
column represents one value of K (default K = 2 to 9); each band shows the
fraction of patients whose assignment changes between consecutive K values. Node
labels show the cluster letter and count.

**Requires `ggsankey`** (not on CRAN):
```r
remotes::install_github("davidsjoberg/ggsankey")
```

## Sample data

```{r}
#| label: sankey_data
dta_san <- sample_cluster_sankey_data(n = 300, seed = 42)
head(dta_san)
table(dta_san$C9)
```

## Default plot (K = 2 to 9)

```{r}
#| label: sankey_default
#| eval: !expr requireNamespace("ggsankey", quietly = TRUE)
#| fig-width: 8
#| fig-height: 5
sk <- hv_sankey(dta_san)
plot(sk) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  hv_theme("poster")
```

## Custom colour palette

Replace the default Set1 colours with a fully custom named vector. Names must
match the node labels in the data.

```{r}
#| label: sankey_custom_cols
#| eval: !expr requireNamespace("ggsankey", quietly = TRUE)
#| fig-width: 8
#| fig-height: 5
my_cols <- c(
  A = "#1f77b4", B = "#ff7f0e", C = "#2ca02c", D = "#d62728",
  E = "#9467bd", F = "#8c564b", G = "#e377c2", H = "#7f7f7f",
  I = "#bcbd22"
)
sk_custom <- hv_sankey(dta_san, node_colours = my_cols)
plot(sk_custom) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  hv_theme("poster")
```

## Subset of K values

Pass a shorter `cluster_cols` vector to the constructor to show only a range of K.

```{r}
#| label: sankey_subset
#| eval: !expr requireNamespace("ggsankey", quietly = TRUE)
#| fig-width: 6
#| fig-height: 5
sk_sub <- hv_sankey(dta_san, cluster_cols = paste0("C", 2:6))
plot(sk_sub) +
  labs(title = "Cluster Stability: K = 2 to 6") +
  hv_theme("poster")
```

## Saving

```{r}
#| label: sankey_save
#| eval: false
p_san <- plot(sk) +
  labs(title = "PAM Cluster Stability") +
  hv_theme("poster")

ggsave("../graphs/sankey_clusters.pdf", p_san, width = 8, height = 5)
```

# Hazard Plot

`hazard_plot()` plots pre-computed parametric survival, hazard, or
cumulative-hazard curves from a fitted Weibull (or other parametric) model,
with optional Kaplan-Meier empirical overlay and population life-table
reference. It ports the entire `tp.hp.dead.*` SAS template family.

The input data comes from two sources that map directly to the SAS output:

| Column set | SAS dataset | R function |
|---|---|---|
| Parametric prediction grid | `predict` (`SSURVIV`, `SCLLSURV`, `SCLUSURV`) | `sample_hazard_data()` |
| KM empirical overlay | `plout` (`CUM_SURV`, `CL_LOWER`, `CL_UPPER`) | `sample_hazard_empirical()` |
| Population life table | `smatched` | `sample_life_table()` |

## Sample data

```{r}
#| label: hazard_data
dat_hp  <- sample_hazard_data(n = 500, time_max = 10)
emp_hp  <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)
head(dat_hp)
```

## Survival curve with KM overlay

The most common `tp.hp.dead.sas` output: smooth parametric survival with
confidence band plus discrete KM empirical circles and error bars.

```{r}
#| label: hazard_survival
hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)") +
  hv_theme("poster")
```

## Hazard rate curve

Switching `estimate_col` to `"hazard"` (and the matching CI columns) shows the
instantaneous hazard rate (%/year) instead of survival.

```{r}
#| label: hazard_rate
hazard_plot(
  dat_hp,
  estimate_col = "hazard",
  lower_col    = "haz_lower",
  upper_col    = "haz_upper"
) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 30),
                     labels = function(x) paste0(x, "%/yr")) +
  labs(x = "Years", y = "Hazard (%/year)") +
  hv_theme("poster")
```

## Cumulative hazard

Used for readmission and repeated-event analyses (`tp.hp.event.weighted.sas`,
`tp.hp.repeated_events.sas`). The `"cumhaz"` column equals `-log(S) * 100`.

```{r}
#| label: hazard_cumhaz
hazard_plot(
  dat_hp,
  estimate_col  = "cumhaz",
  lower_col     = "cumhaz_lower",
  upper_col     = "cumhaz_upper"
) +
  scale_colour_manual(values = c("darkorange"), guide = "none") +
  scale_fill_manual(values = c("darkorange"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Cumulative Hazard (%)") +
  hv_theme("poster")
```

## Stratified by group

Pass `group_col` to compare two or more groups
(`tp.hp.dead.tkdn.stratified.sas`).

```{r}
#| label: hazard_stratified
dat_strat <- sample_hazard_data(
  n = 400, time_max = 10,
  groups = c("No Takedown" = 1.0, "Takedown" = 0.65)
)
emp_strat <- sample_hazard_empirical(
  n = 400, time_max = 10, n_bins = 6,
  groups = c("No Takedown" = 1.0, "Takedown" = 0.65)
)

hazard_plot(
  dat_strat,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  group_col     = "group",
  empirical     = emp_strat,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(
    values = c("No Takedown" = "steelblue", "Takedown" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("No Takedown" = "steelblue", "Takedown" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Esophagostomy", y = "Survival (%)") +
  hv_theme("poster")
```

## Life-table overlay

`tp.hp.dead.age_with_population_life_table.sas` and
`tp.hp.dead.uslife.stratifed.sas` overlay US population life-table survival
(dashed lines) on age-stratified study curves. Pass the life-table data frame
to `reference` and set `ref_group_col` to vary the linetype by age group.

```{r}
#| label: hazard_lifetable
dat_age <- sample_hazard_data(
  n = 600, time_max = 12,
  groups = c("<65" = 0.5, "65\u201380" = 1.0, "\u226580" = 1.8)
)
emp_age <- sample_hazard_empirical(
  n = 600, time_max = 12, n_bins = 6,
  groups = c("<65" = 0.5, "65\u201380" = 1.0, "\u226580" = 1.8)
)
lt <- sample_life_table(
  age_groups = c("<65", "65\u201380", "\u226580"),
  age_mids   = c(55, 72, 85),
  time_max   = 12
)

hazard_plot(
  dat_age,
  estimate_col     = "survival",
  lower_col        = "surv_lower",
  upper_col        = "surv_upper",
  group_col        = "group",
  empirical        = emp_age,
  emp_lower_col    = "lower",
  emp_upper_col    = "upper",
  reference        = lt,
  ref_estimate_col = "survival",
  ref_group_col    = "group"
) +
  scale_colour_manual(
    values = c("<65" = "steelblue", "65\u201380" = "forestgreen",
               "\u226580" = "firebrick"),
    name   = "Age group"
  ) +
  scale_fill_manual(
    values = c("<65" = "steelblue", "65\u201380" = "forestgreen",
               "\u226580" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)",
       caption = "Dashed lines: US population life table") +
  hv_theme("poster")
```

# Survival Difference (Life-Gained) Plot

`survival_difference_plot()` plots the difference `S_2(t) - S_1(t)` between
two groups over time, with an optional confidence band. This ports
`tp.hp.dead.life-gained.sas` (HAZDIFL macro output).

## Sample data

```{r}
#| label: surv_diff_data
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("Control" = 1.0, "Treatment" = 0.70)
)
head(diff_dat)
```

## Survival difference curve

```{r}
#| label: surv_diff_plot
survival_difference_plot(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                      colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-5, 40),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival Difference (%)") +
  hv_theme("poster")
```

## Multiple comparisons

Combine several two-group differences into one long data frame and use
`group_col` to overlay them.

```{r}
#| label: surv_diff_multi
d1 <- sample_survival_difference_data(
  groups = c("Medical Mgmt" = 1.0, "TF-TAVR" = 0.70), seed = 1L
)
d1$comparison <- "TF-TAVR vs Medical Mgmt"

d2 <- sample_survival_difference_data(
  groups = c("TA-TAVR" = 0.90, "TF-TAVR" = 0.70), seed = 2L
)
d2$comparison <- "TF-TAVR vs TA-TAVR"

d3 <- sample_survival_difference_data(
  groups = c("AVR" = 0.80, "TF-TAVR" = 0.70), seed = 3L
)
d3$comparison <- "TF-TAVR vs AVR"

survival_difference_plot(rbind(d1, d2, d3),
                         group_col = "comparison") +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                      colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Survival Difference (%)") +
  hv_theme("poster")
```

# Number Needed to Treat (NNT) Plot

`nnt_plot()` plots the number needed to treat and absolute risk reduction over
time, porting the NNT component of `tp.hp.numtreat.survdiff.matched.sas`.

## Sample data

```{r}
#| label: nnt_data
nnt_dat <- sample_nnt_data(
  n      = 500,
  time_max = 20,
  groups = c("SVG" = 1.0, "ITA" = 0.75)
)
head(nnt_dat)
```

## NNT curve

NNT decreases over time as the treatment benefit accumulates.

```{r}
#| label: nnt_curve
nnt_plot(
  nnt_dat,
  lower_col = "nnt_lower",
  upper_col = "nnt_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) +
  labs(x = "Years", y = "Number Needed to Treat") +
  hv_theme("poster")
```

## Absolute risk reduction

The same data, plotted as ARR (%) instead of NNT.

```{r}
#| label: nnt_arr
nnt_plot(
  nnt_dat,
  estimate_col = "arr",
  lower_col    = "arr_lower",
  upper_col    = "arr_upper"
) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_y_continuous(limits = c(0, 50),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Absolute Risk Reduction (%)") +
  hv_theme("poster")
```

## Saving

```{r}
#| label: hazard_save
#| eval: false
p_hp <- hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  labs(x = "Years", y = "Survival (%)") +
  hv_theme("poster")

ggsave("../graphs/hazard_survival.pdf", p_hp, width = 11.5, height = 8)
```

# Temporal Trend Plot

`hv_trends()` ports the pattern from five SAS/R templates:

| Template | Study | Key pattern |
|---|---|---|
| `tp.rp.trends.sas` | Tricuspid valve replacement, 1968–2000 | Single continuous outcome vs operation year |
| `tp.lp.trends.sas` | Post-infarct VSD, 1969–2000 | Binary % outcomes, x 1970–2000 by 10, y 0–100 |
| `tp.lp.trends.age.sas` | Mitral valve surgery, 1990–1999 | Binary % vs **age** (not year), x 25–85 by 10 |
| `tp.lp.trends.polytomous.sas` | Tricuspid valve repair, 1990–1999 | Polytomous (≥3) groups, x 1990–1999 by 1 |
| `tp.dp.trends.R` | Mitral degeneration, 1985–2015 | NYHA %, LV mass, %CHF, case volume, LOS |

The constructor accepts patient-level data and computes annual summaries (mean
or median) internally. `plot()` returns a bare ggplot. Compose with
`scale_colour_*`, `scale_shape_*`, `scale_x_continuous()`,
`scale_y_continuous()`, `coord_cartesian()`, `labs()`, `annotate()`, and
`hv_theme()`.

## Sample data

```{r}
#| label: trends_data
# year_range matches the 1968-2000 Tricuspid Valve Replacement study in the
# SAS template (tp.rp.trends.sas)
dta_tr <- sample_trends_data(
  n          = 600,
  year_range = c(1968L, 2000L),
  groups     = c("Group I", "Group II", "Group III", "Group IV")
)
head(dta_tr)
```

## Single variable: cases/year

The SAS template plots annual case volume against operation year with
`axisx order=(1968 to 2000 by 4)` and `axisy order=(0 to 10 by 2)`.

```{r}
#| label: trends_n_year_build
one_grp <- dta_tr[dta_tr$group == "Group I", ]
tr1 <- hv_trends(one_grp, group_col = NULL)
```

### Bare plot

```{r}
#| label: trends_n_year_bare
p_tr1 <- plot(tr1)
p_tr1
```

### Adding scales, labels, and theme

```{r}
#| label: trends_n_year_decorated
p_tr1 +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 10),      breaks = seq(0, 10, 2)) +
  labs(x = "Year", y = "Cases/year") +
  hv_theme("poster")
```

## Single variable: age

The age figure uses the same x-axis and `axisy order=(30 to 70 by 10)`.

```{r}
#| label: trends_age
plot(tr1) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(30, 70),     breaks = seq(30, 70, 10)) +
  labs(x = "Year", y = "Age (years)") +
  hv_theme("poster")
```

## Multiple groups with `scale_colour_brewer`

```{r}
#| label: trends_multi_brewer
tr <- hv_trends(dta_tr)
plot(tr) +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "Group"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Outcome (%)") +
  hv_theme("poster")
```

## Median summary + manual colours (NYHA style)

```{r}
#| label: trends_median_manual
tr_med <- hv_trends(dta_tr, summary_fn = "median")
plot(tr_med) +
  scale_colour_manual(
    values = c(
      "Group I"   = "steelblue",
      "Group II"  = "firebrick",
      "Group III" = "forestgreen",
      "Group IV"  = "goldenrod3"
    ),
    name = "NYHA Class"
  ) +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "NYHA Class"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 80),      breaks = seq(0, 80, 20)) +
  labs(x = "Year", y = "%", title = "Preoperative NYHA Class Over Time") +
  annotate("text", x = 1980, y = 75,
           label = "Trend: Preoperative NYHA", size = 4) +
  hv_theme("poster")
```

## With confidence ribbon

```{r}
#| label: trends_ribbon
plot(tr1, se = TRUE, alpha = 0.2) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Cases/year") +
  hv_theme("poster")
```

## Binary % outcomes (tp.lp.trends.sas)

The VSD template plots multiple binary outcomes (cardiogenic shock %, pre-op
IABP %, inotropes %) on the same axes: `axisx order=(1970 to 2000 by 10)`,
`axisy order=(0 to 100 by 10)`. Each outcome becomes a group.

```{r}
#| label: trends_binary_pct
dta_lp <- sample_trends_data(
  n          = 800,
  year_range = c(1970L, 2000L),
  groups     = c("Shock %", "Pre-op IABP %", "Inotropes %")
)

plot(hv_trends(dta_lp)) +
  scale_colour_manual(
    values = c("Shock %"       = "steelblue",
               "Pre-op IABP %" = "firebrick",
               "Inotropes %"   = "forestgreen"),
    name = NULL
  ) +
  scale_shape_manual(
    values = c("Shock %" = 16L, "Pre-op IABP %" = 15L, "Inotropes %" = 17L),
    name = NULL
  ) +
  scale_x_continuous(limits = c(1970, 2000), breaks = seq(1970, 2000, 10)) +
  scale_y_continuous(limits = c(0, 100),     breaks = seq(0, 100, 10)) +
  coord_cartesian(xlim = c(1970, 2000), ylim = c(0, 100)) +
  labs(x = "Year", y = "Percent (%)") +
  hv_theme("poster")
```

## Age as x-axis (tp.lp.trends.age.sas)

The mitral valve template uses patient age (not year) on the x-axis:
`axisx order=(25 to 85 by 10)`, `axisy order=(0 to 100 by 20)`. Pass real
data with an `age` column and set `x_col = "age"`. The example below uses
the sample data's `year` column relabelled for illustration.

```{r}
#| label: trends_age_x
# With real data: x_col = "age", x_col_data has values ~25 to 85
# Illustration using sample_trends_data() with year_range spanning age range:
dta_age <- sample_trends_data(
  n          = 600,
  year_range = c(25L, 85L),
  groups     = c("Repair %", "Bioprosthesis %"),
  seed       = 7L
)

plot(hv_trends(dta_age)) +
  scale_colour_manual(
    values = c("Repair %" = "steelblue", "Bioprosthesis %" = "firebrick"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Repair %" = 16L, "Bioprosthesis %" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(25, 85), breaks = seq(25, 85, 10)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  coord_cartesian(xlim = c(25, 85), ylim = c(0, 100)) +
  labs(x = "Age (years)", y = "Percent (%)") +
  hv_theme("poster")
```

## Polytomous groups — repair types (tp.lp.trends.polytomous.sas)

Four repair categories over a short study period (1990–1999): the SAS template
uses `axisx order=(1990 to 1999 by 1)` for fine year breaks.

```{r}
#| label: trends_polytomous
dta_poly <- sample_trends_data(
  n          = 800,
  year_range = c(1990L, 1999L),
  groups     = c("CE", "Cosgrove", "Periguard", "DeVega"),
  seed       = 5L
)

plot(hv_trends(dta_poly)) +
  scale_colour_manual(
    values = c(CE        = "steelblue",
               Cosgrove  = "firebrick",
               Periguard = "forestgreen",
               DeVega    = "goldenrod3"),
    name = "Repair type"
  ) +
  scale_shape_manual(
    values = c(CE = 15L, Cosgrove = 19L, Periguard = 17L, DeVega = 18L),
    name = "Repair type"
  ) +
  scale_x_continuous(limits = c(1990, 1999), breaks = seq(1990, 1999, 1)) +
  scale_y_continuous(limits = c(0, 100),     breaks = seq(0, 100, 10)) +
  coord_cartesian(xlim = c(1990, 1999), ylim = c(0, 100)) +
  labs(x = "Year", y = "Percent (%)") +
  hv_theme("poster")
```

## LV mass index (tp.dp.trends.R — plot2)

Continuous outcome with a larger y range: `scale_y_continuous(breaks=seq(0, 200, 50))`,
`coord_cartesian(xlim = c(1995, 2016), ylim = c(0, 200))`.

```{r}
#| label: trends_lv_mass
dta_lv <- sample_trends_data(
  n          = 800,
  year_range = c(1995L, 2015L),
  groups     = NULL,
  seed       = 3L
)

plot(hv_trends(dta_lv, group_col = NULL)) +
  scale_x_continuous(limits = c(1995, 2015), breaks = seq(1995, 2015, 5)) +
  scale_y_continuous(limits = c(0, 200),     breaks = seq(0, 200, 50)) +
  coord_cartesian(xlim = c(1995, 2015), ylim = c(0, 200)) +
  labs(x = "Years", y = "LV Mass Index") +
  hv_theme("poster")
```

## Case volume / total surgeries per year (tp.dp.trends.R — plot4)

```{r}
#| label: trends_case_volume
dta_vol <- sample_trends_data(
  n          = 1000,
  year_range = c(1985L, 2015L),
  groups     = NULL,
  seed       = 9L
)

plot(hv_trends(dta_vol, group_col = NULL)) +
  scale_x_continuous(limits = c(1985, 2015), breaks = seq(1985, 2015, 5)) +
  scale_y_continuous(limits = c(0, 400),     breaks = seq(0, 400, 50)) +
  coord_cartesian(xlim = c(1985, 2015), ylim = c(0, 400)) +
  labs(x = "Years", y = "Surgeries (#)") +
  hv_theme("poster")
```

## Annotated trend — hospital LOS (tp.dp.trends.R — plot5)

Template: `coord_cartesian(ylim = c(0, 20))`,
`annotate("text", 1995, 18, label="Trend: Hospital Length of Stay", size=4.5)`.

```{r}
#| label: trends_los
dta_los <- sample_trends_data(
  n          = 800,
  year_range = c(1985L, 2015L),
  groups     = NULL,
  seed       = 11L
)

plot(hv_trends(dta_los, group_col = NULL)) +
  scale_x_continuous(limits = c(1985, 2015), breaks = seq(1985, 2015, 5)) +
  scale_y_continuous(limits = c(0, 20),      breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(1985, 2015), ylim = c(0, 20)) +
  annotate("text", x = 1995, y = 18,
           label = "Trend: Hospital Length of Stay", size = 4.5) +
  labs(x = "Years", y = "Hospital LOS (Days)") +
  hv_theme("poster")
```

## Saving

```{r}
#| label: trends_save
#| eval: false
p_tr <- plot(tr) +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "Group"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Outcome (%)") +
  hv_theme("poster")

ggsave(here::here("graphs", "rp.trends.pdf"), p_tr, width = 11.5, height = 8)
```

# Spaghetti / Profile Plot

`hv_spaghetti()` ports the pattern from `tp.dp.spaghetti.echo.R`: one
trajectory line per subject over time, with optional stratification by a
grouping variable. `plot()` accepts `add_smooth = TRUE` for an optional LOESS
overlay. The original template covers nine figures — unstratified and
sex-stratified variants of three echo outcomes (AV mean gradient, AV area, DVI)
plus an ordinal MV regurgitation grade plot.

`plot()` returns a bare ggplot. Compose with `scale_colour_manual()`,
`scale_x_continuous()`, `scale_y_continuous()`, `coord_cartesian()`, `labs()`,
and `hv_theme()`.

## Sample data

```{r}
#| label: spaghetti_data
# groups mirrors the Female/Male sex stratification in the template (MALE column)
dta_sp <- sample_spaghetti_data(
  n_patients = 150,
  max_obs    = 6,
  groups     = c(Female = 0.45, Male = 0.55),
  seed       = 42L
)
head(dta_sp)

# Build S3 objects for reuse below
sp     <- hv_spaghetti(dta_sp)
sp_col <- hv_spaghetti(dta_sp, colour_col = "group")
```

## Bare plot

```{r}
#| label: spaghetti_bare
p_sp <- plot(sp)
p_sp
```

## Unstratified — AV mean gradient, full range (plot_1)

Template: `scale_y_continuous(breaks=seq(0, 80, 20))`,
`coord_cartesian(xlim = c(0, 5), ylim = c(0, 80))`.

```{r}
#| label: spaghetti_unstrat_full
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hv_theme("poster")
```

## Unstratified — zoomed y-axis (plot_3)

Template: `scale_y_continuous(breaks=seq(0, 30, 10))`,
`coord_cartesian(ylim = c(0, 30))`.

```{r}
#| label: spaghetti_unstrat_zoom
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 30, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 30)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hv_theme("poster")
```

## Stratified by sex — AV mean gradient (plot_2 / plot_4)

Template: `scale_color_manual(breaks = c("0", "1"), values=c("red", "blue"))`.
The quick-start in the template header uses the modernised `"firebrick"` /
`"steelblue"` equivalents.

```{r}
#| label: spaghetti_sex_strat
p_sp <- plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hv_theme("poster")

p_sp
```

## AV area y-scale (plot_5 / plot_6)

Template: `scale_y_continuous(breaks=seq(0, 5, 1))`,
`coord_cartesian(ylim = c(0, 5))`, `ylab('AV Area (EOA) (cm^2)')`.

```{r}
#| label: spaghetti_avarea
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 5, 1)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(x = "Years", y = "AV Area (EOA) (cm\u00b2)") +
  hv_theme("poster")
```

## DVI y-scale (plot_7 / plot_8)

Template: `scale_y_continuous(breaks=seq(0, 1.25, 0.25))`,
`coord_cartesian(ylim = c(0, 1.25))`, `ylab('DVI')`.

```{r}
#| label: spaghetti_dvi
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 1.25, 0.25)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 1.25)) +
  labs(x = "Years", y = "DVI") +
  hv_theme("poster")
```

## Ordinal y-axis — MV regurgitation grade (plot_9)

Template: `scale_y_continuous(labels=c("None", "Mild", "Moderate", "Severe"))`,
`coord_cartesian(xlim = c(0, 6), ylim = c(0, 3))`, two colour groups for
early (blue) and late (red2) cohorts.

```{r}
#| label: spaghetti_ordinal
dta_ord        <- dta_sp
dta_ord$value  <- round(pmin(3, pmax(0, dta_sp$value / 12)))
levels(dta_ord$group) <- c("Early", "Late")

sp_ord <- hv_spaghetti(dta_ord, colour_col = "group")

plot(sp_ord, y_labels = c(None = 0, Mild = 1, Moderate = 2, Severe = 3)) +
  scale_colour_manual(
    values = c(Early = "steelblue", Late = "red2"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 6, 1)) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 3)) +
  labs(x = "Years after Procedure", y = "MV Regurgitation Grade") +
  hv_theme("poster")
```

## With LOESS smooth overlay

```{r}
#| label: spaghetti_smooth
plot(sp_col, add_smooth = TRUE) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hv_theme("poster")
```

## Saving

Template uses `wid = 11, hei = 8.5` in `pdf()`.

```{r}
#| label: spaghetti_save
#| eval: false
ggsave(here::here("graphs", "mp.amngrd_profile.pdf"),
       p_sp, width = 11, height = 8.5)
```

# Nonparametric Temporal Trend Curve

`hv_nonparametric()` prepares pre-computed average curves from a two-phase
nonparametric temporal trend model — the R equivalent of the SAS
`tp.np.*.avrg_curv.*` and `tp.np.*.u.trend.*` template family.

The constructor accepts the SAS `mean_curv` / `boots_ci` datasets exported to
CSV and read in with `read.csv()`. `plot()` returns a bare ggplot for
composition with `scale_colour_*`, `labs()`, and `hv_theme()`.

## Sample data

```{r}
#| label: np_curve_data
curve_dat <- sample_nonparametric_curve_data(
  n            = 500,
  time_max     = 12,
  outcome_type = "probability",
  ci_level     = 0.68       # 68 % CI — one SE, matching SAS cll_p68/clu_p68
)
pts_dat <- sample_nonparametric_curve_points(n = 500, time_max = 12)
head(curve_dat)
```

## Single average curve with 68 % CI ribbon

```{r}
#| label: np_curve_single_build
np <- hv_nonparametric(
  curve_data  = curve_dat,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = pts_dat
)
```

### Bare plot

```{r}
#| label: np_curve_single_bare
p_np_bare <- plot(np)
p_np_bare
```

### Adding scales, labels, and theme

```{r}
#| label: np_curve_single_decorated
p_np <- p_np_bare +
  ggplot2::scale_colour_manual(values = c("steelblue"), guide = "none") +
  ggplot2::scale_fill_manual(values   = c("steelblue"), guide = "none") +
  ggplot2::scale_x_continuous(
    limits = c(0, 12),
    breaks = seq(0, 12, 2),
    labels = function(x) paste(x, "yr")
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  ggplot2::labs(x = "Follow-up (years)", y = "Prevalence (%)") +
  hv_theme("poster")

p_np
```

The shaded ribbon shows the 68 % bootstrap CI (one standard error), matching
`cll_p68`/`clu_p68` in the SAS template. Switch to `ci_level = 0.95` in
`sample_nonparametric_curve_data()` for 95 % CI bands.

## Two-group comparison (analogous to tp.np.avpkgrad_ozak_ind_mtwt.sas)

```{r}
#| label: np_curve_groups
curve_grp <- sample_nonparametric_curve_data(
  n      = 400,
  time_max = 7,
  groups   = c("Ozaki" = 0.7, "CE-Pericardial" = 1.3),
  outcome_type = "continuous",
  ci_level = 0.68
)
pts_grp <- sample_nonparametric_curve_points(
  n = 400, time_max = 7,
  groups = c("Ozaki" = 0.7, "CE-Pericardial" = 1.3),
  outcome_type = "continuous"
)

np_grp <- hv_nonparametric(
  curve_data  = curve_grp,
  group_col   = "group",
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = pts_grp
)

plot(np_grp) +
  ggplot2::scale_colour_manual(
    values = c("Ozaki" = "steelblue", "CE-Pericardial" = "firebrick"),
    name   = NULL
  ) +
  ggplot2::scale_fill_manual(
    values = c("Ozaki" = "steelblue", "CE-Pericardial" = "firebrick"),
    guide  = "none"
  ) +
  ggplot2::scale_x_continuous(limits = c(0, 7), breaks = 0:7) +
  ggplot2::labs(
    x = "Follow-up (years)",
    y = "AV Peak Gradient (mmHg)"
  ) +
  hv_theme("poster") +
  ggplot2::theme(legend.position = c(0.15, 0.85))
```

## Saving

```{r}
#| label: np_curve_save
#| eval: false
# Manuscript PDF
ggsave(here::here("graphs", "np_afib_prevalence.pdf"),
       p_np, width = 11, height = 8.5)

# Editable PowerPoint
save_ppt(p_np,
         template   = system.file("ClevelandClinic.pptx", package = "hvtiPlotR"),
         powerpoint = here::here("graphs", "np_afib_prevalence.pptx"))
```

# Nonparametric Ordinal Outcome Curve

`hv_ordinal()` prepares pre-computed grade-specific probability curves from a
cumulative proportional-odds model — the R equivalent of `tp.np.*.ordinal.*`
SAS templates (e.g. TR grade prevalence, AR severity).

The SAS `predict` dataset stores one column per grade (`p0`, `p1`, `p2`, `p3`).
Before calling this function, reshape it from wide to long format:

```r
library(tidyr)
long <- pivot_longer(
  predict_wide,
  cols      = c(p0, p1, p2, p3),
  names_to  = "grade",
  values_to = "estimate"
)
```

## Sample data

```{r}
#| label: np_ordinal_data
ord_dat <- sample_nonparametric_ordinal_data(
  n            = 800,
  time_max     = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)
ord_pts <- sample_nonparametric_ordinal_points(
  n            = 800,
  time_max     = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)
head(ord_dat)
```

## Grade probability curves with data summary points

```{r}
#| label: np_ordinal_basic_build
np_ord <- hv_ordinal(curve_data = ord_dat, data_points = ord_pts)
```

### Bare plot

```{r}
#| label: np_ordinal_basic_bare
p_ord_bare <- plot(np_ord)
p_ord_bare
```

### Adding scales, labels, and theme

```{r}
#| label: np_ordinal_basic_decorated
p_ord <- p_ord_bare +
  ggplot2::scale_colour_manual(
    values = c(
      "None"     = "grey40",
      "Mild"     = "steelblue",
      "Moderate" = "darkorange",
      "Severe"   = "firebrick"
    ),
    name = "AR Grade"
  ) +
  ggplot2::scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  ggplot2::labs(
    x = "Follow-up (years)",
    y = "Grade prevalence (%)"
  ) +
  hv_theme("poster") +
  ggplot2::theme(legend.position = c(0.75, 0.6))

p_ord
```

Each line is one grade level. Points show binned data summary values
(analogous to the SAS `means` dataset with `smntr0`–`smntr3` columns).

## Combined mild + moderate (p34 = p3 + p4 pattern)

For analyses that collapse higher grades (e.g. Moderate + Severe combined),
subset and re-label before passing to the constructor:

```{r}
#| label: np_ordinal_collapsed
ord_collapsed <- ord_dat
# Merge Moderate and Severe into one level
ord_collapsed <- subset(ord_collapsed, grade %in% c("None", "Mild", "Moderate/Severe"))

# If your data has Moderate and Severe as separate rows, combine with:
# library(dplyr)
# ord_collapsed <- ord_dat |>
#   dplyr::filter(grade %in% c("Moderate", "Severe")) |>
#   dplyr::summarise(estimate = sum(estimate), .by = time) |>
#   dplyr::mutate(grade = "Moderate/Severe") |>
#   dplyr::bind_rows(dplyr::filter(ord_dat, grade %in% c("None", "Mild")))

# Show three-level version using only two grades from sample data
ord_two <- subset(ord_dat, grade %in% c("None", "Severe"))

plot(hv_ordinal(curve_data = ord_two)) +
  ggplot2::scale_colour_manual(
    values = c("None" = "steelblue", "Severe" = "firebrick"),
    name   = NULL
  ) +
  ggplot2::scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                               labels = scales::percent) +
  ggplot2::labs(x = "Follow-up (years)", y = "Grade prevalence (%)") +
  hv_theme("poster")
```

## Saving

```{r}
#| label: np_ordinal_save
#| eval: false
ggsave(here::here("graphs", "np_tr_ordinal.pdf"),
       p_ord, width = 11, height = 8.5)
```

# Longitudinal Participation Counts

`hv_longitudinal()` reproduces the two-panel layout from
`tp.dp.longitudinal_patients_measures.*`: a grouped bar chart of patients and
measurements at each follow-up window, paired with a numeric summary table
below. Call `plot(lc)` for the bar chart and `plot(lc, type = "table")` for the
table panel. Compose the two panels with `patchwork`.

Input is **pre-aggregated long-format data** — one row per (time window, series)
combination. Use `sample_longitudinal_counts_data()` to derive this from
patient-level data, or build it yourself from your registry data.

## Sample data

```{r}
#| label: lc_data
lc_dat <- sample_longitudinal_counts_data(n_patients = 300, seed = 42L)
lc_dat

# Build the S3 object once; use for both panels
lc <- hv_longitudinal(lc_dat)
```

## Bare plot

```{r}
#| label: lc_bare
p_lc_bare <- plot(lc)
p_lc_bare
```

## Bar chart

```{r}
#| label: lc_bar
p_lc_bar <- plot(lc) +
  ggplot2::scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq(0, 2000, 500),
    expand = c(0, 0)
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 2200)) +
  ggplot2::labs(x = NULL, y = "Count (n)") +
  hv_theme("poster") +
  ggplot2::theme(legend.position = c(0.85, 0.85))

p_lc_bar
```

## Numeric table panel

```{r}
#| label: lc_table
p_lc_tbl <- plot(lc, type = "table") +
  ggplot2::scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  ) +
  hv_theme("poster")

p_lc_tbl
```

## Two-panel layout with patchwork

```{r}
#| label: lc_combined
#| fig-height: 5
library(patchwork)

p_lc_bar / p_lc_tbl +
  patchwork::plot_layout(heights = c(3, 1))
```

## Saving the combined figure

```{r}
#| label: lc_save
#| eval: false
combined <- p_lc_bar / p_lc_tbl +
  patchwork::plot_layout(heights = c(3, 1))

ggsave(here::here("graphs", "longitudinal_participation.pdf"),
       combined, width = 11, height = 6)

# PowerPoint (bar chart only — patchwork composites may need ggsave first)
save_ppt(p_lc_bar,
         template   = system.file("ClevelandClinic.pptx", package = "hvtiPlotR"),
         powerpoint = here::here("graphs", "longitudinal_participation.pptx"))
```

# UpSet Plot

`hv_upset()` wraps `ComplexUpset::upset()` to visualise surgical procedure
co-occurrences or any set membership data. It ports the pattern from
`tp.complexUpset.R`, replacing hard-coded colours with `scale_fill_*`
composition and hard-coded themes with `hv_theme()`.

Unlike other hvtiPlotR functions, `plot.hv_upset` returns a **patchwork
composite** built internally by ComplexUpset. A theme must be applied via
the patchwork `&` operator (not `+`) to cover all sub-panels simultaneously.
`&` is also required for correct rendering: ComplexUpset's internal theme
sets `axis.title.x` in a form that newer ggplot2 rejects unless overridden,
so `& hv_theme()` is part of the *minimum* working call, not an optional
decoration:

```r
# Minimum working call — & hv_theme() is required, not optional
plot(hu) & hv_theme("poster")
```

## Sample data

```{r}
#| label: upset_data
#| eval: !expr .Platform$OS.type != "windows"
sets <- c("AV_Replacement", "AV_Repair", "MV_Replacement", "MV_Repair",
          "TV_Repair", "Aorta", "CABG")

upset_dta <- sample_upset_data(n = 400, seed = 42)
head(upset_dta)
colSums(upset_dta)

hu <- hv_upset(upset_dta, intersect = sets)
```

## Default plot

```{r}
#| label: upset_basic
#| fig-width: 9
#| fig-height: 5
#| eval: !expr .Platform$OS.type != "windows"
plot(hu) &
  hv_theme("poster")
```

## Custom intersection bar colour (scale_fill_manual)

```{r}
#| label: upset_fill
#| fig-width: 9
#| fig-height: 5
#| eval: !expr .Platform$OS.type != "windows"
plot(hu,
     base_annotations = list(
       "Intersection size" = ComplexUpset::intersection_size(
         mapping = ggplot2::aes(fill = "n")
       ) +
         ggplot2::scale_fill_manual(
           values = c("n" = "steelblue"),
           guide  = "none"
         ) +
         ggplot2::labs(y = "Patients (n)")
     )) &
  hv_theme("poster")
```

## Colour bars by era

```{r}
#| label: upset_era
#| fig-width: 9
#| fig-height: 5
#| eval: !expr .Platform$OS.type != "windows"
upset_dta$era <- ifelse(seq_len(nrow(upset_dta)) <= 200, "Early", "Recent")
hu_era <- hv_upset(upset_dta, intersect = sets)

plot(hu_era,
     base_annotations = list(
       "Intersection size" = ComplexUpset::intersection_size(
         counts  = FALSE,
         mapping = ggplot2::aes(fill = era)
       ) +
         ggplot2::scale_fill_manual(
           values = c("Early" = "grey60", "Recent" = "steelblue"),
           name   = "Era"
         ) +
         ggplot2::labs(y = "Patients (n)")
     )) &
  hv_theme("poster")
```

## Saving

```{r}
#| label: upset_save
#| eval: false
p_upset <- plot(hu) & hv_theme("poster")

# UpSet plots are patchwork composites — use ggsave() via ggplot2
ggplot2::ggsave(here::here("graphs", "procedure_cooccurrence.pdf"),
                p_upset, width = 11, height = 6)
```

# Draft Footnotes

`make_footnote()` writes a small annotation to the **bottom-right corner** of
the current graphics device using grid. Call it *after* printing a plot to mark
the figure as a work-in-progress during analysis. For publication-ready output,
simply omit the call — the plot itself is unchanged.

```
# During analysis                   # For publication
print(p)                             ggsave("fig1.pdf", p, ...)
make_footnote("R/analysis.R")        # <- no footnote call
```

## Draft annotation pattern

```{r}
#| label: footnote_draft
#| fig-width: 7
#| fig-height: 5
dta_fn <- sample_hazard_data(n = 300, time_max = 10)
emp_fn <- sample_hazard_empirical(n = 300, time_max = 10, n_bins = 5)

p_draft <- hazard_plot(
  dta_fn,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_fn,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)") +
  hv_theme("poster")

print(p_draft)
make_footnote("vignettes/plot-functions.qmd")
```

## Custom analyst tag

Pass any string as `text` and set `timestamp = FALSE` for a stable label.

```{r}
#| label: footnote_analyst
#| fig-width: 7
#| fig-height: 5
print(p_draft)
make_footnote(
  text      = paste("J. Ehrlinger |", Sys.Date()),
  timestamp = FALSE,
  prefix    = ""
)
```

## Saving without the footnote

The `ggsave()` call writes the ggplot object directly — `make_footnote()` is
never called, so the saved file is clean.

```{r}
#| label: footnote_save
#| eval: false
ggsave("../graphs/fig1_survival.pdf", p_draft, width = 11, height = 8.5)
```
