---
title: "SAS Template Migration Guide: Finding Your Plot in R"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{SAS Template Migration Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
# On Windows, R CMD check sets R_LIBS but Quarto subprocesses may not inherit
# the full library path. Explicitly add it before loading the package.
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(hvtiPlotR)
library(ggplot2)
```

# Overview

This guide maps every SAS template in the HVTI statistics group template
library to its R equivalent in the **hvtiPlotR** package. If you know the
SAS template name (e.g., `tp.np.afib.ivwristm.avrg_curv.binary.sas`), look it
up in the table below and jump to the corresponding section for a working R
example.

The guide is organized by template family (the two-letter prefix after `tp.`).
New ports are added to this document as they become available.

## Key concepts for SAS users

**ggplot2 builds plots in layers.** Instead of one macro call with many
`color=`, `xaxis=`, and `footnote=` options, you chain `+` operations:

```r
np <- hvti_nonparametric(dat, ...)
plot(np) +
  scale_colour_manual(values = c("black", "gray40")) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Years after Operation", y = "Prevalence (%)") +
  hvti_theme("manuscript")
```

**`hvti_theme()` replaces SAS `device=` / style options.** Use
`"dark_ppt"` / `"ppt"` for PowerPoint slides, `"light_ppt"` for
light-background slides, `"manuscript"` for journal figures, and
`"poster"` for conference posters.

**Functions return a ggplot object; they do not display it.** Call the
plot object at the top level (or use `print()`) to render it, or save with
`ggsave()` / `save_ppt()`.

**Pre-fitted models, not raw data.** Most functions accept the *output*
of a statistical model (curve datasets, probability estimates) rather than
individual patient records — mirroring the SAS workflow where `%decompos()`
or `%kaplan` computes estimates and a separate template step produces the
figure.

---

# Template lookup table

| SAS Template | Family | R Constructor | Section |
|---|---|---|---|
| `tp.np.afib.ivwristm.avrg_curv.binary.sas` | np | `hvti_nonparametric()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.afib.ivwristm.pt_spec_phases.binary.sas` | np | `hvti_nonparametric()` | [Phase decomposition](#np-phases) |
| `tp.np.afib.ivwristm.pt_specific.binary.sas` | np | `hvti_nonparametric()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.afib.mult.avrg_curv.binary.sas` | np | `hvti_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.afib.mult.pt_spec.binary.sas` | np | `hvti_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.avpkgrad_ozak_ind_mtwt.sas` | np | `hvti_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.fev.double.univariate.continuous.sas` | np | `hvti_nonparametric()` | [Continuous outcome](#np-continuous) |
| `tp.np.fev.multivariate.continuous.sas` | np | `hvti_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.fev.u.trend.continuous.sas` | np | `hvti_nonparametric()` | [Continuous outcome](#np-continuous) |
| `tp.np.tr.icdpr.avg_curv.sas` | np | `hvti_nonparametric()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.tr.ivecho.average_curv.ordinal.sas` | np | `hvti_ordinal()` | [Ordinal outcomes](#np-ordinal) |
| `tp.np.tr.ivecho.independence.sas` | np | `hvti_ordinal()` | [Ordinal independence](#np-ordinal-independence) |
| `tp.np.tr.ivecho.u.phases.sas` | np | `hvti_ordinal()` | [Ordinal phases](#np-ordinal-phases) |
| `tp.np.po_ar.u_multi.ordinal.sas` | np | `hvti_ordinal()` | [Ordinal multi-scenario](#np-ordinal-multi) |
| `tp.np.z0axdpo.continuous.bmi_xaxis.sas` | np | `hvti_nonparametric()` | [Covariate x-axis](#np-covariate-xaxis) |
| `tp.ac.dead.sas` (via `%kaplan` / `%nelsont`) | ac | `hvti_survival()` | [Kaplan–Meier survival](#ac-dead) |
| `tp.cp.dead.sas` | cp | `hvti_survival()` | [Kaplan–Meier survival](#ac-dead) |
| `tp.dp.gfup.R` | dp | `hvti_followup()` | [Goodness of follow-up](#dp-gfup) |
| `tp.lp.propen.cov_balance.R` | lp | `hvti_balance()` | [Covariate balance](#lp-covbal) |
| `tp.dp.female_bicus_preAR_sankey.R` | dp | `hvti_alluvial()` | [Alluvial](#dp-sankey) |
| PAM cluster stability analysis | dp | `hvti_sankey()` | [Cluster stability Sankey](#dp-cluster-sankey) |
| `tp.complexUpset.R` | dp | `hvti_upset()` | [UpSet plot](#dp-upset) |
| `tp.dp.spaghetti.echo.R` | dp | `hvti_spaghetti()` | [Spaghetti / individual trajectories](#dp-spaghetti) |
| `tp.rp.trends.sas` | rp | `hvti_trends()` | [Trends over time](#dp-trends) |
| `tp.lp.trends.sas` | lp | `hvti_trends()` | [Trends over time](#dp-trends) |
| `tp.lp.trends.age.sas` | lp | `hvti_trends()` | [Trends over time](#dp-trends) |
| `tp.lp.trends.polytomous.sas` | lp | `hvti_trends()` | [Trends over time](#dp-trends) |
| `tp.dp.trends.R` | dp | `hvti_trends()` | [Trends over time](#dp-trends) |
| `tp.dp.longitudinal_patients_measures.R` | dp | `hvti_longitudinal()` | [Longitudinal counts](#dp-long-counts) |
| `tp.lp.mirror-histogram_SAVR-TF-TAVR.R` | lp | `hvti_mirror_hist()` | [Mirror histogram — binary-match](#dp-mirror) |
| `tp.lp.mirror_histo_before_after_wt.R` | lp | `hvti_mirror_hist(weight_col = ...)` | [Mirror histogram — weighted IPTW](#dp-mirror) |
| Stacked histogram | dp | `hvti_stacked()` | [Stacked histogram](#dp-stacked) |
| `tp.hs.dead.setup.sas` | hs | `hazard_plot()` | [Parametric hazard/survival](#hs-dead) |
| `tp.hs.dead_uses_setup.sas` | hs | `hazard_plot()` | [Parametric hazard/survival](#hs-dead) |
| `tp.hs.dead.procedure.tdepth.sas` | hs | `hazard_plot()` | [Parametric hazard/survival](#hs-dead) |
| `tp.hs.dead.conditional.setup.sas` | hs | `hazard_plot()` | [Conditional survival](#hs-dead) |
| `tp.hs.dead.conditional.uses_setup.sas` | hs | `hazard_plot()` | [Conditional survival](#hs-dead) |
| `tp.hs.dead.compare_benefit.setup.sas` | hs | `survival_difference_plot()` | [Treatment benefit distribution](#hs-dead) |
| `tp.hs.uslife_estimates_generate_stratify_.age.sas` | hs | `hazard_plot()` | [US life-table overlay](#hs-dead) |
| `tp.hs.uslife_generates_matched_estimates.sas` | hs | `hazard_plot()` | [US life-table overlay](#hs-dead) |

---

# Nonparametric temporal trends (`tp.np.*`) {#np}

All `tp.np.*` templates share a common SAS workflow:

1. Fit patient-specific temporal profiles with `%decompos()`.
2. Average profiles across patients with `PROC SUMMARY` → `mean_curv` dataset.
3. Optionally compute bootstrap confidence intervals → `boots_ci` dataset.
4. Compute binned patient-level summaries (deciles/quintiles) → `means` dataset.
5. Call the plotting template.

The R port replaces step 5. Steps 1–4 still run in SAS; export the resulting
datasets to CSV and read them into R:

```r
curve_data <- read.csv("mean_curv.csv")   # or boots_ci if CI bands are needed
data_pts   <- read.csv("means.csv")       # optional binned data points
```

## SAS column name mapping

| SAS column | R argument | Notes |
|---|---|---|
| `iv_echo` / `iv_wristm` / `iv_fyrs` | `x_col` | Time variable |
| `prev` / `mnprev` / `_p_` / `est` | `estimate_col` | Curve estimate |
| `cll_p68` / `cll_p95` | `lower_col` | CI lower band |
| `clu_p68` / `clu_p95` | `upper_col` | CI upper band |
| Grouping variable (e.g., `group`) | `group_col` | NULL for single curve |
| `mtime` / `mmtime` (means dataset) | `dp_x_col` | Binned point x |
| `mprev` / `mnprev` (means dataset) | `dp_y_col` | Binned point y |

## Average curve — binary outcome {#np-binary-avg}

**Ports:** `tp.np.afib.ivwristm.avrg_curv.binary.sas`,
`tp.np.afib.ivwristm.pt_specific.binary.sas`,
`tp.np.tr.icdpr.avg_curv.sas`

```{r np-binary-avg-example}
dat     <- sample_nonparametric_curve_data(
  n            = 500,
  time_max     = 12,
  outcome_type = "probability"
)
dat_pts <- sample_nonparametric_curve_points(n = 500, time_max = 12)

# Minimal plot
plot(hvti_nonparametric(dat)) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation Prevalence"
  ) +
  hvti_theme("manuscript")
```

Add 68% CI bands (one standard error; matches SAS `boots_ci` with `cll_p68`
/ `clu_p68`):

```{r np-binary-avg-ci}
plot(hvti_nonparametric(
  dat,
  lower_col = "lower",
  upper_col = "upper"
)) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation — 68% CI"
  ) +
  hvti_theme("manuscript")
```

Add binned data summary points (matches the SAS `means` dataset):

```{r np-binary-avg-points}
plot(hvti_nonparametric(
  dat,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_pts
)) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation with Binned Observations"
  ) +
  hvti_theme("manuscript")
```

## Continuous outcome {#np-continuous}

**Ports:** `tp.np.fev.double.univariate.continuous.sas`,
`tp.np.fev.u.trend.continuous.sas`

For continuous outcomes (FEV1, AV peak gradient) set `outcome_type =
"continuous"`. The y-axis label and scale change accordingly; everything
else is identical.

```{r np-continuous-example}
dat_cont     <- sample_nonparametric_curve_data(
  n            = 400,
  time_max     = 10,
  outcome_type = "continuous"
)
dat_cont_pts <- sample_nonparametric_curve_points(
  n            = 400,
  time_max     = 10,
  outcome_type = "continuous"
)

plot(hvti_nonparametric(
  dat_cont,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_cont_pts
)) +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(
    x     = "Years after operation",
    y     = expression(FEV[1] ~ (L)),
    title = "FEV\u2081 Temporal Trend"
  ) +
  annotate("text",
    x = 9, y = 3.5,
    label = "68% CI band",
    hjust = 1, size = 3
  ) +
  hvti_theme("manuscript")
```

## Multi-group comparison {#np-multigroup}

**Ports:** `tp.np.afib.mult.avrg_curv.binary.sas`,
`tp.np.afib.mult.pt_spec.binary.sas`,
`tp.np.avpkgrad_ozak_ind_mtwt.sas`,
`tp.np.fev.multivariate.continuous.sas`

Provide a named vector to `groups` when generating sample data, or set
`group_col` when supplying your own curve data.

```{r np-multigroup-example}
grp_def <- c("Ozaki" = 0.7, "CE-Pericardial" = 1.1, "Homograft" = 1.4)
dat_grp     <- sample_nonparametric_curve_data(
  n        = 600,
  time_max = 12,
  groups   = grp_def
)
dat_grp_pts <- sample_nonparametric_curve_points(
  n        = 600,
  time_max = 12,
  groups   = grp_def
)

plot(hvti_nonparametric(
  dat_grp,
  group_col   = "group",
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_grp_pts
)) +
  scale_colour_manual(
    values = c("Ozaki" = "#003087", "CE-Pericardial" = "#CC0000", "Homograft" = "#666666")
  ) +
  scale_fill_manual(
    values = c("Ozaki" = "#003087", "CE-Pericardial" = "#CC0000", "Homograft" = "#666666")
  ) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Prevalence (%)",
    colour = "Valve type",
    fill   = "Valve type",
    title  = "Atrial Fibrillation by Valve Type"
  ) +
  hvti_theme("manuscript")
```

## Phase decomposition {#np-phases}

**Ports:** `tp.np.afib.ivwristm.pt_spec_phases.binary.sas`,
`tp.np.tr.ivecho.u.phases.sas`

Phase plots separate the early (bell-shaped, incomplete hazard) and late
(Weibull CDF, complete hazard) components. Supply a `group_col` whose
levels label the phases.

```{r np-phases-example}
dat_phase <- sample_nonparametric_curve_data(
  n      = 500,
  groups = c("Early phase" = 1.5, "Late phase" = 0.6, "Overall" = 1.0)
)

plot(hvti_nonparametric(
  dat_phase,
  group_col = "group"
)) +
  scale_colour_manual(
    values = c("Early phase" = "#CC0000",
               "Late phase"  = "#003087",
               "Overall"     = "black")
  ) +
  scale_linetype_manual(
    values = c("Early phase" = "dashed",
               "Late phase"  = "dashed",
               "Overall"     = "solid")
  ) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Prevalence (%)",
    colour = NULL,
    title  = "AF — Early and Late Phase Decomposition"
  ) +
  hvti_theme("manuscript")
```

## Covariate on the x-axis {#np-covariate-xaxis}

**Port:** `tp.np.z0axdpo.continuous.bmi_xaxis.sas`

When BMI or another continuous covariate (rather than time) is on the x-axis,
the function signature is identical — simply pass the covariate column name to
`x_col`.

```{r np-bmi-xaxis}
# Simulate covariate (BMI) x-axis data
set.seed(42)
n_pts  <- 300
bmi    <- seq(18, 45, length.out = n_pts)
est    <- plogis(-3 + 0.08 * bmi)
se     <- sqrt(est * (1 - est) / 50)
bmi_curve <- data.frame(
  bmi   = bmi,
  est   = est,
  lower = pmax(0, est - qnorm(0.84) * se),
  upper = pmin(1, est + qnorm(0.84) * se)
)

plot(hvti_nonparametric(
  bmi_curve,
  x_col        = "bmi",
  estimate_col = "est",
  lower_col    = "lower",
  upper_col    = "upper"
)) +
  scale_x_continuous(
    breaks = seq(18, 45, 3),
    limits = c(18, 45)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1),
    labels = scales::percent
  ) +
  labs(
    x     = expression(BMI ~ (kg/m^2)),
    y     = "Estimated Probability",
    title = "Outcome Probability vs. BMI at Operation"
  ) +
  hvti_theme("manuscript")
```

## Ordinal outcomes {#np-ordinal}

**Port:** `tp.np.tr.ivecho.average_curv.ordinal.sas`

Ordinal templates (TR grade, AR grade) use `hvti_ordinal()`.
The critical migration step is reshaping the SAS wide-format `predict`
dataset (one column per grade) to long format.

**SAS reshape (do this before reading into R):**

```sas
/* SAS wide format: p0, p1, p2, p3 */
data predict_wide;
  set predict;
run;
```

**R reshape:**

```r
library(tidyr)
long <- pivot_longer(
  predict_wide,
  cols      = c(p0, p1, p2, p3),
  names_to  = "grade",
  values_to = "estimate"
)
```

```{r np-ordinal-example}
ord_labels <- c("None", "Mild", "Moderate", "Severe")
dat_ord     <- sample_nonparametric_ordinal_data(
  n            = 1000,
  time_max     = 5,
  grade_labels = ord_labels
)
dat_ord_pts <- sample_nonparametric_ordinal_points(
  n            = 1000,
  time_max     = 5,
  grade_labels = ord_labels
)

plot(hvti_ordinal(
  dat_ord,
  grade_col   = "grade",
  data_points = dat_ord_pts
)) +
  scale_colour_manual(
    values = c(
      "None"     = "#003087",
      "Mild"     = "#55A51C",
      "Moderate" = "#FFA500",
      "Severe"   = "#CC0000"
    )
  ) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Grade probability",
    colour = "TR Grade",
    title  = "Tricuspid Regurgitation Grade"
  ) +
  hvti_theme("manuscript")
```

### Ordinal multi-scenario comparison {#np-ordinal-multi}

**Port:** `tp.np.po_ar.u_multi.ordinal.sas`

```{r np-ordinal-multi-example}
dat_ar1 <- sample_nonparametric_ordinal_data(seed = 1)
dat_ar2 <- sample_nonparametric_ordinal_data(seed = 99)

dat_ar1$scenario <- "Before repair"
dat_ar2$scenario <- "After repair"
combined <- rbind(dat_ar1, dat_ar2)

plot(hvti_ordinal(combined, grade_col = "grade")) +
  facet_wrap(~scenario) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    x      = "Years",
    y      = "Grade probability",
    colour = "AR Grade",
    title  = "AR Grade — Before vs. After Repair"
  ) +
  hvti_theme("manuscript")
```

### Ordinal phase independence {#np-ordinal-independence}

**Port:** `tp.np.tr.ivecho.independence.sas`

To examine a single grade in isolation, filter the long-format curve
before passing it to the function:

```{r np-ordinal-independence-example}
dat_ind    <- sample_nonparametric_ordinal_data(n = 800)
pts_ind    <- sample_nonparametric_ordinal_points(n = 800)
grade_2    <- dat_ind[dat_ind$grade == "Grade 2", ]
dp_grade_2 <- pts_ind[pts_ind$grade == "Grade 2", ]

plot(hvti_ordinal(
  grade_2,
  grade_col   = "grade",
  data_points = dp_grade_2
)) +
  scale_colour_manual(values = c("Grade 2" = "#CC0000")) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(
    limits = c(0, 0.5),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Probability",
    title = "Probability of TR Grade 2"
  ) +
  hvti_theme("manuscript")
```

### Ordinal phases {#np-ordinal-phases}

**Port:** `tp.np.tr.ivecho.u.phases.sas`

Phase-decomposed ordinal plots combine phase labels with grade colours.
Create the figure by plotting grade-specific curves and annotating early
vs. late phase regions:

```{r np-ordinal-phases-example}
dat_ph <- sample_nonparametric_ordinal_data(n = 800, seed = 7)

plot(hvti_ordinal(dat_ph, grade_col = "grade")) +
  annotate("rect",
    xmin = 0, xmax = 2,  ymin = -Inf, ymax = Inf,
    fill = "steelblue", alpha = 0.07
  ) +
  annotate("text",
    x = 1, y = 0.95, label = "Early\nphase",
    size = 3, colour = "steelblue", fontface = "italic"
  ) +
  annotate("rect",
    xmin = 2, xmax = 5, ymin = -Inf, ymax = Inf,
    fill = "tomato", alpha = 0.07
  ) +
  annotate("text",
    x = 3.5, y = 0.95, label = "Late\nphase",
    size = 3, colour = "tomato", fontface = "italic"
  ) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x      = "Years after operation",
    y      = "Grade probability",
    colour = "TR Grade",
    title  = "TR Grade — Early and Late Phase"
  ) +
  hvti_theme("manuscript")
```

---

# Survival analysis (`tp.ac.dead.*`, `tp.cp.dead.*`) {#ac-dead}

**Ports:** `tp.ac.dead.sas`, `tp.cp.dead.sas`

**R equivalent:** `hvti_survival()`

`hvti_survival()` wraps `survfit()` from the **survival** package and returns
an S3 object. Call `plot(km, type = ...)` to render one of the five plot types
matching the SAS `%kaplan` / `%nelsont` macro output flags (`PLOTS`, `PLOTC`,
`PLOTH`, `PLOTL`). Tidy data frames are accessible via `km$tables`.

```r
dta <- sample_survival_data(n = 500, seed = 42)
km  <- hvti_survival(dta)

# Kaplan–Meier survival (PLOTS)
plot(km) +
  scale_y_continuous(
    breaks = seq(0, 100, 20),
    labels = function(x) paste0(x, "%")
  ) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    x     = "Years after operation",
    y     = "Survival (%)",
    title = "Freedom from Death"
  ) +
  hvti_theme("manuscript")

# Cumulative hazard (PLOTC)
plot(km, type = "cumhaz") +
  labs(x = "Years", y = "Cumulative Hazard") +
  hvti_theme("manuscript")

# Hazard rate (PLOTH)
plot(km, type = "hazard") +
  labs(x = "Years", y = "Instantaneous Hazard") +
  hvti_theme("manuscript")

# Log-log survival (PH check)
plot(km, type = "loglog") +
  labs(x = "log(Years)", y = "log(-log S(t))") +
  hvti_theme("manuscript")

# Integrated survivorship (PLOTL)
plot(km, type = "life") +
  labs(x = "Years", y = "Restricted Mean Survival (years)") +
  hvti_theme("manuscript")

# Access the KM data, risk table, and report table
km$data              # tidy KM data frame
km$tables$risk       # numbers-at-risk table
km$tables$report     # survival at report times
```

### SAS → R argument mapping

| SAS `%kaplan` option | R argument | Notes |
|---|---|---|
| `data=` | `data` | Patient-level data frame |
| `time=` | `time_col` | Time-to-event column name |
| `event=` | `event_col` | Event indicator column (1 = event) |
| `group=` | `group_col` | Stratification variable |
| `method=kaplan` | `method = "kaplan-meier"` | Default |
| `method=nelsont` | `method = "nelson-aalen"` | Fleming–Harrington |
| `alpha=0.05` | `conf_level = 0.95` | Default is 0.95 |
| `tp=0 1 2 3 5 7 10` | `report_times` | Table time points |

---

# Goodness of follow-up (`tp.dp.gfup.R`) {#dp-gfup}

**Port:** `tp.dp.gfup.R`

**R equivalent:** `hvti_followup()`

```r
dta <- sample_goodness_followup_data(n = 300)
gf  <- hvti_followup(dta)

plot(gf) +
  labs(title = "Goodness of Follow-Up") +
  hvti_theme("manuscript")
```

---

# Covariate balance (`tp.lp.propen.cov_balance.R`) {#lp-covbal}

**Port:** `tp.lp.propen.cov_balance.R`

**R equivalent:** `hvti_balance()`

The SAS export arrives wide (one column per time-point). Reshape to long format
before calling `hvti_balance()`. Pass `var_levels` to control the
bottom-to-top display order of covariates (matches `ylabel` in the original
script).

```r
# Reshape wide → long (mirrors the original script)
ylabel <- c(
  "Cardiac Output", "BAV", "Cardiac Index", "NYHA -Functional Class",
  "COPD-Oxy", "Date of Surg", "Creatinine", "Age", "CAD",
  "Female", "LV PWT", "Mismatch", "HTN", "PVD", "Sinitibular Jcn: diam"
)

names(dta) <- c("variable", "Before match", "After match")

dta_long <- reshape(
  dta,
  direction = "long",
  varying   = c("Before match", "After match"),
  v.names   = "std_diff",
  timevar   = "group",
  times     = c("Before match", "After match"),
  idvar     = "variable"
)

n_vars    <- length(ylabel)
cb        <- hvti_balance(
  data         = dta_long,
  variable_col = "variable",
  group_col    = "group",
  std_diff_col = "std_diff",
  var_levels   = ylabel
)

stdifPlot <- plot(cb) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(-40, 30), breaks = seq(-40, 30, 10)) +
  labs(x = "Standardized difference: SAVR - TF:TAVR (%)", y = "") +
  annotate("text", x = -30, y = 0,      label = "More likely TF-TAVR", size = 4.5) +
  annotate("text", x =  22, y = n_vars, label = "More likely SAVR",    size = 4.5) +
  theme(legend.position = c(0.20, 0.935)) +
  hvti_theme("manuscript")

ggsave(here::here("graphs", "lp_cov-balance-SAVR_TF-TAVR.pdf"),
       plot = stdifPlot, height = 7, width = 8)
```

---

# Alluvial flow (`tp.dp.female_bicus_preAR_sankey.R`) {#dp-sankey}

**Port:** `tp.dp.female_bicus_preAR_sankey.R`

**R equivalent:** `hvti_alluvial()`

```r
dta <- sample_alluvial_data(n = 200)
al  <- hvti_alluvial(dta)

plot(al) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Patient Flow Between States") +
  hvti_theme("manuscript")
```

---

# Cluster stability Sankey (PAM analysis) {#dp-cluster-sankey}

**Source:** PAM clustering analysis pipeline (R script using `ggsankey`)

**R equivalent:** `hvti_sankey()`

`hvti_sankey()` draws a Sankey diagram showing how patients flow between
letter-labelled clusters as the number of clusters K increases from 2 to 9.
Each column is one K; flow bands show assignment changes between consecutive K
values; node labels show cluster letter + patient count.

The original code used `ggsankey::make_long()` with hard-coded column names and
ordering vectors. The R port accepts any cluster columns and ordering via
`cluster_cols` and `node_levels` arguments.

**Requires `ggsankey`** (install from GitHub):

```r
remotes::install_github("davidsjoberg/ggsankey")
```

### SAS/R workflow → R equivalent

| Original step | R equivalent |
|---|---|
| `sid_dta$C2 <- factor(...)` with `gr2_names` ordering | Factor levels set by `node_levels` argument |
| `make_long(C2, …, C9)` | `.make_sankey_long()` (internal) |
| `geom_sankey()` + `geom_sankey_label()` | `hvti_sankey()` + `plot()` |
| `brewer.pal(9, "Set1")[c(2,6,8,4,3,5,7,1,9)]` | Default `node_colours` |

```r
dta_san <- sample_cluster_sankey_data(n = 300, seed = 42)
sk      <- hvti_sankey(dta_san)

plot(sk) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  hvti_theme("manuscript")
```

To reproduce the original analysis with patient-level cluster assignments:

```r
# Build a data frame with one row per patient, columns C2..C9 (factor)
grp_dta <- data.frame(
  C2 = factor(pm2clusters_labels, levels = gr2_names),
  C3 = factor(pm3clusters_labels, levels = gr3_names),
  # ... etc.
  C9 = factor(pm9clusters_labels, levels = gr9_names)
)

sk_grp <- hvti_sankey(
  grp_dta,
  cluster_cols = paste0("C", 2:9),
  node_levels  = gr9_names
)

plot(sk_grp) +
  labs(x = NULL) +
  hvti_theme("manuscript")
```

---

# UpSet plot (`tp.complexUpset.R`) {#dp-upset}

**Port:** `tp.complexUpset.R`

**R equivalent:** `hvti_upset()`

```r
sets <- c("CABG", "Valve", "MAZE", "Aorta")
dta  <- sample_upset_data(n = 300, sets = sets)
hu   <- hvti_upset(data = dta, intersect = sets)

plot(hu)
```

---

# Spaghetti / individual trajectories (`tp.dp.spaghetti.echo.R`) {#dp-spaghetti}

**Port:** `tp.dp.spaghetti.echo.R`

**R equivalent:** `hvti_spaghetti()`

The template covers nine figures across three echo outcomes (AV mean gradient,
AV area, DVI) in unstratified and sex-stratified variants, plus an ordinal MV
regurgitation grade plot (plot_9).

```r
dta    <- sample_spaghetti_data(n_patients = 150, max_obs = 6)
sp     <- hvti_spaghetti(dta)
sp_col <- hvti_spaghetti(dta, colour_col = "group")
```

## Unstratified — AV mean gradient full range (plot_1)

```r
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hvti_theme("manuscript")
```

## Unstratified — zoomed y-axis (plot_3)

```r
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 30, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 30)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hvti_theme("manuscript")
```

## Stratified by sex (plot_2 / plot_4)

Template uses `values=c("red", "blue")`; modernised equivalents below.

```r
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  hvti_theme("manuscript")
```

## AV area y-scale (plot_5 / plot_6)

```r
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 5, 1)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(x = "Years", y = "AV Area (EOA) (cm\u00b2)") +
  hvti_theme("manuscript")
```

## DVI y-scale (plot_7 / plot_8)

```r
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 1.25, 0.25)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 1.25)) +
  labs(x = "Years", y = "DVI") +
  hvti_theme("manuscript")
```

## Ordinal y-axis — MV regurgitation grade (plot_9)

```r
dta_ord       <- dta
dta_ord$value <- round(pmin(3, pmax(0, dta$value / 12)))
levels(dta_ord$group) <- c("Early", "Late")

sp_ord <- hvti_spaghetti(dta_ord, colour_col = "group")

plot(sp_ord, y_labels = c(None = 0, Mild = 1, Moderate = 2, Severe = 3)) +
  scale_colour_manual(
    values = c(Early = "steelblue", Late = "red2"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 6, 1)) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 3)) +
  labs(x = "Years after Procedure", y = "MV Regurgitation Grade") +
  hvti_theme("manuscript")
```

---

# Trends over time (`tp.*.trends.*`) {#dp-trends}

**Ports:** `tp.rp.trends.sas`, `tp.lp.trends.sas`, `tp.lp.trends.age.sas`,
`tp.lp.trends.polytomous.sas`, `tp.dp.trends.R`

**R equivalent:** `hvti_trends()`

## tp.rp.trends.sas — cases/year and age (1968–2000 by 4)

```r
one    <- sample_trends_data(n = 600, year_range = c(1968L, 2000L), groups = NULL)
tr_one <- hvti_trends(one, group_col = NULL)

# Cases/year: axisy order=(0 to 10 by 2)
plot(tr_one) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 10),      breaks = seq(0, 10, 2)) +
  labs(x = "Year", y = "Cases/year") +
  hvti_theme("manuscript")

# Age: axisy order=(30 to 70 by 10)
plot(tr_one) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(30, 70),     breaks = seq(30, 70, 10)) +
  labs(x = "Year", y = "Age (years)") +
  hvti_theme("manuscript")
```

## tp.lp.trends.sas — binary % outcomes (1970–2000 by 10, y 0–100 by 10)

Multiple outcomes on one figure. CGM version: `axisx order=(1970 to 2000 by 10)`,
`axisy order=(0 to 100 by 20)`.

```r
dta_lp <- sample_trends_data(
  n = 800, year_range = c(1970L, 2000L),
  groups = c("Shock %", "Pre-op IABP %", "Inotropes %"))
tr_lp <- hvti_trends(dta_lp)

plot(tr_lp) +
  scale_colour_manual(
    values = c("Shock %" = "steelblue", "Pre-op IABP %" = "firebrick",
               "Inotropes %" = "forestgreen"), name = NULL) +
  scale_shape_manual(
    values = c("Shock %" = 16L, "Pre-op IABP %" = 15L, "Inotropes %" = 17L),
    name = NULL) +
  scale_x_continuous(limits = c(1970, 2000), breaks = seq(1970, 2000, 10)) +
  scale_y_continuous(limits = c(0, 100),     breaks = seq(0, 100, 10)) +
  coord_cartesian(xlim = c(1970, 2000), ylim = c(0, 100)) +
  labs(x = "Year", y = "Percent (%)") +
  hvti_theme("manuscript")
```

## tp.lp.trends.age.sas — age on x-axis (25–85 by 10, y 0–100 by 20)

```r
dta_age <- sample_trends_data(
  n = 600, year_range = c(25L, 85L),
  groups = c("Repair %", "Bioprosthesis %"), seed = 7L)
tr_age <- hvti_trends(dta_age)

plot(tr_age) +
  scale_colour_manual(
    values = c("Repair %" = "steelblue", "Bioprosthesis %" = "firebrick"),
    name = NULL) +
  scale_x_continuous(limits = c(25, 85), breaks = seq(25, 85, 10)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  coord_cartesian(xlim = c(25, 85), ylim = c(0, 100)) +
  labs(x = "Age (years)", y = "Percent (%)") +
  hvti_theme("manuscript")
```

## tp.lp.trends.polytomous.sas — repair types (1990–1999 by 1, y 0–100 by 10)

```r
dta_poly <- sample_trends_data(
  n = 800, year_range = c(1990L, 1999L),
  groups = c("CE", "Cosgrove", "Periguard", "DeVega"), seed = 5L)
tr_poly <- hvti_trends(dta_poly)

plot(tr_poly) +
  scale_colour_manual(
    values = c(CE = "steelblue", Cosgrove = "firebrick",
               Periguard = "forestgreen", DeVega = "goldenrod3"),
    name = "Repair type") +
  scale_shape_manual(
    values = c(CE = 15L, Cosgrove = 19L, Periguard = 17L, DeVega = 18L),
    name = "Repair type") +
  scale_x_continuous(limits = c(1990, 1999), breaks = seq(1990, 1999, 1)) +
  scale_y_continuous(limits = c(0, 100),     breaks = seq(0, 100, 10)) +
  coord_cartesian(xlim = c(1990, 1999), ylim = c(0, 100)) +
  labs(x = "Year", y = "Percent (%)") +
  hvti_theme("manuscript")
```

## tp.dp.trends.R — LV mass index (1995–2015 by 5, y 0–200 by 50)

```r
dta_lv <- sample_trends_data(n = 800, year_range = c(1995L, 2015L),
                              groups = NULL, seed = 3L)
tr_lv  <- hvti_trends(dta_lv, group_col = NULL)

plot(tr_lv) +
  scale_x_continuous(limits = c(1995, 2015), breaks = seq(1995, 2015, 5)) +
  scale_y_continuous(limits = c(0, 200),     breaks = seq(0, 200, 50)) +
  coord_cartesian(xlim = c(1995, 2015), ylim = c(0, 200)) +
  labs(x = "Years", y = "LV Mass Index") +
  hvti_theme("manuscript")
```

## tp.dp.trends.R — hospital LOS with annotation (1985–2015 by 5, y 0–20 by 5)

```r
dta_los <- sample_trends_data(n = 800, year_range = c(1985L, 2015L),
                               groups = NULL, seed = 11L)
tr_los  <- hvti_trends(dta_los, group_col = NULL)

plot(tr_los) +
  scale_x_continuous(limits = c(1985, 2015), breaks = seq(1985, 2015, 5)) +
  scale_y_continuous(limits = c(0, 20),      breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(1985, 2015), ylim = c(0, 20)) +
  annotate("text", x = 1995, y = 18,
           label = "Trend: Hospital Length of Stay", size = 4.5) +
  labs(x = "Years", y = "Hospital LOS (Days)") +
  hvti_theme("manuscript")
```

---

# Longitudinal patient counts (`tp.dp.longitudinal_patients_measures.R`) {#dp-long-counts}

**Port:** `tp.dp.longitudinal_patients_measures.R`

**R equivalent:** `hvti_longitudinal()`

```r
dta <- sample_longitudinal_counts_data(n_patients = 150)
lc  <- hvti_longitudinal(dta)

plot(lc) +
  scale_x_continuous(breaks = 0:10) +
  labs(
    x     = "Years after operation",
    y     = "Patients with measurement",
    title = "Follow-Up Echocardiogram Availability"
  ) +
  hvti_theme("manuscript")
```

---

# Mirror histogram (propensity score) {#dp-mirror}

**R equivalents:** `tp.lp.mirror-histogram_SAVR-TF-TAVR.R` (binary-match),
`tp.lp.mirror_histo_before_after_wt.R` (weighted IPTW)

Both scripts are superseded by `hvti_mirror_hist()`. Call `plot(mh)` to render the
histogram. Diagnostics are accessible via `mh$tables$diagnostics`; working
data via `mh$tables$working`. Compose scales, annotations, and theme with the
usual `+` operator.

## Binary-match mode (`tp.lp.mirror-histogram_SAVR-TF-TAVR.R`) {#dp-mirror-binary}

```r
dta <- sample_mirror_histogram_data(n = 400, separation = 1.5)
mh  <- hvti_mirror_hist(dta)   # defaults: prob_t / tavr / match

plot(mh) +
  scale_fill_manual(
    values = c(
      before_g0  = "white",  matched_g0 = "green1",
      before_g1  = "white",  matched_g1 = "green4"
    ),
    guide = "none"
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  annotate("text", x = 20, y =  100, label = "SAVR",    size = 7) +
  annotate("text", x = 20, y = -100, label = "TF-TAVR", size = 7) +
  labs(x = "Propensity Score (%)", y = "Number of Patients") +
  hvti_theme("manuscript")
```

## Weighted IPTW mode (`tp.lp.mirror_histo_before_after_wt.R`) {#dp-mirror-weighted}

```r
dta   <- sample_mirror_histogram_data(n = 400, add_weights = TRUE)
mh_wt <- hvti_mirror_hist(
  dta,
  group_labels = c("Limited", "Extended"),
  weight_col   = "mt_wt"
)

plot(mh_wt) +
  scale_fill_manual(
    values = c(
      before_g0   = "white", weighted_g0 = "blue",
      before_g1   = "white", weighted_g1 = "red"
    ),
    guide = "none"
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  annotate("text", x = 30, y =  150, label = "Limited",  color = "blue", size = 5) +
  annotate("text", x = 30, y = -70,  label = "Extended", color = "red",  size = 5) +
  labs(x = "Propensity Score (%)", y = "#") +
  hvti_theme("manuscript")
```

---

# Stacked histogram {#dp-stacked}

**R equivalent:** `hvti_stacked()`

```r
dta <- sample_stacked_histogram_data()
sh  <- hvti_stacked(dta)

plot(sh) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x     = "Operation year",
    y     = "Count",
    fill  = "Group",
    title = "Annual Case Volume by Group"
  ) +
  hvti_theme("manuscript")
```

---

# Using themes

All `plot()` calls return an unstyled ggplot object. Add a theme as the
final layer:

| Context | Call | Equivalent SAS device |
|---|---|---|
| Dark PowerPoint slide | `hvti_theme("dark_ppt")` | `device=ppt` |
| Light PowerPoint slide | `hvti_theme("light_ppt")` | `device=ppt` with white background |
| Journal manuscript | `hvti_theme("manuscript")` | `device=eps` / PDF |
| Conference poster | `hvti_theme("poster")` | Large-font poster |

You can pass `base_size` to any theme to scale all text simultaneously:

```r
p + hvti_theme("manuscript", base_size = 10)   # smaller text for double-column
p + hvti_theme("poster",     base_size = 24)   # larger text for A0 poster
```

---

# Saving figures

## PowerPoint

`save_ppt()` inserts ggplot objects into a PowerPoint file as **editable
DrawingML vector graphics** via the `officer` and `rvg` packages — shapes,
lines, and text remain individually selectable in PowerPoint after export.
The first argument is `object` (not `plot`); the output path is `powerpoint`
(not `file` or `filename`).

```r
# Locate the bundled Cleveland Clinic slide template
template <- system.file("ClevelandClinic.pptx", package = "hvtiPlotR")

# Single slide — apply a PPT theme before saving
p_ppt <- p +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values   = c("steelblue"), guide = "none") +
  labs(x = "Years", y = "Prevalence (%)") +
  hvti_theme("dark_ppt")

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = "figures/afib_prevalence.pptx",
  slide_titles = "AF Prevalence over Time"
)

# Multiple plots — one slide per figure in a single call
save_ppt(
  object       = list(fig1 = p_binary, fig2 = p_multi),
  template     = template,
  powerpoint   = "figures/np_curves.pptx",
  slide_titles = c("Binary Outcome", "Multi-group Comparison")
)
```

## PDF / TIFF for journals

```r
ggsave("figures/afib_prevalence.pdf",  p, width = 3.5, height = 3.5, units = "in")
ggsave("figures/afib_prevalence.tiff", p, width = 3.5, height = 3.5,
       units = "in", dpi = 600)
```

---

# Parametric hazard/survival (`tp.hs.*`) {#hs-dead}

**Ports:** `tp.hs.dead.setup.sas`, `tp.hs.dead_uses_setup.sas`,
`tp.hs.dead.procedure.tdepth.sas`, `tp.hs.dead.conditional.setup.sas`,
`tp.hs.dead.conditional.uses_setup.sas`,
`tp.hs.dead.compare_benefit.setup.sas`,
`tp.hs.uslife_estimates_generate_stratify_.age.sas`,
`tp.hs.uslife_generates_matched_estimates.sas`

**R equivalents:** `hazard_plot()`, `survival_difference_plot()`

The `tp.hs.*` family extends `tp.hp.dead.*` by using the `%hazpred` macro to
generate patient-specific parametric survival predictions from a fitted
multivariable hazard model. The *setup* templates compute and store:

- **Cumulative hazard per patient** at last follow-up (for observed vs.
  expected goodness-of-fit tests via `%chisqgf`).
- **Individual survivorship curves** on a fine time grid (for mean-curve
  aggregation with `PROC SUMMARY`).

Subsequent *use* templates stratify the aggregated mean curves by a covariate
and overlay Kaplan-Meier estimates from `%kaplan`. The resulting SAS datasets
map directly to `hazard_plot()` in R:

| SAS dataset | Columns used | R source |
|---|---|---|
| `means` (from `PROC SUMMARY` over `predict`) | `SSURVIV`, `SCLLSURV`, `SCLUSURV` | `sample_hazard_data()` |
| `plout` (from `%kaplan`) | `CUM_SURV`, `CL_LOWER`, `CL_UPPER` | `sample_hazard_empirical()` |
| `est.uslife_*` (from `%usmatchd`) | `SMATCHED` | `sample_life_table()` |

## Parametric survival with KM overlay

Reproduces the core figure from `tp.hs.dead_uses_setup.sas` and
`tp.hs.dead.procedure.tdepth.sas`: a mean parametric survival curve (solid line
with confidence band) overlaid with Kaplan-Meier empirical estimates (symbols
with error bars).

```r
dat_hp <- sample_hazard_data(n = 500, time_max = 10)
emp_hp <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)

hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  hvti_theme("manuscript")
```

## Stratified by covariate

`tp.hs.dead.procedure.tdepth.sas` overlays curves for depth-of-invasion groups
(`tdepth = 1, 2, 3`). Pass `group_col` to `hazard_plot()` and supply matched
`sample_hazard_data()` / `sample_hazard_empirical()` calls with a `groups`
vector whose names become the legend labels.

```r
dat_strat <- sample_hazard_data(
  n = 500, time_max = 10,
  groups = c("pT1" = 0.4, "pT2" = 1.0, "pT3" = 1.8)
)
emp_strat <- sample_hazard_empirical(
  n = 500, time_max = 10, n_bins = 5,
  groups = c("pT1" = 0.4, "pT2" = 1.0, "pT3" = 1.8)
)

hazard_plot(
  dat_strat,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  group_col     = "group",
  empirical     = emp_strat,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(
    values = c("pT1" = "steelblue", "pT2" = "forestgreen", "pT3" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("pT1" = "steelblue", "pT2" = "forestgreen", "pT3" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  hvti_theme("manuscript")
```

## Conditional survival after hospital discharge {#hs-conditional}

`tp.hs.dead.conditional.setup.sas` computes individual survivorship curves
starting from the time of hospital discharge rather than the operation date.
The conditional survival `S(t | discharge)` is `S(t) / S(t_discharge)` for
each patient. In R, the same `hazard_plot()` call is used; only the x-axis
label and data preparation differ.

```r
dat_cond <- sample_hazard_data(n = 500, time_max = 10)
emp_cond <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)

hazard_plot(
  dat_cond,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_cond,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Discharge", y = "Survival after Discharge (%)") +
  hvti_theme("manuscript")
```

## US life-table overlay {#hs-uslife}

`tp.hs.uslife_*` templates call `%usmatchd` to generate age-, sex-, and
race-matched US population life-table survival curves for each patient.
`sample_life_table()` provides representative matched curves stratified by age
group for use in R examples.

```r
dat_age <- sample_hazard_data(
  n = 600, time_max = 10,
  groups = c("<60" = 0.4, "60\u201385" = 1.0, "\u226585" = 2.0)
)
emp_age <- sample_hazard_empirical(
  n = 600, time_max = 10, n_bins = 6,
  groups = c("<60" = 0.4, "60\u201385" = 1.0, "\u226585" = 2.0)
)
lt <- sample_life_table(
  age_groups = c("<60", "60\u201385", "\u226585"),
  age_mids   = c(50, 72, 88),
  time_max   = 10
)

hazard_plot(
  dat_age,
  estimate_col     = "survival",
  lower_col        = "surv_lower",
  upper_col        = "surv_upper",
  group_col        = "group",
  empirical        = emp_age,
  emp_lower_col    = "lower",
  emp_upper_col    = "upper",
  reference        = lt,
  ref_estimate_col = "survival",
  ref_group_col    = "group"
) +
  scale_colour_manual(
    values = c("<60" = "steelblue", "60\u201385" = "forestgreen",
               "\u226585" = "firebrick"),
    name   = "Age group"
  ) +
  scale_fill_manual(
    values = c("<60" = "steelblue", "60\u201385" = "forestgreen",
               "\u226585" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Survival (%)",
       caption = "Dashed lines: US population life table") +
  hvti_theme("manuscript")
```

## Treatment benefit distribution {#hs-benefit}

`tp.hs.dead.compare_benefit.setup.sas` computes, for each patient, the
difference in predicted survival at a fixed time point (5 years) under two
treatment arms (e.g. ASA vs. no ASA). The distribution of these individual
differences shows who benefits and by how much. Use `survival_difference_plot()`
to plot the mean curve over time.

```r
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("No ASA" = 1.0, "ASA" = 0.75)
)

survival_difference_plot(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                      colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-5, 40),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival Difference (%)") +
  hvti_theme("manuscript")
```

---

# Adding new ports {#adding-new-ports}

As additional SAS templates are ported to R, update this guide:

1. **Add a row** to the lookup table at the top of this file with the SAS
   template name, family prefix, R function, and section anchor.

2. **Add a section** following the existing pattern: name the port, describe
   the SAS workflow, show the R equivalent with a runnable example, and
   document any column name mapping differences.

3. **Export the new R function** by adding `@export` to its roxygen block and
   running `devtools::document()` to regenerate `NAMESPACE`.

4. **Update `DESCRIPTION`** if a new package dependency is required.

Templates currently planned for porting:

- `tp.ce.states.*` competing-events / state-occupancy → (pending)
- `tp.gp.*` grouped longitudinal ordinal models → (pending)

---

# Session info

```{r session-info}
sessionInfo()
```
