---
title: "SAS Template Migration Guide: Finding Your Plot in R"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{SAS Template Migration Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
library(hvtiPlotR)
library(ggplot2)
```

# Overview

This guide maps every SAS template in the HVTI statistics group template
library to its R equivalent in the **hvtiPlotR** package. If you know the
SAS template name (e.g., `tp.np.afib.ivwristm.avrg_curv.binary.sas`), look it
up in the table below and jump to the corresponding section for a working R
example.

The guide is organized by template family (the two-letter prefix after `tp.`).
New ports are added to this document as they become available.

## Key concepts for SAS users

**ggplot2 builds plots in layers.** Instead of one macro call with many
`color=`, `xaxis=`, and `footnote=` options, you chain `+` operations:

```r
nonparametric_curve_plot(dat$curve, ...) +
  scale_colour_manual(values = c("black", "gray40")) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Years after Operation", y = "Prevalence (%)") +
  hvti_theme("manuscript")
```

**`hvti_theme()` replaces SAS `device=` / style options.** Use
`"dark_ppt"` / `"ppt"` for PowerPoint slides, `"light_ppt"` for
light-background slides, `"manuscript"` for journal figures, and
`"poster"` for conference posters.

**Functions return a ggplot object; they do not display it.** Call the
plot object at the top level (or use `print()`) to render it, or save with
`ggsave()` / `save_ppt()`.

**Pre-fitted models, not raw data.** Most functions accept the *output*
of a statistical model (curve datasets, probability estimates) rather than
individual patient records — mirroring the SAS workflow where `%decompos()`
or `%kaplan` computes estimates and a separate template step produces the
figure.

---

# Template lookup table

| SAS Template | Family | R Function | Section |
|---|---|---|---|
| `tp.np.afib.ivwristm.avrg_curv.binary.sas` | np | `nonparametric_curve_plot()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.afib.ivwristm.pt_spec_phases.binary.sas` | np | `nonparametric_curve_plot()` | [Phase decomposition](#np-phases) |
| `tp.np.afib.ivwristm.pt_specific.binary.sas` | np | `nonparametric_curve_plot()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.afib.mult.avrg_curv.binary.sas` | np | `nonparametric_curve_plot()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.afib.mult.pt_spec.binary.sas` | np | `nonparametric_curve_plot()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.avpkgrad_ozak_ind_mtwt.sas` | np | `nonparametric_curve_plot()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.fev.double.univariate.continuous.sas` | np | `nonparametric_curve_plot()` | [Continuous outcome](#np-continuous) |
| `tp.np.fev.multivariate.continuous.sas` | np | `nonparametric_curve_plot()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.fev.u.trend.continuous.sas` | np | `nonparametric_curve_plot()` | [Continuous outcome](#np-continuous) |
| `tp.np.tr.icdpr.avg_curv.sas` | np | `nonparametric_curve_plot()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.tr.ivecho.average_curv.ordinal.sas` | np | `nonparametric_ordinal_plot()` | [Ordinal outcomes](#np-ordinal) |
| `tp.np.tr.ivecho.independence.sas` | np | `nonparametric_ordinal_plot()` | [Ordinal independence](#np-ordinal-independence) |
| `tp.np.tr.ivecho.u.phases.sas` | np | `nonparametric_ordinal_plot()` | [Ordinal phases](#np-ordinal-phases) |
| `tp.np.po_ar.u_multi.ordinal.sas` | np | `nonparametric_ordinal_plot()` | [Ordinal multi-scenario](#np-ordinal-multi) |
| `tp.np.z0axdpo.continuous.bmi_xaxis.sas` | np | `nonparametric_curve_plot()` | [Covariate x-axis](#np-covariate-xaxis) |
| `tp.ac.dead.sas` (via `%kaplan` / `%nelsont`) | ac | `survival_curve()` | [Kaplan–Meier survival](#ac-dead) |
| `tp.cp.dead.sas` | cp | `survival_curve()` | [Kaplan–Meier survival](#ac-dead) |
| `tp.dp.gfup.R` | dp | `goodness_followup()` | [Goodness of follow-up](#dp-gfup) |
| `tp.lp.propen.cov_balance.R` | lp | `covariate_balance()` | [Covariate balance](#lp-covbal) |
| `tp.dp.female_bicus_preAR_sankey.R` | dp | `sankey_plot()` | [Sankey / alluvial](#dp-sankey) |
| `tp.complexUpset.R` | dp | `upset_plot()` | [UpSet plot](#dp-upset) |
| `tp.dp.spaghetti.echo.R` | dp | `spaghetti_plot()` | [Spaghetti / individual trajectories](#dp-spaghetti) |
| `tp.dp.trends.R` | dp | `trends_plot()` | [Trends over time](#dp-trends) |
| `tp.dp.longitudinal_patients_measures.R` | dp | `longitudinal_counts_plot()` | [Longitudinal counts](#dp-long-counts) |
| Mirror histogram (propensity matching) | dp | `mirror_histogram()` | [Mirror histogram](#dp-mirror) |
| Stacked histogram | dp | `stacked_histogram()` | [Stacked histogram](#dp-stacked) |

---

# Nonparametric temporal trends (`tp.np.*`) {#np}

All `tp.np.*` templates share a common SAS workflow:

1. Fit patient-specific temporal profiles with `%decompos()`.
2. Average profiles across patients with `PROC SUMMARY` → `mean_curv` dataset.
3. Optionally compute bootstrap confidence intervals → `boots_ci` dataset.
4. Compute binned patient-level summaries (deciles/quintiles) → `means` dataset.
5. Call the plotting template.

The R port replaces step 5. Steps 1–4 still run in SAS; export the resulting
datasets to CSV and read them into R:

```r
curve_data <- read.csv("mean_curv.csv")   # or boots_ci if CI bands are needed
data_pts   <- read.csv("means.csv")       # optional binned data points
```

## SAS column name mapping

| SAS column | R argument | Notes |
|---|---|---|
| `iv_echo` / `iv_wristm` / `iv_fyrs` | `x_col` | Time variable |
| `prev` / `mnprev` / `_p_` / `est` | `estimate_col` | Curve estimate |
| `cll_p68` / `cll_p95` | `lower_col` | CI lower band |
| `clu_p68` / `clu_p95` | `upper_col` | CI upper band |
| Grouping variable (e.g., `group`) | `group_col` | NULL for single curve |
| `mtime` / `mmtime` (means dataset) | `dp_x_col` | Binned point x |
| `mprev` / `mnprev` (means dataset) | `dp_y_col` | Binned point y |

## Average curve — binary outcome {#np-binary-avg}

**Ports:** `tp.np.afib.ivwristm.avrg_curv.binary.sas`,
`tp.np.afib.ivwristm.pt_specific.binary.sas`,
`tp.np.tr.icdpr.avg_curv.sas`

```{r np-binary-avg-example}
dat <- sample_nonparametric_curve_data(
  n            = 500,
  time_max     = 12,
  outcome_type = "probability"
)

# Minimal plot
nonparametric_curve_plot(dat$curve) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation Prevalence"
  ) +
  hvti_theme("manuscript")
```

Add 68% CI bands (one standard error; matches SAS `boots_ci` with `cll_p68`
/ `clu_p68`):

```{r np-binary-avg-ci}
nonparametric_curve_plot(
  dat$curve,
  lower_col = "lower",
  upper_col = "upper"
) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation — 68% CI"
  ) +
  hvti_theme("manuscript")
```

Add binned data summary points (matches the SAS `means` dataset):

```{r np-binary-avg-points}
nonparametric_curve_plot(
  dat$curve,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat$data_points
) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation with Binned Observations"
  ) +
  hvti_theme("manuscript")
```

## Continuous outcome {#np-continuous}

**Ports:** `tp.np.fev.double.univariate.continuous.sas`,
`tp.np.fev.u.trend.continuous.sas`

For continuous outcomes (FEV1, AV peak gradient) set `outcome_type =
"continuous"`. The y-axis label and scale change accordingly; everything
else is identical.

```{r np-continuous-example}
dat_cont <- sample_nonparametric_curve_data(
  n            = 400,
  time_max     = 10,
  outcome_type = "continuous"
)

nonparametric_curve_plot(
  dat_cont$curve,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_cont$data_points
) +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(
    x     = "Years after operation",
    y     = expression(FEV[1] ~ (L)),
    title = "FEV\u2081 Temporal Trend"
  ) +
  annotate("text",
    x = 9, y = 3.5,
    label = "68% CI band",
    hjust = 1, size = 3
  ) +
  hvti_theme("manuscript")
```

## Multi-group comparison {#np-multigroup}

**Ports:** `tp.np.afib.mult.avrg_curv.binary.sas`,
`tp.np.afib.mult.pt_spec.binary.sas`,
`tp.np.avpkgrad_ozak_ind_mtwt.sas`,
`tp.np.fev.multivariate.continuous.sas`

Provide a named vector to `groups` when generating sample data, or set
`group_col` when supplying your own curve data.

```{r np-multigroup-example}
dat_grp <- sample_nonparametric_curve_data(
  n        = 600,
  time_max = 12,
  groups   = c("Ozaki" = 0.7, "CE-Pericardial" = 1.1, "Homograft" = 1.4)
)

nonparametric_curve_plot(
  dat_grp$curve,
  group_col   = "group",
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_grp$data_points,
  dp_group_col = "group"
) +
  scale_colour_manual(
    values = c("Ozaki" = "#003087", "CE-Pericardial" = "#CC0000", "Homograft" = "#666666")
  ) +
  scale_fill_manual(
    values = c("Ozaki" = "#003087", "CE-Pericardial" = "#CC0000", "Homograft" = "#666666")
  ) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Prevalence (%)",
    colour = "Valve type",
    fill   = "Valve type",
    title  = "Atrial Fibrillation by Valve Type"
  ) +
  hvti_theme("manuscript")
```

## Phase decomposition {#np-phases}

**Ports:** `tp.np.afib.ivwristm.pt_spec_phases.binary.sas`,
`tp.np.tr.ivecho.u.phases.sas`

Phase plots separate the early (bell-shaped, incomplete hazard) and late
(Weibull CDF, complete hazard) components. Supply a `group_col` whose
levels label the phases.

```{r np-phases-example}
dat_phase <- sample_nonparametric_curve_data(
  n      = 500,
  groups = c("Early phase" = 1.5, "Late phase" = 0.6, "Overall" = 1.0)
)

nonparametric_curve_plot(
  dat_phase$curve,
  group_col = "group"
) +
  scale_colour_manual(
    values = c("Early phase" = "#CC0000",
               "Late phase"  = "#003087",
               "Overall"     = "black")
  ) +
  scale_linetype_manual(
    values = c("Early phase" = "dashed",
               "Late phase"  = "dashed",
               "Overall"     = "solid")
  ) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Prevalence (%)",
    colour = NULL,
    title  = "AF — Early and Late Phase Decomposition"
  ) +
  hvti_theme("manuscript")
```

## Covariate on the x-axis {#np-covariate-xaxis}

**Port:** `tp.np.z0axdpo.continuous.bmi_xaxis.sas`

When BMI or another continuous covariate (rather than time) is on the x-axis,
the function signature is identical — simply pass the covariate column name to
`x_col`.

```{r np-bmi-xaxis}
# Simulate covariate (BMI) x-axis data
set.seed(42)
n_pts  <- 300
bmi    <- seq(18, 45, length.out = n_pts)
est    <- plogis(-3 + 0.08 * bmi)
se     <- sqrt(est * (1 - est) / 50)
bmi_curve <- data.frame(
  bmi   = bmi,
  est   = est,
  lower = pmax(0, est - qnorm(0.84) * se),
  upper = pmin(1, est + qnorm(0.84) * se)
)

nonparametric_curve_plot(
  bmi_curve,
  x_col        = "bmi",
  estimate_col = "est",
  lower_col    = "lower",
  upper_col    = "upper"
) +
  scale_x_continuous(
    breaks = seq(18, 45, 3),
    limits = c(18, 45)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1),
    labels = scales::percent
  ) +
  labs(
    x     = expression(BMI ~ (kg/m^2)),
    y     = "Estimated Probability",
    title = "Outcome Probability vs. BMI at Operation"
  ) +
  hvti_theme("manuscript")
```

## Ordinal outcomes {#np-ordinal}

**Port:** `tp.np.tr.ivecho.average_curv.ordinal.sas`

Ordinal templates (TR grade, AR grade) use `nonparametric_ordinal_plot()`.
The critical migration step is reshaping the SAS wide-format `predict`
dataset (one column per grade) to long format.

**SAS reshape (do this before reading into R):**

```sas
/* SAS wide format: p0, p1, p2, p3 */
data predict_wide;
  set predict;
run;
```

**R reshape:**

```r
library(tidyr)
long <- pivot_longer(
  predict_wide,
  cols      = c(p0, p1, p2, p3),
  names_to  = "grade",
  values_to = "estimate"
)
```

```{r np-ordinal-example}
dat_ord <- sample_nonparametric_ordinal_data(
  n           = 1000,
  time_max    = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)

nonparametric_ordinal_plot(
  dat_ord$curve,
  grade_col   = "grade",
  data_points = dat_ord$data_points
) +
  scale_colour_manual(
    values = c(
      "None"     = "#003087",
      "Mild"     = "#55A51C",
      "Moderate" = "#FFA500",
      "Severe"   = "#CC0000"
    )
  ) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Grade probability",
    colour = "TR Grade",
    title  = "Tricuspid Regurgitation Grade"
  ) +
  hvti_theme("manuscript")
```

### Ordinal multi-scenario comparison {#np-ordinal-multi}

**Port:** `tp.np.po_ar.u_multi.ordinal.sas`

```{r np-ordinal-multi-example}
dat_ar1 <- sample_nonparametric_ordinal_data(seed = 1)
dat_ar2 <- sample_nonparametric_ordinal_data(seed = 99)

dat_ar1$curve$scenario <- "Before repair"
dat_ar2$curve$scenario <- "After repair"
combined <- rbind(dat_ar1$curve, dat_ar2$curve)

nonparametric_ordinal_plot(combined, grade_col = "grade") +
  facet_wrap(~scenario) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    x      = "Years",
    y      = "Grade probability",
    colour = "AR Grade",
    title  = "AR Grade — Before vs. After Repair"
  ) +
  hvti_theme("manuscript")
```

### Ordinal phase independence {#np-ordinal-independence}

**Port:** `tp.np.tr.ivecho.independence.sas`

To examine a single grade in isolation, filter the long-format curve
before passing it to the function:

```{r np-ordinal-independence-example}
dat_ind  <- sample_nonparametric_ordinal_data(n = 800)
grade_2  <- dat_ind$curve[dat_ind$curve$grade == "Grade 2", ]
dp_grade_2 <- dat_ind$data_points[dat_ind$data_points$grade == "Grade 2", ]

nonparametric_ordinal_plot(
  grade_2,
  grade_col   = "grade",
  data_points = dp_grade_2
) +
  scale_colour_manual(values = c("Grade 2" = "#CC0000")) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(
    limits = c(0, 0.5),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Probability",
    title = "Probability of TR Grade 2"
  ) +
  hvti_theme("manuscript")
```

### Ordinal phases {#np-ordinal-phases}

**Port:** `tp.np.tr.ivecho.u.phases.sas`

Phase-decomposed ordinal plots combine phase labels with grade colours.
Create the figure by plotting grade-specific curves and annotating early
vs. late phase regions:

```{r np-ordinal-phases-example}
dat_ph <- sample_nonparametric_ordinal_data(n = 800, seed = 7)

nonparametric_ordinal_plot(dat_ph$curve, grade_col = "grade") +
  annotate("rect",
    xmin = 0, xmax = 2,  ymin = -Inf, ymax = Inf,
    fill = "steelblue", alpha = 0.07
  ) +
  annotate("text",
    x = 1, y = 0.95, label = "Early\nphase",
    size = 3, colour = "steelblue", fontface = "italic"
  ) +
  annotate("rect",
    xmin = 2, xmax = 5, ymin = -Inf, ymax = Inf,
    fill = "tomato", alpha = 0.07
  ) +
  annotate("text",
    x = 3.5, y = 0.95, label = "Late\nphase",
    size = 3, colour = "tomato", fontface = "italic"
  ) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x      = "Years after operation",
    y      = "Grade probability",
    colour = "TR Grade",
    title  = "TR Grade — Early and Late Phase"
  ) +
  hvti_theme("manuscript")
```

---

# Survival analysis (`tp.ac.dead.*`, `tp.cp.dead.*`) {#ac-dead}

**Ports:** `tp.ac.dead.sas`, `tp.cp.dead.sas`

**R equivalent:** `survival_curve()`

`survival_curve()` wraps `survfit()` from the **survival** package and
returns a named list of ggplot objects matching the SAS `%kaplan` / `%nelsont`
macro output flags (`PLOTS`, `PLOTC`, `PLOTH`, `PLOTL`).

```r
dta    <- sample_survival_data(n = 500, seed = 42)
result <- survival_curve(dta)

# Kaplan–Meier survival (PLOTS)
result$survival_plot +
  scale_y_continuous(
    breaks = seq(0, 100, 20),
    labels = function(x) paste0(x, "%")
  ) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    x     = "Years after operation",
    y     = "Survival (%)",
    title = "Freedom from Death"
  ) +
  hvti_theme("manuscript")
```

### SAS → R argument mapping

| SAS `%kaplan` option | R argument | Notes |
|---|---|---|
| `data=` | `data` | Patient-level data frame |
| `time=` | `time_col` | Time-to-event column name |
| `event=` | `event_col` | Event indicator column (1 = event) |
| `group=` | `group_col` | Stratification variable |
| `method=kaplan` | `method = "kaplan-meier"` | Default |
| `method=nelsont` | `method = "nelson-aalen"` | Fleming–Harrington |
| `alpha=0.05` | `conf_int = 0.95` | 1 – alpha |
| `tp=0 1 2 3 5 7 10` | `report_times` | Table time points |

---

# Goodness of follow-up (`tp.dp.gfup.R`) {#dp-gfup}

**Port:** `tp.dp.gfup.R`

**R equivalent:** `goodness_followup()`

```r
dta <- sample_goodness_followup_data(n = 300)

goodness_followup(dta) +
  labs(title = "Goodness of Follow-Up") +
  hvti_theme("manuscript")
```

---

# Covariate balance (`tp.lp.propen.cov_balance.R`) {#lp-covbal}

**Port:** `tp.lp.propen.cov_balance.R`

**R equivalent:** `covariate_balance()`

```r
dta <- sample_covariate_balance_data(n_vars = 15)

covariate_balance(dta) +
  labs(title = "Standardised Mean Differences Before and After Matching") +
  hvti_theme("manuscript")
```

---

# Sankey / alluvial flow (`tp.dp.female_bicus_preAR_sankey.R`) {#dp-sankey}

**Port:** `tp.dp.female_bicus_preAR_sankey.R`

**R equivalent:** `sankey_plot()`

```r
dta <- sample_sankey_data(n = 200)

sankey_plot(dta, show_labels = TRUE) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Patient Flow Between States") +
  hvti_theme("manuscript")
```

---

# UpSet plot (`tp.complexUpset.R`) {#dp-upset}

**Port:** `tp.complexUpset.R`

**R equivalent:** `upset_plot()`

```r
sets <- c("CABG", "Valve", "MAZE", "Aorta")
dta  <- sample_upset_data(n = 300, sets = sets)

upset_plot(data = dta, intersect = sets)
```

---

# Spaghetti / individual trajectories (`tp.dp.spaghetti.echo.R`) {#dp-spaghetti}

**Port:** `tp.dp.spaghetti.echo.R`

**R equivalent:** `spaghetti_plot()`

```r
dta <- sample_spaghetti_data(n_patients = 60, n_visits = 5)

spaghetti_plot(dta) +
  scale_colour_viridis_c(option = "plasma") +
  labs(
    x      = "Years after operation",
    y      = "LVEF (%)",
    colour = "Age",
    title  = "Individual LV Ejection Fraction Trajectories"
  ) +
  hvti_theme("manuscript")
```

---

# Trends over time (`tp.dp.trends.R`) {#dp-trends}

**Port:** `tp.dp.trends.R`

**R equivalent:** `trends_plot()`

```r
dta <- sample_trends_data(n = 200)

trends_plot(dta) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    x     = "Operation year",
    y     = "Count",
    fill  = "Procedure",
    title = "Procedure Volume by Year"
  ) +
  hvti_theme("manuscript")
```

---

# Longitudinal patient counts (`tp.dp.longitudinal_patients_measures.R`) {#dp-long-counts}

**Port:** `tp.dp.longitudinal_patients_measures.R`

**R equivalent:** `longitudinal_counts_plot()`

```r
dta <- sample_longitudinal_counts_data(n_patients = 150)

longitudinal_counts_plot(dta) +
  scale_x_continuous(breaks = 0:10) +
  labs(
    x     = "Years after operation",
    y     = "Patients with measurement",
    title = "Follow-Up Echocardiogram Availability"
  ) +
  hvti_theme("manuscript")
```

---

# Mirror histogram (propensity score) {#dp-mirror}

**R equivalent:** `mirror_histogram()`

Used for propensity score overlap plots before and after matching or
weighting. No direct SAS template equivalent — this replaces ad-hoc SAS
histogram code common in propensity analyses.

```r
dta <- sample_mirror_histogram_data(n = 400)

mirror_histogram(dta) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(
    x     = "Propensity score",
    y     = "Count",
    title = "Propensity Score Distribution Before and After Matching"
  ) +
  hvti_theme("manuscript")
```

---

# Stacked histogram {#dp-stacked}

**R equivalent:** `stacked_histogram()`

```r
dta <- sample_stacked_histogram_data()

stacked_histogram(dta) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x     = "Operation year",
    y     = "Count",
    fill  = "Group",
    title = "Annual Case Volume by Group"
  ) +
  hvti_theme("manuscript")
```

---

# Using themes

All plot functions return an unstyled ggplot object. Add a theme as the
final layer:

| Context | Call | Equivalent SAS device |
|---|---|---|
| Dark PowerPoint slide | `hvti_theme("dark_ppt")` | `device=ppt` |
| Light PowerPoint slide | `hvti_theme("light_ppt")` | `device=ppt` with white background |
| Journal manuscript | `hvti_theme("manuscript")` | `device=eps` / PDF |
| Conference poster | `hvti_theme("poster")` | Large-font poster |

You can pass `base_size` to any theme to scale all text simultaneously:

```r
p + hvti_theme("manuscript", base_size = 10)   # smaller text for double-column
p + hvti_theme("poster",     base_size = 24)   # larger text for A0 poster
```

---

# Saving figures

## PowerPoint

```r
save_ppt(p, file = "figures/afib_prevalence.pptx")
```

## PDF / TIFF for journals

```r
ggsave("figures/afib_prevalence.pdf",  p, width = 3.5, height = 3.5, units = "in")
ggsave("figures/afib_prevalence.tiff", p, width = 3.5, height = 3.5,
       units = "in", dpi = 600)
```

---

# Adding new ports {#adding-new-ports}

As additional SAS templates are ported to R, update this guide:

1. **Add a row** to the lookup table at the top of this file with the SAS
   template name, family prefix, R function, and section anchor.

2. **Add a section** following the existing pattern: name the port, describe
   the SAS workflow, show the R equivalent with a runnable example, and
   document any column name mapping differences.

3. **Register the new R function** in `R/generics.R` under `hvti_plot()` if
   it is a new plot type.

4. **Update `DESCRIPTION`** if a new package dependency is required.

Templates currently planned for porting:

- `tp.hp.dead.*` hazard-function plots → `hazard_plot()` (pending)
- `tp.ce.states.*` competing-events / state-occupancy → (pending)
- `tp.gp.*` grouped longitudinal ordinal models → (pending)

---

# Session info

```{r session-info}
sessionInfo()
```
