---
title: "SAS Template Migration Guide: Finding Your Plot in R"
author: "John Ehrlinger <ehrlinj@ccf.org>"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
editor:
  markdown:
    wrap: 80
vignette: >
  %\VignetteIndexEntry{SAS Template Migration Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
# On Windows, R CMD check sets R_LIBS but Quarto subprocesses may not inherit
# the full library path. Explicitly add it before loading the package.
local({
  r_libs <- trimws(Sys.getenv("R_LIBS"))
  if (nzchar(r_libs)) {
    sep   <- if (.Platform$OS.type == "windows") ";" else ":"
    paths <- strsplit(r_libs, sep, fixed = TRUE)[[1]]
    .libPaths(unique(c(paths, .libPaths())))
  }
})
library(hvtiPlotR)
library(ggplot2)
```

# Overview

This guide maps every SAS template in the HVTI statistics group template
library to its R equivalent in the **hvtiPlotR** package. If you know the
SAS template name (e.g., `tp.np.afib.ivwristm.avrg_curv.binary.sas`), look it
up in the table below and jump to the corresponding section for a working R
example.

The guide is organized by template family (the two-letter prefix after `tp.`).
We add new ports as they become available.

## Key concepts for SAS users

Two concepts trip up most SAS users on first contact. First, R does not have a
`device=` / `GOPTIONS` call that sets the output context globally; you pick a
theme function and append it to each plot. Second, the constructor never writes
a file -- you have to call `print()` or assign the result to an object and save
separately. Everything else follows from these two facts.

**ggplot2 builds plots in layers.** Instead of one macro call with many
`color=`, `xaxis=`, and `footnote=` options, you chain `+` operations:

```r
np <- hv_nonparametric(dat, ...)
plot(np) +
  scale_colour_manual(values = c("black", "gray40")) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Years after Operation", y = "Prevalence (%)") +
  theme_hv_poster()
```

**The `theme_hv_*()` functions replace SAS `device=` / style options.** Use
`theme_hv_ppt_dark()` for PowerPoint slides on dark backgrounds,
`theme_hv_ppt_light()` for light/transparent slides,
`theme_hv_manuscript()` for journal figures, and `theme_hv_poster()` for
conference posters.

**Functions return a ggplot object; they do not display it.** Call the
plot object at the top level (or use `print()`) to render it, or save with
`ggsave()` / `save_ppt()`.

**Pre-fitted models, not raw data.** Most functions accept the *output*
of a statistical model (curve datasets, probability estimates) rather than
individual patient records — mirroring the SAS workflow where `%decompos()`
or `%kaplan` computes estimates and a separate template step produces the
figure.

---

# Template lookup table

| SAS Template | Family | R Constructor | Section |
|---|---|---|---|
| `tp.np.afib.ivwristm.avrg_curv.binary.sas` | np | `hv_nonparametric()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.afib.ivwristm.pt_spec_phases.binary.sas` | np | `hv_nonparametric()` | [Phase decomposition](#np-phases) |
| `tp.np.afib.ivwristm.pt_specific.binary.sas` | np | `hv_nonparametric()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.afib.mult.avrg_curv.binary.sas` | np | `hv_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.afib.mult.pt_spec.binary.sas` | np | `hv_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.avpkgrad_ozak_ind_mtwt.sas` | np | `hv_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.fev.double.univariate.continuous.sas` | np | `hv_nonparametric()` | [Continuous outcome](#np-continuous) |
| `tp.np.fev.multivariate.continuous.sas` | np | `hv_nonparametric()` | [Multi-group comparison](#np-multigroup) |
| `tp.np.fev.u.trend.continuous.sas` | np | `hv_nonparametric()` | [Continuous outcome](#np-continuous) |
| `tp.np.tr.icdpr.avg_curv.sas` | np | `hv_nonparametric()` | [Average curve — binary](#np-binary-avg) |
| `tp.np.tr.ivecho.average_curv.ordinal.sas` | np | `hv_ordinal()` | [Ordinal outcomes](#np-ordinal) |
| `tp.np.tr.ivecho.independence.sas` | np | `hv_ordinal()` | [Ordinal independence](#np-ordinal-independence) |
| `tp.np.tr.ivecho.u.phases.sas` | np | `hv_ordinal()` | [Ordinal phases](#np-ordinal-phases) |
| `tp.np.po_ar.u_multi.ordinal.sas` | np | `hv_ordinal()` | [Ordinal multi-scenario](#np-ordinal-multi) |
| `tp.np.z0axdpo.continuous.bmi_xaxis.sas` | np | `hv_nonparametric()` | [Covariate x-axis](#np-covariate-xaxis) |
| `tp.ac.dead.sas` (via `%kaplan` / `%nelsont`) | ac | `hv_survival()` | [Kaplan–Meier survival](#ac-dead) |
| `tp.cp.dead.sas` | cp | `hv_survival()` | [Kaplan–Meier survival](#ac-dead) |
| `tp.dp.gfup.R` | dp | `hv_followup()` | [Goodness of follow-up](#dp-gfup) |
| `tp.lp.propen.cov_balance.R` | lp | `hv_balance()` | [Covariate balance](#lp-covbal) |
| `tp.dp.female_bicus_preAR_sankey.R` | dp | `hv_alluvial()` | [Alluvial](#dp-sankey) |
| PAM cluster stability analysis | dp | `hv_sankey()` | [Cluster stability Sankey](#dp-cluster-sankey) |
| `tp.complexUpset.R` | dp | `hv_upset()` | [UpSet plot](#dp-upset) |
| `tp.dp.spaghetti.echo.R` | dp | `hv_spaghetti()` | [Spaghetti / individual trajectories](#dp-spaghetti) |
| `tp.rp.trends.sas` | rp | `hv_trends()` | [Trends over time](#dp-trends) |
| `tp.lp.trends.sas` | lp | `hv_trends()` | [Trends over time](#dp-trends) |
| `tp.lp.trends.age.sas` | lp | `hv_trends()` | [Trends over time](#dp-trends) |
| `tp.lp.trends.polytomous.sas` | lp | `hv_trends()` | [Trends over time](#dp-trends) |
| `tp.dp.trends.R` | dp | `hv_trends()` | [Trends over time](#dp-trends) |
| `tp.dp.longitudinal_patients_measures.R` | dp | `hv_longitudinal()` | [Longitudinal counts](#dp-long-counts) |
| `tp.lp.mirror-histogram_SAVR-TF-TAVR.R` | lp | `hv_mirror_hist()` | [Mirror histogram — binary-match](#dp-mirror) |
| `tp.lp.mirror_histo_before_after_wt.R` | lp | `hv_mirror_hist(weight_col = ...)` | [Mirror histogram — weighted IPTW](#dp-mirror) |
| Stacked histogram | dp | `hv_stacked()` | [Stacked histogram](#dp-stacked) |
| `tp.hs.dead.setup.sas` | hs | `hazard_plot()` | [Parametric hazard/survival](#hs-dead) |
| `tp.hs.dead_uses_setup.sas` | hs | `hazard_plot()` | [Parametric hazard/survival](#hs-dead) |
| `tp.hs.dead.procedure.tdepth.sas` | hs | `hazard_plot()` | [Parametric hazard/survival](#hs-dead) |
| `tp.hs.dead.conditional.setup.sas` | hs | `hazard_plot()` | [Conditional survival](#hs-dead) |
| `tp.hs.dead.conditional.uses_setup.sas` | hs | `hazard_plot()` | [Conditional survival](#hs-dead) |
| `tp.hs.dead.compare_benefit.setup.sas` | hs | `survival_difference_plot()` | [Treatment benefit distribution](#hs-dead) |
| `tp.hs.uslife_estimates_generate_stratify_.age.sas` | hs | `hazard_plot()` | [US life-table overlay](#hs-dead) |
| `tp.hs.uslife_generates_matched_estimates.sas` | hs | `hazard_plot()` | [US life-table overlay](#hs-dead) |

---

# Nonparametric temporal trends (`tp.np.*`) {#np}

All `tp.np.*` templates share a common SAS workflow:

1. Fit patient-specific temporal profiles with `%decompos()`.
2. Average profiles across patients with `PROC SUMMARY` → `mean_curv` dataset.
3. Optionally compute bootstrap confidence intervals → `boots_ci` dataset.
4. Compute binned patient-level summaries (deciles/quintiles) → `means` dataset.
5. Call the plotting template.

The R port replaces step 5. Steps 1–4 still run in SAS; export the resulting
datasets to CSV and read them into R:

```r
curve_data <- read.csv("mean_curv.csv")   # or boots_ci if CI bands are needed
data_pts   <- read.csv("means.csv")       # optional binned data points
```

## SAS column name mapping

| SAS column | R argument | Notes |
|---|---|---|
| `iv_echo` / `iv_wristm` / `iv_fyrs` | `x_col` | Time variable |
| `prev` / `mnprev` / `_p_` / `est` | `estimate_col` | Curve estimate |
| `cll_p68` / `cll_p95` | `lower_col` | CI lower band |
| `clu_p68` / `clu_p95` | `upper_col` | CI upper band |
| Grouping variable (e.g., `group`) | `group_col` | NULL for single curve |
| `mtime` / `mmtime` (means dataset) | `dp_x_col` | Binned point x |
| `mprev` / `mnprev` (means dataset) | `dp_y_col` | Binned point y |

## Average curve — binary outcome {#np-binary-avg}

**Ports:** `tp.np.afib.ivwristm.avrg_curv.binary.sas`,
`tp.np.afib.ivwristm.pt_specific.binary.sas`,
`tp.np.tr.icdpr.avg_curv.sas`

```{r np-binary-avg-example}
dat     <- sample_nonparametric_curve_data(
  n            = 500,
  time_max     = 12,
  outcome_type = "probability"
)
dat_pts <- sample_nonparametric_curve_points(n = 500, time_max = 12)

# Minimal plot
plot(hv_nonparametric(dat)) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation Prevalence"
  ) +
  theme_hv_poster()
```

Add 68% CI bands (one standard error; matches SAS `boots_ci` with `cll_p68`
/ `clu_p68`):

```{r np-binary-avg-ci}
plot(hv_nonparametric(
  dat,
  lower_col = "lower",
  upper_col = "upper"
)) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation — 68% CI"
  ) +
  theme_hv_poster()
```

Add binned data summary points (matches the SAS `means` dataset):

```{r np-binary-avg-points}
plot(hv_nonparametric(
  dat,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_pts
)) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Prevalence (%)",
    title = "Atrial Fibrillation with Binned Observations"
  ) +
  theme_hv_poster()
```

## Continuous outcome {#np-continuous}

**Ports:** `tp.np.fev.double.univariate.continuous.sas`,
`tp.np.fev.u.trend.continuous.sas`

For continuous outcomes (FEV1, AV peak gradient) set `outcome_type =
"continuous"`. The y-axis label and scale change accordingly; everything
else is identical.

```{r np-continuous-example}
dat_cont     <- sample_nonparametric_curve_data(
  n            = 400,
  time_max     = 10,
  outcome_type = "continuous"
)
dat_cont_pts <- sample_nonparametric_curve_points(
  n            = 400,
  time_max     = 10,
  outcome_type = "continuous"
)

plot(hv_nonparametric(
  dat_cont,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_cont_pts
)) +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  scale_y_continuous(limits = c(0, 4)) +
  labs(
    x     = "Years after operation",
    y     = expression(FEV[1] ~ (L)),
    title = "FEV\u2081 Temporal Trend"
  ) +
  annotate("text",
    x = 9, y = 3.5,
    label = "68% CI band",
    hjust = 1, size = 3
  ) +
  theme_hv_poster()
```

## Multi-group comparison {#np-multigroup}

**Ports:** `tp.np.afib.mult.avrg_curv.binary.sas`,
`tp.np.afib.mult.pt_spec.binary.sas`,
`tp.np.avpkgrad_ozak_ind_mtwt.sas`,
`tp.np.fev.multivariate.continuous.sas`

Provide a named vector to `groups` when generating sample data, or set
`group_col` when supplying your own curve data.

```{r np-multigroup-example}
grp_def <- c("Ozaki" = 0.7, "CE-Pericardial" = 1.1, "Homograft" = 1.4)
dat_grp     <- sample_nonparametric_curve_data(
  n        = 600,
  time_max = 12,
  groups   = grp_def
)
dat_grp_pts <- sample_nonparametric_curve_points(
  n        = 600,
  time_max = 12,
  groups   = grp_def
)

plot(hv_nonparametric(
  dat_grp,
  group_col   = "group",
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = dat_grp_pts
)) +
  scale_colour_manual(
    values = c("Ozaki" = "#003087", "CE-Pericardial" = "#CC0000", "Homograft" = "#666666")
  ) +
  scale_fill_manual(
    values = c("Ozaki" = "#003087", "CE-Pericardial" = "#CC0000", "Homograft" = "#666666")
  ) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Prevalence (%)",
    colour = "Valve type",
    fill   = "Valve type",
    title  = "Atrial Fibrillation by Valve Type"
  ) +
  theme_hv_poster()
```

## Phase decomposition {#np-phases}

**Ports:** `tp.np.afib.ivwristm.pt_spec_phases.binary.sas`,
`tp.np.tr.ivecho.u.phases.sas`

Phase plots separate the early (bell-shaped, incomplete hazard) and late
(Weibull CDF, complete hazard) components. Supply a `group_col` whose
levels label the phases.

```{r np-phases-example}
dat_phase <- sample_nonparametric_curve_data(
  n      = 500,
  groups = c("Early phase" = 1.5, "Late phase" = 0.6, "Overall" = 1.0)
)

plot(hv_nonparametric(
  dat_phase,
  group_col = "group"
)) +
  scale_colour_manual(
    values = c("Early phase" = "#CC0000",
               "Late phase"  = "#003087",
               "Overall"     = "black")
  ) +
  scale_linetype_manual(
    values = c("Early phase" = "dashed",
               "Late phase"  = "dashed",
               "Overall"     = "solid")
  ) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Prevalence (%)",
    colour = NULL,
    title  = "AF — Early and Late Phase Decomposition"
  ) +
  theme_hv_poster()
```

## Covariate on the x-axis {#np-covariate-xaxis}

**Port:** `tp.np.z0axdpo.continuous.bmi_xaxis.sas`

When BMI or another continuous covariate (rather than time) is on the x-axis,
the function signature is identical — simply pass the covariate column name to
`x_col`.

```{r np-bmi-xaxis}
# Simulate covariate (BMI) x-axis data
set.seed(42)
n_pts  <- 300
bmi    <- seq(18, 45, length.out = n_pts)
est    <- plogis(-3 + 0.08 * bmi)
se     <- sqrt(est * (1 - est) / 50)
bmi_curve <- data.frame(
  bmi   = bmi,
  est   = est,
  lower = pmax(0, est - qnorm(0.84) * se),
  upper = pmin(1, est + qnorm(0.84) * se)
)

plot(hv_nonparametric(
  bmi_curve,
  x_col        = "bmi",
  estimate_col = "est",
  lower_col    = "lower",
  upper_col    = "upper"
)) +
  scale_x_continuous(
    breaks = seq(18, 45, 3),
    limits = c(18, 45)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1),
    labels = scales::percent
  ) +
  labs(
    x     = expression(BMI ~ (kg/m^2)),
    y     = "Estimated Probability",
    title = "Outcome Probability vs. BMI at Operation"
  ) +
  theme_hv_poster()
```

## Ordinal outcomes {#np-ordinal}

**Port:** `tp.np.tr.ivecho.average_curv.ordinal.sas`

Ordinal templates (TR grade, AR grade) use `hv_ordinal()`.
The main migration step is reshaping the SAS wide-format `predict`
dataset (one column per grade) to long format.

**SAS reshape (do this before reading into R):**

```sas
/* SAS wide format: p0, p1, p2, p3 */
data predict_wide;
  set predict;
run;
```

**R reshape:**

```r
library(tidyr)
long <- pivot_longer(
  predict_wide,
  cols      = c(p0, p1, p2, p3),
  names_to  = "grade",
  values_to = "estimate"
)
```

```{r np-ordinal-example}
ord_labels <- c("None", "Mild", "Moderate", "Severe")
dat_ord     <- sample_nonparametric_ordinal_data(
  n            = 1000,
  time_max     = 5,
  grade_labels = ord_labels
)
dat_ord_pts <- sample_nonparametric_ordinal_points(
  n            = 1000,
  time_max     = 5,
  grade_labels = ord_labels
)

plot(hv_ordinal(
  dat_ord,
  grade_col   = "grade",
  data_points = dat_ord_pts
)) +
  scale_colour_manual(
    values = c(
      "None"     = "#003087",
      "Mild"     = "#55A51C",
      "Moderate" = "#FFA500",
      "Severe"   = "#CC0000"
    )
  ) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent
  ) +
  labs(
    x      = "Years after operation",
    y      = "Grade probability",
    colour = "TR Grade",
    title  = "Tricuspid Regurgitation Grade"
  ) +
  theme_hv_poster()
```

### Ordinal multi-scenario comparison {#np-ordinal-multi}

**Port:** `tp.np.po_ar.u_multi.ordinal.sas`

The multi-scenario template (`po_ar` = post-op aortic regurgitation) places
two or more ordinal grade curves side by side so you can read the grade
distribution before repair in one panel and after repair in the next. In SAS
this required separate `%ordinal` calls and manual page layout; here you bind
the datasets, add a `scenario` column, and let `facet_wrap()` do the layout.

```{r np-ordinal-multi-example}
dat_ar1 <- sample_nonparametric_ordinal_data(seed = 1)
dat_ar2 <- sample_nonparametric_ordinal_data(seed = 99)

dat_ar1$scenario <- "Before repair"
dat_ar2$scenario <- "After repair"
combined <- rbind(dat_ar1, dat_ar2)

plot(hv_ordinal(combined, grade_col = "grade")) +
  facet_wrap(~scenario) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(
    x      = "Years",
    y      = "Grade probability",
    colour = "AR Grade",
    title  = "AR Grade — Before vs. After Repair"
  ) +
  theme_hv_poster()
```

### Ordinal phase independence {#np-ordinal-independence}

**Port:** `tp.np.tr.ivecho.independence.sas`

To examine a single grade in isolation, filter the long-format curve
before passing it to the function:

```{r np-ordinal-independence-example}
dat_ind    <- sample_nonparametric_ordinal_data(n = 800)
pts_ind    <- sample_nonparametric_ordinal_points(n = 800)
grade_2    <- dat_ind[dat_ind$grade == "Grade 2", ]
dp_grade_2 <- pts_ind[pts_ind$grade == "Grade 2", ]

plot(hv_ordinal(
  grade_2,
  grade_col   = "grade",
  data_points = dp_grade_2
)) +
  scale_colour_manual(values = c("Grade 2" = "#CC0000")) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(
    limits = c(0, 0.5),
    labels = scales::percent
  ) +
  labs(
    x     = "Years after operation",
    y     = "Probability",
    title = "Probability of TR Grade 2"
  ) +
  theme_hv_poster()
```

### Ordinal phases {#np-ordinal-phases}

**Port:** `tp.np.tr.ivecho.u.phases.sas`

Phase-decomposed ordinal plots combine phase labels with grade colours.
Create the figure by plotting grade-specific curves and annotating early
vs. late phase regions:

```{r np-ordinal-phases-example}
dat_ph <- sample_nonparametric_ordinal_data(n = 800, seed = 7)

plot(hv_ordinal(dat_ph, grade_col = "grade")) +
  annotate("rect",
    xmin = 0, xmax = 2,  ymin = -Inf, ymax = Inf,
    fill = "steelblue", alpha = 0.07
  ) +
  annotate("text",
    x = 1, y = 0.95, label = "Early\nphase",
    size = 3, colour = "steelblue", fontface = "italic"
  ) +
  annotate("rect",
    xmin = 2, xmax = 5, ymin = -Inf, ymax = Inf,
    fill = "tomato", alpha = 0.07
  ) +
  annotate("text",
    x = 3.5, y = 0.95, label = "Late\nphase",
    size = 3, colour = "tomato", fontface = "italic"
  ) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x      = "Years after operation",
    y      = "Grade probability",
    colour = "TR Grade",
    title  = "TR Grade — Early and Late Phase"
  ) +
  theme_hv_poster()
```

---

# Survival analysis (`tp.ac.dead.*`, `tp.cp.dead.*`) {#ac-dead}

**Ports:** `tp.ac.dead.sas`, `tp.cp.dead.sas`

**R equivalent:** `hv_survival()`

`hv_survival()` wraps `survfit()` from the **survival** package and returns
an S3 object. Call `plot(km, type = ...)` to render one of the five plot types
matching the SAS `%kaplan` / `%nelsont` macro output flags (`PLOTS`, `PLOTC`,
`PLOTH`, `PLOTL`). Tidy data frames live in `km$tables`.

```r
dta <- sample_survival_data(n = 500, seed = 42)
km  <- hv_survival(dta)

# Kaplan–Meier survival (PLOTS)
plot(km) +
  scale_y_continuous(
    breaks = seq(0, 100, 20),
    labels = function(x) paste0(x, "%")
  ) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    x     = "Years after operation",
    y     = "Survival (%)",
    title = "Freedom from Death"
  ) +
  theme_hv_poster()

# Cumulative hazard (PLOTC)
plot(km, type = "cumhaz") +
  labs(x = "Years", y = "Cumulative Hazard") +
  theme_hv_poster()

# Hazard rate (PLOTH)
plot(km, type = "hazard") +
  labs(x = "Years", y = "Instantaneous Hazard") +
  theme_hv_poster()

# Log-log survival (PH check)
plot(km, type = "loglog") +
  labs(x = "log(Years)", y = "log(-log S(t))") +
  theme_hv_poster()

# Integrated survivorship (PLOTL)
plot(km, type = "life") +
  labs(x = "Years", y = "Restricted Mean Survival (years)") +
  theme_hv_poster()

# Access the KM data, risk table, and report table
km$data              # tidy KM data frame
km$tables$risk       # numbers-at-risk table
km$tables$report     # survival at report times
```

### SAS → R argument mapping

| SAS `%kaplan` option | R argument | Notes |
|---|---|---|
| `data=` | `data` | Patient-level data frame |
| `time=` | `time_col` | Time-to-event column name |
| `event=` | `event_col` | Event indicator column (1 = event) |
| `group=` | `group_col` | Stratification variable |
| `method=kaplan` | `method = "kaplan-meier"` | Default |
| `method=nelsont` | `method = "nelson-aalen"` | Fleming–Harrington |
| `alpha=0.05` | `conf_level = 0.95` | Default is 0.95 |
| `tp=0 1 2 3 5 7 10` | `report_times` | Table time points |

---

# Goodness of follow-up (`tp.dp.gfup.R`) {#dp-gfup}

**Port:** `tp.dp.gfup.R`

**R equivalent:** `hv_followup()`

`tp.dp.gfup.R` summarises how many patients remain under active follow-up at
each time point -- the kind of quality-check figure you run before committing
to a temporal-prevalence analysis. The SAS version uses `PROC FREQ` output
aggregated by year; `hv_followup()` accepts the same aggregated data frame
(one row per time point, with counts for total patients and patients with a
measurement). The output is a bar chart; any goodness-of-follow-up threshold
you add in SAS (a dashed horizontal line) becomes a
`geom_hline(yintercept = ..., linetype = "dashed")` call in R.

```r
dta <- sample_goodness_followup_data(n = 300)
gf  <- hv_followup(dta)

plot(gf) +
  labs(title = "Goodness of Follow-Up") +
  theme_hv_poster()
```

---

# Covariate balance (`tp.lp.propen.cov_balance.R`) {#lp-covbal}

**Port:** `tp.lp.propen.cov_balance.R`

**R equivalent:** `hv_balance()`

The SAS export arrives wide (one column per time-point). Reshape to long format
before you call `hv_balance()`. Pass `var_levels` to control the
bottom-to-top display order of covariates — this matches `ylabel` in the
original script.

```r
# Reshape wide → long (mirrors the original script)
ylabel <- c(
  "Cardiac Output", "BAV", "Cardiac Index", "NYHA -Functional Class",
  "COPD-Oxy", "Date of Surg", "Creatinine", "Age", "CAD",
  "Female", "LV PWT", "Mismatch", "HTN", "PVD", "Sinitibular Jcn: diam"
)

names(dta) <- c("variable", "Before match", "After match")

dta_long <- reshape(
  dta,
  direction = "long",
  varying   = c("Before match", "After match"),
  v.names   = "std_diff",
  timevar   = "group",
  times     = c("Before match", "After match"),
  idvar     = "variable"
)

n_vars    <- length(ylabel)
cb        <- hv_balance(
  data         = dta_long,
  variable_col = "variable",
  group_col    = "group",
  std_diff_col = "std_diff",
  var_levels   = ylabel
)

stdifPlot <- plot(cb) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(-40, 30), breaks = seq(-40, 30, 10)) +
  labs(x = "Standardized difference: SAVR - TF:TAVR (%)", y = "") +
  annotate("text", x = -30, y = 0,      label = "More likely TF-TAVR", size = 4.5) +
  annotate("text", x =  22, y = n_vars, label = "More likely SAVR",    size = 4.5) +
  theme(legend.position = c(0.20, 0.935)) +
  theme_hv_poster()

ggsave(here::here("graphs", "lp_cov-balance-SAVR_TF-TAVR.pdf"),
       plot = stdifPlot, height = 7, width = 8)
```

---

# Alluvial flow (`tp.dp.female_bicus_preAR_sankey.R`) {#dp-sankey}

**Port:** `tp.dp.female_bicus_preAR_sankey.R`

**R equivalent:** `hv_alluvial()`

`tp.dp.female_bicus_preAR_sankey.R` traces each patient through a sequence of
categorical states (e.g. pre-op AR grade to procedure type to sex subgroup),
producing the flowing band diagram that the original script builds with
`ggalluvial`. The R port wraps this with `hv_alluvial()`, which handles the
internal `to_lodes_form()` reshape and sets default fill aesthetics. You supply
a data frame with one column per state and let `scale_fill_brewer()` or a
manual fill scale control the colours. See the
[Alluvial section of the plot-functions vignette](plot-functions.html#dp-sankey)
for a decorated worked example with labels.

```r
dta <- sample_alluvial_data(n = 200)
al  <- hv_alluvial(dta)

plot(al) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Patient Flow Between States") +
  theme_hv_poster()
```

---

# Cluster stability Sankey (PAM analysis) {#dp-cluster-sankey}

**Source:** PAM clustering analysis pipeline (R script using `ggsankey`)

**R equivalent:** `hv_sankey()`

`hv_sankey()` draws a Sankey diagram showing how patients flow between
letter-labelled clusters as the number of clusters K increases from 2 to 9.
Each column is one K; flow bands show assignment changes between consecutive K
values; node labels show cluster letter + patient count.

The original code used `ggsankey::make_long()` with hard-coded column names and
ordering vectors. The R port accepts any cluster columns and ordering via
`cluster_cols` and `node_levels` arguments.

**Requires `ggsankey`** (install from GitHub):

```r
remotes::install_github("davidsjoberg/ggsankey")
```

### SAS/R workflow → R equivalent

| Original step | R equivalent |
|---|---|
| `sid_dta$C2 <- factor(...)` with `gr2_names` ordering | Factor levels set by `node_levels` argument |
| `make_long(C2, …, C9)` | `.make_sankey_long()` (internal) |
| `geom_sankey()` + `geom_sankey_label()` | `hv_sankey()` + `plot()` |
| `brewer.pal(9, "Set1")[c(2,6,8,4,3,5,7,1,9)]` | Default `node_colours` |

```r
dta_san <- sample_cluster_sankey_data(n = 300, seed = 42)
sk      <- hv_sankey(dta_san)

plot(sk) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  theme_hv_poster()
```

To reproduce the original analysis with patient-level cluster assignments:

```r
# Build a data frame with one row per patient, columns C2..C9 (factor)
grp_dta <- data.frame(
  C2 = factor(pm2clusters_labels, levels = gr2_names),
  C3 = factor(pm3clusters_labels, levels = gr3_names),
  # ... etc.
  C9 = factor(pm9clusters_labels, levels = gr9_names)
)

sk_grp <- hv_sankey(
  grp_dta,
  cluster_cols = paste0("C", 2:9),
  node_levels  = gr9_names
)

plot(sk_grp) +
  labs(x = NULL) +
  theme_hv_poster()
```

---

# UpSet plot (`tp.complexUpset.R`) {#dp-upset}

**Port:** `tp.complexUpset.R`

**R equivalent:** `hv_upset()`

`tp.complexUpset.R` visualises how many patients fall into each combination of
procedure categories (CABG, Valve, MAZE, Aorta, etc.) -- the combinatorial
overlap question that Venn diagrams can't answer cleanly past three sets. The
SAS version used a custom `PROC TABULATE` / `SGPLOT` workaround; the R port
replaces it with the `ComplexUpset` package wrapped in `hv_upset()`. You pass
a binary-indicator data frame (one column per set, 1 = member) and the
`intersect` argument names the sets to include.

```r
sets <- c("CABG", "Valve", "MAZE", "Aorta")
dta  <- sample_upset_data(n = 300, sets = sets)
hu   <- hv_upset(data = dta, intersect = sets)

plot(hu)
```

---

# Spaghetti / individual trajectories (`tp.dp.spaghetti.echo.R`) {#dp-spaghetti}

**Port:** `tp.dp.spaghetti.echo.R`

**R equivalent:** `hv_spaghetti()`

The template covers nine figures across three echo outcomes (AV mean gradient,
AV area, DVI) in unstratified and sex-stratified variants, plus an ordinal MV
regurgitation grade plot (plot_9).

```r
dta    <- sample_spaghetti_data(n_patients = 150, max_obs = 6)
sp     <- hv_spaghetti(dta)
sp_col <- hv_spaghetti(dta, colour_col = "group")
```

## Unstratified — AV mean gradient full range (plot_1)

The SAS template `tp.dp.spaghetti.echo.R` sets `AXISY ORDER=(0 TO 80 BY 20)`
for the full gradient range. Reproduce that scale with `coord_cartesian()` so
you keep the ggplot clipping behaviour rather than silently dropping
out-of-range trajectories.

```r
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_poster()
```

## Unstratified — zoomed y-axis (plot_3)

Plot_3 in the SAS script tightens the y-axis to `ORDER=(0 TO 30 BY 10)` to
reveal structure in the low-gradient patients that is invisible at the 0-80
range. Only `coord_cartesian()` changes; the constructor call and data are
identical to plot_1. Look for: trajectories that were flat in plot_1 now
showing meaningful within-patient variation -- if none appear, the patient
population may not have a low-gradient subgroup.

```r
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 30, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 30)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_poster()
```

## Stratified by sex (plot_2 / plot_4)

Template uses `values=c("red", "blue")`; modernised equivalents below. Pass
`colour_col = "group"` to `hv_spaghetti()` when constructing `sp_col` so each
patient trajectory inherits the group colour.

```r
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 80, 20)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 80)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_poster()
```

## AV area y-scale (plot_5 / plot_6)

Plot_5 (unstratified) and plot_6 (by sex) switch the outcome to effective
orifice area (EOA, cm^2^) with `AXISY ORDER=(0 TO 5 BY 1)`. The only
change from the gradient plots is the y-axis range and the label; the
constructor object and colour scale reuse `sp` / `sp_col` from above.

```r
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 5, 1)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(x = "Years", y = "AV Area (EOA) (cm\u00b2)") +
  theme_hv_poster()
```

## DVI y-scale (plot_7 / plot_8)

Plot_7 (unstratified) and plot_8 (by sex) track the dimensionless velocity
index (DVI), a unitless ratio bounded roughly 0-1.25. The SAS option is
`AXISY ORDER=(0 TO 1.25 BY 0.25)`. This is the only spaghetti variant where
a y-axis maximum above 1 is correct -- do not treat values above 1 as
outliers.

```r
plot(sp_col) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 1.25, 0.25)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 1.25)) +
  labs(x = "Years", y = "DVI") +
  theme_hv_poster()
```

## Ordinal y-axis — MV regurgitation grade (plot_9)

Plot_9 is the outlier in this template family: MV regurgitation grade is
ordinal (None / Mild / Moderate / Severe = 0-3), not continuous. In SAS the
y-axis was labelled with `AXISY ORDER=(0 TO 3 BY 1)` and annotated manually;
here you pass `y_labels` to `plot()` to replace the numeric tick marks with
grade names. The pre-processing step rounds the raw echo value to an integer
0-3 before constructing the spaghetti object.

```r
dta_ord       <- dta
dta_ord$value <- round(pmin(3, pmax(0, dta$value / 12)))
levels(dta_ord$group) <- c("Early", "Late")

sp_ord <- hv_spaghetti(dta_ord, colour_col = "group")

plot(sp_ord, y_labels = c(None = 0, Mild = 1, Moderate = 2, Severe = 3)) +
  scale_colour_manual(
    values = c(Early = "steelblue", Late = "red2"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 6, 1)) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 3)) +
  labs(x = "Years after Procedure", y = "MV Regurgitation Grade") +
  theme_hv_poster()
```

---

# Trends over time (`tp.*.trends.*`) {#dp-trends}

**Ports:** `tp.rp.trends.sas`, `tp.lp.trends.sas`, `tp.lp.trends.age.sas`,
`tp.lp.trends.polytomous.sas`, `tp.dp.trends.R`

**R equivalent:** `hv_trends()`

## tp.rp.trends.sas — cases/year and age (1968–2000 by 4)

`tp.rp.trends.sas` is the right-panel trends template: a single smoothed
curve for cases-per-year and a second figure for median operative age, both
spanning 1968-2000. In SAS the two figures are separate `SGPLOT` calls sharing
the same `axisx order=(1968 to 2000 by 4)` statement. In R you call the same
`hv_trends()` object twice with different y-axis scales -- one for counts, one
for age -- so you only need to build the constructor once.

```r
one    <- sample_trends_data(n = 600, year_range = c(1968L, 2000L), groups = NULL)
tr_one <- hv_trends(one, group_col = NULL)

# Cases/year: axisy order=(0 to 10 by 2)
plot(tr_one) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 10),      breaks = seq(0, 10, 2)) +
  labs(x = "Year", y = "Cases/year") +
  theme_hv_poster()

# Age: axisy order=(30 to 70 by 10)
plot(tr_one) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(30, 70),     breaks = seq(30, 70, 10)) +
  labs(x = "Year", y = "Age (years)") +
  theme_hv_poster()
```

## tp.lp.trends.sas — binary % outcomes (1970–2000 by 10, y 0–100 by 10)

`tp.lp.trends.sas` is the left-panel trends template for binary percentage
outcomes plotted together on a single figure -- shock rate, pre-op IABP use,
inotrope use, etc. The SAS `SGPLOT` overlays multiple `SCATTER` / `REG`
statement pairs, one per outcome, with `axisx order=(1970 to 2000 by 10)` and
`axisy order=(0 to 100 by 20)`. In R you encode the outcome identity as a
group variable and use `scale_colour_manual()` / `scale_shape_manual()` to
assign the same per-outcome styling. CGM axis spec: `axisx order=(1970 to 2000
by 10)`, `axisy order=(0 to 100 by 20)`.

```r
dta_lp <- sample_trends_data(
  n = 800, year_range = c(1970L, 2000L),
  groups = c("Shock %", "Pre-op IABP %", "Inotropes %"))
tr_lp <- hv_trends(dta_lp)

plot(tr_lp) +
  scale_colour_manual(
    values = c("Shock %" = "steelblue", "Pre-op IABP %" = "firebrick",
               "Inotropes %" = "forestgreen"), name = NULL) +
  scale_shape_manual(
    values = c("Shock %" = 16L, "Pre-op IABP %" = 15L, "Inotropes %" = 17L),
    name = NULL) +
  scale_x_continuous(limits = c(1970, 2000), breaks = seq(1970, 2000, 10)) +
  scale_y_continuous(limits = c(0, 100),     breaks = seq(0, 100, 10)) +
  coord_cartesian(xlim = c(1970, 2000), ylim = c(0, 100)) +
  labs(x = "Year", y = "Percent (%)") +
  theme_hv_poster()
```

## tp.lp.trends.age.sas — age on x-axis (25–85 by 10, y 0–100 by 20)

`tp.lp.trends.age.sas` flips the x-axis from calendar year to patient age at
operation, showing how the proportion receiving each procedure type shifts with
age. The SAS `axisx order=(25 to 85 by 10)` statement is the tell; in R you
use the same `hv_trends()` constructor but pass age as the x variable and set
`scale_x_continuous(limits = c(25, 85), breaks = seq(25, 85, 10))`.

```r
dta_age <- sample_trends_data(
  n = 600, year_range = c(25L, 85L),
  groups = c("Repair %", "Bioprosthesis %"), seed = 7L)
tr_age <- hv_trends(dta_age)

plot(tr_age) +
  scale_colour_manual(
    values = c("Repair %" = "steelblue", "Bioprosthesis %" = "firebrick"),
    name = NULL) +
  scale_x_continuous(limits = c(25, 85), breaks = seq(25, 85, 10)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  coord_cartesian(xlim = c(25, 85), ylim = c(0, 100)) +
  labs(x = "Age (years)", y = "Percent (%)") +
  theme_hv_poster()
```

## tp.lp.trends.polytomous.sas — repair types (1990–1999 by 1, y 0–100 by 10)

`tp.lp.trends.polytomous.sas` extends the binary trends template to three or
more mutually exclusive categorical outcomes (CE, Cosgrove, Periguard, DeVega
repair types) that together sum to 100%. The SAS template uses a separate
`SCATTER` / `REG` overlay per repair type; here each type is a group level.
Because the percents are compositional (they sum to roughly 100), use
`coord_cartesian()` rather than `scale_y_continuous(limits = ...)` to avoid
silently dropping early time points where sample sizes are small.

```r
dta_poly <- sample_trends_data(
  n = 800, year_range = c(1990L, 1999L),
  groups = c("CE", "Cosgrove", "Periguard", "DeVega"), seed = 5L)
tr_poly <- hv_trends(dta_poly)

plot(tr_poly) +
  scale_colour_manual(
    values = c(CE = "steelblue", Cosgrove = "firebrick",
               Periguard = "forestgreen", DeVega = "goldenrod3"),
    name = "Repair type") +
  scale_shape_manual(
    values = c(CE = 15L, Cosgrove = 19L, Periguard = 17L, DeVega = 18L),
    name = "Repair type") +
  scale_x_continuous(limits = c(1990, 1999), breaks = seq(1990, 1999, 1)) +
  scale_y_continuous(limits = c(0, 100),     breaks = seq(0, 100, 10)) +
  coord_cartesian(xlim = c(1990, 1999), ylim = c(0, 100)) +
  labs(x = "Year", y = "Percent (%)") +
  theme_hv_poster()
```

## tp.dp.trends.R — LV mass index (1995–2015 by 5, y 0–200 by 50)

`tp.dp.trends.R` covers continuous-outcome trends, here LV mass index (g/m^2^)
over a 20-year window. Unlike the SAS templates above this is an R script
origin rather than a `.sas` file, so there is no `axisx` option to translate
literally -- the axis spec (`1995 to 2015 by 5`, y `0 to 200 by 50`) comes
from the `scale_*_continuous()` calls below.

```r
dta_lv <- sample_trends_data(n = 800, year_range = c(1995L, 2015L),
                              groups = NULL, seed = 3L)
tr_lv  <- hv_trends(dta_lv, group_col = NULL)

plot(tr_lv) +
  scale_x_continuous(limits = c(1995, 2015), breaks = seq(1995, 2015, 5)) +
  scale_y_continuous(limits = c(0, 200),     breaks = seq(0, 200, 50)) +
  coord_cartesian(xlim = c(1995, 2015), ylim = c(0, 200)) +
  labs(x = "Years", y = "LV Mass Index") +
  theme_hv_poster()
```

## tp.dp.trends.R — hospital LOS with annotation (1985–2015 by 5, y 0–20 by 5)

This variant of `tp.dp.trends.R` tracks hospital length of stay (days) and
adds a text annotation inside the plot panel -- the kind of label the SAS
`SGPLOT` `INSET` statement would place. In R an `annotate("text", ...)` call
places it at absolute data coordinates; adjust the `x` and `y` values to
avoid overlapping the fitted curve.

```r
dta_los <- sample_trends_data(n = 800, year_range = c(1985L, 2015L),
                               groups = NULL, seed = 11L)
tr_los  <- hv_trends(dta_los, group_col = NULL)

plot(tr_los) +
  scale_x_continuous(limits = c(1985, 2015), breaks = seq(1985, 2015, 5)) +
  scale_y_continuous(limits = c(0, 20),      breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(1985, 2015), ylim = c(0, 20)) +
  annotate("text", x = 1995, y = 18,
           label = "Trend: Hospital Length of Stay", size = 4.5) +
  labs(x = "Years", y = "Hospital LOS (Days)") +
  theme_hv_poster()
```

---

# Longitudinal patient counts (`tp.dp.longitudinal_patients_measures.R`) {#dp-long-counts}

**Port:** `tp.dp.longitudinal_patients_measures.R`

**R equivalent:** `hv_longitudinal()`

`tp.dp.longitudinal_patients_measures.R` answers the question "how many
patients have an echocardiogram measurement at each follow-up year?" -- a
companion to the goodness-of-follow-up chart above, but focused on measurement
availability rather than vital-status follow-up. The original R script used
`PROC FREQ`-style counting over a long-format echo dataset; `hv_longitudinal()`
accepts the same patient-level long-format frame and produces a bar chart with
count annotations. You need one row per patient per observation, with columns
for patient ID, follow-up year, and (optionally) a measurement flag.

```r
dta <- sample_longitudinal_counts_data(n_patients = 150)
lc  <- hv_longitudinal(dta)

plot(lc) +
  scale_x_continuous(breaks = 0:10) +
  labs(
    x     = "Years after operation",
    y     = "Patients with measurement",
    title = "Follow-Up Echocardiogram Availability"
  ) +
  theme_hv_poster()
```

---

# Mirror histogram (propensity score) {#dp-mirror}

**R equivalents:** `tp.lp.mirror-histogram_SAVR-TF-TAVR.R` (binary-match),
`tp.lp.mirror_histo_before_after_wt.R` (weighted IPTW)

`hv_mirror_hist()` replaces both scripts. Call `plot(mh)` to render.
Diagnostics are in `mh$tables$diagnostics`, working data in
`mh$tables$working`; layer scales, annotations, and a theme the usual way.

## Binary-match mode (`tp.lp.mirror-histogram_SAVR-TF-TAVR.R`) {#dp-mirror-binary}

`tp.lp.mirror-histogram_SAVR-TF-TAVR.R` compares propensity score distributions
for SAVR and TF-TAVR patients before and after 1:1 nearest-neighbour matching.
Upper bars show pre-match counts; a darker overlaid bar shows the matched
subset in each bin. Pass `match_col` to `hv_mirror_hist()` to activate this
mode -- the four internal fill levels (`before_g0`, `matched_g0`, `before_g1`,
`matched_g1`) then map directly to the four `scale_fill_manual()` values.

```r
dta <- sample_mirror_histogram_data(n = 400, separation = 1.5)
mh  <- hv_mirror_hist(dta)   # defaults: prob_t / tavr / match

plot(mh) +
  scale_fill_manual(
    values = c(
      before_g0  = "white",  matched_g0 = "green1",
      before_g1  = "white",  matched_g1 = "green4"
    ),
    guide = "none"
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  annotate("text", x = 20, y =  100, label = "SAVR",    size = 7) +
  annotate("text", x = 20, y = -100, label = "TF-TAVR", size = 7) +
  labs(x = "Propensity Score (%)", y = "Number of Patients") +
  theme_hv_poster()
```

## Weighted IPTW mode (`tp.lp.mirror_histo_before_after_wt.R`) {#dp-mirror-weighted}

`tp.lp.mirror_histo_before_after_wt.R` replaces the matched-subset bars with
IPTW weight sums per bin -- the right display when you are weighting rather
than matching. Pass `weight_col` instead of `match_col` to switch modes. The
fill levels change to `before_g0`, `weighted_g0`, `before_g1`, `weighted_g1`,
so update your `scale_fill_manual()` values accordingly.

```r
dta   <- sample_mirror_histogram_data(n = 400, add_weights = TRUE)
mh_wt <- hv_mirror_hist(
  dta,
  group_labels = c("Limited", "Extended"),
  weight_col   = "mt_wt"
)

plot(mh_wt) +
  scale_fill_manual(
    values = c(
      before_g0   = "white", weighted_g0 = "blue",
      before_g1   = "white", weighted_g1 = "red"
    ),
    guide = "none"
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  annotate("text", x = 30, y =  150, label = "Limited",  color = "blue", size = 5) +
  annotate("text", x = 30, y = -70,  label = "Extended", color = "red",  size = 5) +
  labs(x = "Propensity Score (%)", y = "#") +
  theme_hv_poster()
```

---

# Stacked histogram {#dp-stacked}

**R equivalent:** `hv_stacked()`

This plot has no direct SAS template predecessor -- it is a new addition
designed for the annual case-volume figures that were previously hand-built
with `PROC SGPLOT` `VBAR` / `VBARPARM` calls. `hv_stacked()` accepts a
data frame with one column for the x variable (typically operation year) and
one column for the group/fill variable. It returns a bar chart with stacked
fills; apply `scale_fill_brewer()` or `scale_fill_manual()` to set colours
consistent with your other figures in the same presentation.

```r
dta <- sample_stacked_histogram_data()
sh  <- hv_stacked(dta)

plot(sh) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x     = "Operation year",
    y     = "Count",
    fill  = "Group",
    title = "Annual Case Volume by Group"
  ) +
  theme_hv_poster()
```

---

# Using themes

All `plot()` calls return an unstyled ggplot object. Add a theme as the
final layer:

| Context | Call | Equivalent SAS device |
|---|---|---|
| Dark PowerPoint slide | `theme_hv_ppt_dark()` | `device=ppt` |
| Light PowerPoint slide | `theme_hv_ppt_light()` | `device=ppt` with white background |
| Journal manuscript | `theme_hv_manuscript()` | `device=eps` / PDF |
| Conference poster | `theme_hv_poster()` | Large-font poster |

You can pass `base_size` to any theme to scale all text simultaneously:

```r
p + theme_hv_manuscript(base_size = 10)   # smaller text for double-column
p + theme_hv_poster(base_size = 24)   # larger text for A0 poster
```

---

# Saving figures

## PowerPoint

`save_ppt()` inserts ggplot objects into a PowerPoint file as **editable
DrawingML vector graphics** via the `officer` and `rvg` packages — shapes,
lines, and text remain individually selectable in PowerPoint after export.
The first argument is `object` (not `plot`); the output path is `powerpoint`
(not `file` or `filename`).

```r
# Locate the bundled Cleveland Clinic slide template
template <- system.file("ClevelandClinic.pptx", package = "hvtiPlotR")

# Single slide — apply a PPT theme before saving
p_ppt <- p +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values   = c("steelblue"), guide = "none") +
  labs(x = "Years", y = "Prevalence (%)") +
  theme_hv_ppt_dark()

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = "figures/afib_prevalence.pptx",
  slide_titles = "AF Prevalence over Time"
)

# Multiple plots — one slide per figure in a single call
save_ppt(
  object       = list(fig1 = p_binary, fig2 = p_multi),
  template     = template,
  powerpoint   = "figures/np_curves.pptx",
  slide_titles = c("Binary Outcome", "Multi-group Comparison")
)
```

## PDF / TIFF for journals

`ggsave()` replaces the SAS `device=eps` / `device=tiff` options. For most
journals, 3.5 x 3.5 inches (single column) at 600 dpi satisfies TIFF
submission requirements; check the target journal's figure guidelines for the
exact width -- double-column figures typically need 7 inches. Use
`theme_hv_manuscript()` rather than `theme_hv_poster()` before saving, since
manuscript figures need smaller base text sizes than poster figures.

```r
ggsave("figures/afib_prevalence.pdf",  p, width = 3.5, height = 3.5, units = "in")
ggsave("figures/afib_prevalence.tiff", p, width = 3.5, height = 3.5,
       units = "in", dpi = 600)
```

---

# Parametric hazard/survival (`tp.hs.*`) {#hs-dead}

**Ports:** `tp.hs.dead.setup.sas`, `tp.hs.dead_uses_setup.sas`,
`tp.hs.dead.procedure.tdepth.sas`, `tp.hs.dead.conditional.setup.sas`,
`tp.hs.dead.conditional.uses_setup.sas`,
`tp.hs.dead.compare_benefit.setup.sas`,
`tp.hs.uslife_estimates_generate_stratify_.age.sas`,
`tp.hs.uslife_generates_matched_estimates.sas`

**R equivalents:** `hazard_plot()`, `survival_difference_plot()`

The `tp.hs.*` family extends `tp.hp.dead.*` by using the `%hazpred` macro to
generate patient-specific parametric survival predictions from a fitted
multivariable hazard model. The *setup* templates compute and store:

- **Cumulative hazard per patient** at last follow-up (for observed vs.
  expected goodness-of-fit tests via `%chisqgf`).
- **Individual survivorship curves** on a fine time grid (for mean-curve
  aggregation with `PROC SUMMARY`).

Subsequent *use* templates stratify the aggregated mean curves by a covariate
and overlay Kaplan-Meier estimates from `%kaplan`. The resulting SAS datasets
map directly to `hazard_plot()` in R:

| SAS dataset | Columns used | R source |
|---|---|---|
| `means` (from `PROC SUMMARY` over `predict`) | `SSURVIV`, `SCLLSURV`, `SCLUSURV` | `sample_hazard_data()` |
| `plout` (from `%kaplan`) | `CUM_SURV`, `CL_LOWER`, `CL_UPPER` | `sample_hazard_empirical()` |
| `est.uslife_*` (from `%usmatchd`) | `SMATCHED` | `sample_life_table()` |

## Parametric survival with KM overlay

Reproduces the core figure from `tp.hs.dead_uses_setup.sas` and
`tp.hs.dead.procedure.tdepth.sas`: a mean parametric survival curve (solid line
with confidence band) overlaid with Kaplan-Meier empirical estimates (symbols
with error bars).

```r
dat_hp <- sample_hazard_data(n = 500, time_max = 10)
emp_hp <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)

hazard_plot(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  theme_hv_poster()
```

## Stratified by covariate

`tp.hs.dead.procedure.tdepth.sas` overlays curves for depth-of-invasion groups
(`tdepth = 1, 2, 3`). Pass `group_col` to `hazard_plot()` and supply matched
`sample_hazard_data()` / `sample_hazard_empirical()` calls with a `groups`
vector whose names become the legend labels.

```r
dat_strat <- sample_hazard_data(
  n = 500, time_max = 10,
  groups = c("pT1" = 0.4, "pT2" = 1.0, "pT3" = 1.8)
)
emp_strat <- sample_hazard_empirical(
  n = 500, time_max = 10, n_bins = 5,
  groups = c("pT1" = 0.4, "pT2" = 1.0, "pT3" = 1.8)
)

hazard_plot(
  dat_strat,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  group_col     = "group",
  empirical     = emp_strat,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(
    values = c("pT1" = "steelblue", "pT2" = "forestgreen", "pT3" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("pT1" = "steelblue", "pT2" = "forestgreen", "pT3" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  theme_hv_poster()
```

## Conditional survival after hospital discharge {#hs-conditional}

`tp.hs.dead.conditional.setup.sas` computes individual survivorship curves
starting from the time of hospital discharge rather than the operation date.
The conditional survival `S(t | discharge)` is `S(t) / S(t_discharge)` for
each patient. In R, you use the same `hazard_plot()` call — only the x-axis
label and data preparation change.

```r
dat_cond <- sample_hazard_data(n = 500, time_max = 10)
emp_cond <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)

hazard_plot(
  dat_cond,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_cond,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Discharge", y = "Survival after Discharge (%)") +
  theme_hv_poster()
```

## US life-table overlay {#hs-uslife}

`tp.hs.uslife_*` templates call `%usmatchd` to generate age-, sex-, and
race-matched US population life-table survival curves for each patient.
`sample_life_table()` provides representative matched curves stratified by age
group for use in R examples.

```r
dat_age <- sample_hazard_data(
  n = 600, time_max = 10,
  groups = c("<60" = 0.4, "60\u201385" = 1.0, "\u226585" = 2.0)
)
emp_age <- sample_hazard_empirical(
  n = 600, time_max = 10, n_bins = 6,
  groups = c("<60" = 0.4, "60\u201385" = 1.0, "\u226585" = 2.0)
)
lt <- sample_life_table(
  age_groups = c("<60", "60\u201385", "\u226585"),
  age_mids   = c(50, 72, 88),
  time_max   = 10
)

hazard_plot(
  dat_age,
  estimate_col     = "survival",
  lower_col        = "surv_lower",
  upper_col        = "surv_upper",
  group_col        = "group",
  empirical        = emp_age,
  emp_lower_col    = "lower",
  emp_upper_col    = "upper",
  reference        = lt,
  ref_estimate_col = "survival",
  ref_group_col    = "group"
) +
  scale_colour_manual(
    values = c("<60" = "steelblue", "60\u201385" = "forestgreen",
               "\u226585" = "firebrick"),
    name   = "Age group"
  ) +
  scale_fill_manual(
    values = c("<60" = "steelblue", "60\u201385" = "forestgreen",
               "\u226585" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Survival (%)",
       caption = "Dashed lines: US population life table") +
  theme_hv_poster()
```

## Treatment benefit distribution {#hs-benefit}

`tp.hs.dead.compare_benefit.setup.sas` computes, for each patient, the
difference in predicted survival at a fixed time point (5 years) under two
treatment arms (e.g. ASA vs. no ASA). That per-patient distribution is the answer to "who benefits and by how much."
`survival_difference_plot()` plots the mean difference curve over time.

```r
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("No ASA" = 1.0, "ASA" = 0.75)
)

survival_difference_plot(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                      colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(-5, 40),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival Difference (%)") +
  theme_hv_poster()
```

---

# Adding new ports {#adding-new-ports}

As additional SAS templates are ported to R, update this guide:

1. **Add a row** to the lookup table at the top of this file with the SAS
   template name, family prefix, R function, and section anchor.

2. **Add a section** following the existing pattern: name the port, describe
   the SAS workflow, show the R equivalent with a runnable example, and
   document any column name mapping differences.

3. **Export the new R function** by adding `@export` to its roxygen block and
   running `devtools::document()` to regenerate `NAMESPACE`.

4. **Update `DESCRIPTION`** if a new package dependency is required.

Templates currently planned for porting:

- `tp.ce.states.*` competing-events / state-occupancy → (pending)
- `tp.gp.*` grouped longitudinal ordinal models → (pending)

---

# Session info

```{r session-info}
sessionInfo()
```
